---
title: "SpatialVDJ - Clonal analysis vs TLS cluster"
author: "Guilhem Pupier"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmdformats::readthedown:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In order to analyze BCR and TCR clones in a ccRCC tumors samples at a scale closer to single cell, we used the [SpatialVDJ](https://www.science.org/doi/10.1126/science.adf8486?url_ver=Z39.88-2003&rfr_id=ori:rid:crossref.org&rfr_dat=cr_pub%20%200pubmed) method developped by Engblom et al. It is an adaptation of Visium (10X Genomics) on Fresh Frozen slices with a capture of BCR and TCR full length cDNA, before fragmentation, and long read sequencing using PacBio sequencing platform. On an other part, spatial gene expression is also performed.

The goal is to analyze BCR and TCR relationships regarding to TLS : clonal evolution (expansion, SHM, CSR) vs TLS at spatial resolution.

Sections :

- **Preprocess** :  
All pre-process step of **SpatialVDJ** data following clonotype preparation (see script Figure2_SpatialVDJ_clonotypes.Rmd) adding clonal grouping step and lineage tree reconstruction.    

- **Reload merged dataset** :  
Reload data to generate plots.

- **Figures** : 
Script to reproduce Figure 3 and S8.  

# Load required packages


```{r,eval=T}
library(STutility)
library(Seurat) # to manage spatial transcriptomic data

library(parallel) # parApply ...

library(patchwork)
library(ggplot2)
library(ggpubr)
library(ggalluvial)
library(grid)
library(gridExtra)
library(scales) # plot functions
#from Heather Turner @stackoverflow ## Adapt y scale in log scale 
base_breaksLog10 <- function(n = 10){
  function(x) {
    axisTicks(log10(range(x, na.rm = TRUE)), log = T, n = n)
  }
}

base_breaksLog1p <- function(n = 10){
  function(x) {
    axisTicks(log1p(range(x, na.rm = TRUE)), log = T, n = n)
  }
}


library(pastecs)
library(RColorBrewer)
library(viridis) # color palettes

library(stringi)
library(stringr) # string manipulations

library(dplyr) 
library(tidyr)
library(reshape2) # table manipulations

# from Maxime Meylan
plot_contingency <- function(data,
           variables = c("TLS_pos", "HBV"),
           hide_legend = F,
           text_size=6,
           create_labels = T,
           inverse_factor2=F,
           order_factor=NULL,
           palette = NA,
           test_name = c("chisq.test", "fisher.test")[1],
           labs = c(xlab = "",
                    ylab = "",
                    legend_title = ""),
           hide_test = F,
           title = ""){
    if(inverse_factor2){
      data[,variables[2]] <- factor(data[,variables[2]],levels=rev(levels(data[,variables[2]])))
    }
    cont_table <- table(data[,variables[1]],data[,variables[2]])
    #clean data table to avoid empty rows
    non_empty_rows <- rowSums(cont_table)!=0
    cont_table <- cont_table[non_empty_rows,]

    if(create_labels){
      #sapply(levels(factor(data[,variables[2]])), function(x) paste(variables[2],x,sep="_"))
      colnames(cont_table) <- sapply(levels(factor(data[,variables[2]])), function(x) paste(variables[2],x,sep="_"))
      rownames(cont_table) <- sapply(levels(factor(data[,variables[1]])), function(x) paste(variables[1],x,sep="_"))
    }
  if(test_name != "fisher.test" & test_name != "chisq.test"){
    stop("incorrect test, valid options are fisher.test or chisq.test")
  }
  if(test_name == "fisher.test"){
    error_test <-  if(class(try(fisher.test(cont_table),silent=T)) == "try-error"){
      test_res <- fisher.test(cont_table,simulate.p.value = T)
    }else{
      test_res <- fisher.test(cont_table)
    }
  }
  if(test_name == "chisq.test"){
    test_res <- chisq.test(cont_table)
  }
  #make sure there is a matrix (enough values in both sides)
  if(!is.null(dim(cont_table))){
    cont_table_perc <-apply(cont_table,1,function(x) {
      res <- x / sum(x) *100
    })
    cont_table_m <- reshape2::melt(cont_table_perc)
    cont_table_m$Var1 <- factor(cont_table_m$Var1)
    p <- ggplot(data=cont_table_m, aes(x=Var2,y=value,fill=Var1,label=value))
    p <- p + geom_bar(stat="identity",width=0.95)
    #to reverse order of bars
    #p <- p + geom_bar(stat="identity",width=0.95,position = position_stack(reverse = TRUE))
    p <- p + geom_text(label = paste0(as.vector(t(cont_table)),"\n(",format(cont_table_m$value,digits = 1),"%)"),
                       position = position_stack(vjust = .5),size=text_size)
    p <- p + theme_minimal()

    if(any(!is.na(palette))){
      p <- p + scale_fill_manual(values = palette)
    }
    #p <- p + theme(legend.title = element_blank())#,axis.text.x = element_text(angle = 25, hjust = 1))
    p <- p + guides(fill=guide_legend(title=labs[3]))
    if(hide_test){
      p <- p + labs(subtitle=paste("p = ",format(test_res$p.value,digits=3),sep=" "),title=title,x=labs[1],y=labs[2])
      if(test_res$p.value < 0.05 ) p <- p + theme(plot.subtitle=element_text(face="bold"))
    }else{
      p <- p + labs(subtitle=paste(test_name, " p = ",format(test_res$p.value,digits=3),sep=" "),title=title,x=labs[1],y=labs[2])
      if(test_res$p.value < 0.05 ) p <- p + theme(plot.subtitle=element_text(face="bold"))
    }

    if(hide_legend){
      p <- p + guides(fill=FALSE)
    }
  }else{
    p <- ggplot()
    test_res <-1
  }
  return(list(p,test_res,cont_table))

}

library(scatterpie)
library(ggfun) 
library(ggforce) # Packages for spatial pie plot
geom_scatterpie_edit <- function (mapping = NULL, data, cols, pie_scale = 1, sorted_by_radius = FALSE, 
    legend_name = "type", long_format = FALSE, ...) 
{
    if (is.null(mapping)) 
        mapping <- aes_(x = ~x, y = ~y)
    mapping <- modifyList(mapping, aes_(r0 = 0, fill = as.formula(paste0("~", 
        legend_name)), amount = ~value))
    if (!"r" %in% names(mapping)) {
        xvar <- get_aes_var(mapping, "x")
        size <- 5                         ###### Change here to get always the same size of pie
        data$r <- size
        mapping <- modifyList(mapping, aes_(r = size))
    }
    names(mapping)[match(c("x", "y"), names(mapping))] <- c("x0", 
        "y0")
    if (long_format == TRUE) {
        df <- data
        names(df)[which(names(df) == cols)] = legend_name
        cols2 <- enquo(cols)
    }
    else {
        data <- data[rowSums(data[, cols]) > 0, ]
        cols2 <- enquo(cols)
        df <- gather(data, "type", "value", !!cols2)
        df$type <- factor(df$type, levels = cols)
        names(df)[which(names(df) == "type")] = legend_name
    }
    if (!"group" %in% names(mapping)) {
        xvar <- get_aes_var(mapping, "x0")
        yvar <- get_aes_var(mapping, "y0")
        df <- as.list(dplyr::group_split(dplyr::group_by(df, 
            !!xvar, !!yvar)))
        names(df) <- seq_len(length(df))
        df <- dplyr::bind_rows(df, .id = ".group.id")
        mapping <- modifyList(mapping, aes_(group = ~.group.id))
    }
    if (!sorted_by_radius) {
        return(geom_arc_bar(mapping, data = df, stat = "pie", 
            inherit.aes = FALSE, ...))
    }
    if ("r" %in% colnames(df)) {
        rvar <- "r"
    }
    else {
        rvar <- get_aes_var(mapping, "r")
    }
    lapply(split(df, df[, rvar, drop = TRUE])[as.character(sort(unique(df[, 
        rvar, drop = TRUE]), decreasing = TRUE))], function(d) {
        geom_arc_bar(mapping, data = d, stat = "pie", inherit.aes = FALSE, 
            ...)
    })
}
 

### Packages for repertoire
library(airr)
library(data.table) # function fread() for efficient and fast read of tab file
library(alakazam)
library(shazam)

library(dowser) # lineage tree with Immcantation pipeline

library(ggtree) 
library(ggraph)
library(tidygraph) # reshape tree with ggplot/tidygraph

## pretty palette
my_colors <- c("#66b803", "#f94144", "#f8961e", "#3574b8", "#DE77AE", "#9970AB", "#D5E4A2", "#71D0F5", "#B1746F", "#ADE2D0", "#20DE8BFF", "#CCDE8BFF", "#f7f31c", "#FFA88BFF", "#FF6A8BFF")

# Give the path of the Immcantation folder : repertoire data for SpatialVDJ preprocessed with Immcantation pipeline (see paper method). With a back / at the end
ImmcantationFolder <- "PATH/"

# Load table of identifier matching with anonym and original id : Table not shared due to patient privacy requirement. Only anonym id provided (original id not necessary to reproduce the script. Just need to adapt corresponding lines)
Anonym_table <- "" # not shared

# Following vector are just built for consistency with original script. Again, just need to adapt with the correct id shared
#### A vecotr name FileID contained the SpatialVDJ id
FileID <- Anonym_table$SpatialVDJ_id
names(FileID) <- FileID

#### A vector named IDmatch contained the SpatialGEX id and SpatialVDJ in names
IDmatch <- Anonym_table$SpatialGEX_id
names(IDmatch) <- FileID

# Output folder to retrieve the files and plot
res_folder <-"PATH/Figure3_S8/"
```

# Preprocess - Not rendered

## SpatialVDJ clonal grouping

All script bellow are run only one time - not re run for plot/render markdown. 
They should be considered with preprocess steps of the previous script :  Figure2_SpatialVDJ_clonotypes.Rmd

Note : the final clonotype tables (1 table per sample) generated from previous script were named as follow (file_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI.tsv). The file 20260721_airr_igblast-db_full_database.tsv provided corresponds to a merge of all individual clonotype table (see previous script,Figure2_SpatialVDJ_clonotypes.Rmd, section Merged all seurat and all metadata).

### Clustering sequences into clonal groups


#### Calculate the Distance to the Nearest Distribution and find a threshold

```{r,eval=F}

SampleId <- Anonym_table$SpatialVDJ_id

IDmatch <- Anonym_table$SpatialGEX_id
names(IDmatch) <- SampleId


for(file in SampleId){
  # Do not need to adapt from here - path of Igblast quality filtered database
  file_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI.tsv")
  
  # Load airr 
  airr <- read_rearrangement(file_path)
  
  airr_heavy <- airr %>%
  filter(locus == "IGH")
  
  # Open a file connection for writing a log file ##un comment if needed
  sink(paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igh_dist_nearest_and_threshold_log.txt"), append = FALSE, type = c("output"))
  
  print(paste("Dimension of the IGH repertoire :",dim(airr_heavy)[[1]],"x",dim(airr_heavy)[[2]]))
  
  print(paste("Start distToNearest for file :",file))
  t1 <- Sys.time()
  airr_heavy <- distToNearest(airr_heavy, sequenceColumn="junction",vCallColumn="v_call", jCallColumn="j_call", model="ham", first=FALSE, normalize="len", nproc=12)
  t2 <- Sys.time()
  print(paste("distToNearest time with nproc=12 - Hamming length normalize :"))
  print(t2-t1)
  
  print(paste("Start findThreshold for file :",file))
  t1 <- Sys.time()
  threshold <- findThreshold(airr_heavy[["dist_nearest"]], method="density")
  t2 <- Sys.time()
  print(paste("findThreshold time :"))
  print(t2-t1)
  
  clone_threshold <- round(threshold@threshold,digits = 2) 
  
  print(paste("findThreshold - density method - round 2 digits :",clone_threshold))
  
  # Close the sink connection to save the output to the file ## un comment if needed
  sink()
  
  #Save the plot
  #create directory to store results per file
  if(!file.exists(paste0(res_folder,"03_IgBlast/Clone_Dist_Threshold"))) dir.create(paste0(res_folder,"03_IgBlast/Clone_Dist_Threshold")) 

  p<-plot(threshold, binwidth=0.02, silent=FALSE)
  pdf(paste0(res_folder,"03_IgBlast/Clone_Dist_Threshold/",file,"_Hamming_length_normalized_igh_threshold.pdf"), width = 8, height = 7)
  print(p)
  dev.off()

  # Save the repertoire data after collapse duplicates
  write_rearrangement(airr_heavy, file=paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist.tsv"))
  
  rm(list=setdiff(ls(),c("FileID","SampleId","IDmatch","ImmcantationFolder","res_folder")))
  gc()
  
}

```

#### Define clones

Define clone according to previously defined threshold. 
Create germline according to clone information.

```{bash,eval=F}
# in a environment : conda activate pRESTO

LongRead_dir="PATH/"

cd ${LongRead_dir}

#List file 
files=`cat "${LongRead_dir}sampleId_batch1_run3_SMRTbellCRC.txt"`

for file in $files; do

  echo $file
  
  in="${LongRead_dir}Immcantation/8_PostIgBlastFilter/${file}_igh_dist_nearest_and_threshold_log.txt"
  out="${LongRead_dir}Immcantation/8_PostIgBlastFilter/${file}_igh_dist_nearest_and_threshold.log"
  
  # first sed remove [1] "
  # Second sed remove remaining "
  # Third sed remoe explicit \n lign
  
  cat $in | sed -E "s/^\[.{4}//g" | sed -E "s/\"//g" | sed -E "s/^.n$//g" > $out
  
  #Retrieve thresold from log of the correct sample.
  ## read log ## get the lign : 
  Threshlog="${LongRead_dir}Immcantation/8_PostIgBlastFilter/${file}_igh_dist_nearest_and_threshold.log"
  threshold=`grep "findThreshold - density method - round 2 digits : " $Threshlog | sed "s/findThreshold.*: //g"`
  
  # If threshold is NA or threshold is greater than 0.15, then replace by 0.15
  if [[ "$threshold" == "NA" || $(echo "$threshold > 0.15" | bc -l) -eq 1 ]]; then
    threshold=0.15
  fi
  
echo "Runing DefineClones.py"
DefineClones.py -d "${LongRead_dir}Immcantation/8_PostIgBlastFilter/${file}_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist.tsv" --model ham --norm len --dist $threshold --mode gene --act set --nproc 12 --outdir "${LongRead_dir}Immcantation/8_PostIgBlastFilter/" --format airr --log "${LongRead_dir}Immcantation/8_PostIgBlastFilter/${file}_DefineClone.log" >> "${LongRead_dir}Immcantation/pipeline_log/${file}_ChangeO_and_PostProcess_perf.log" 2>&1

echo "Runing CreateGermlines.py"
CreateGermlines.py -d "${LongRead_dir}Immcantation/8_PostIgBlastFilter/${file}_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass.tsv" -r PATH/germlines/imgt/human/vdj -g dmask --format airr --cloned --outdir "${LongRead_dir}Immcantation/8_PostIgBlastFilter/" >> "${LongRead_dir}Immcantation/pipeline_log/${file}_ChangeO_and_PostProcess_perf.log" 2>&1

done

```



### Clonal lineage tree analysis with Immcantation pipeline

This script is provided (regular Immcantation lineage tree analysis) but is not actualy used in the final analysis / article >> see TRIBAL lineage tree section after.

```{r,eval=FALSE}
library(dowser)
library(alakazam)
library(airr)
library(dplyr)
library(ggtree)
```


```{r,eval=F}

SampleId <- Anonym_table$SpatialVDJ_id

IDmatch <- Anonym_table$SpatialGEX_id
names(IDmatch) <- SampleId


for(file in FileID){
  
  #folder to save plot
  if(!file.exists(paste0(res_folder,"03_IgBlast/IGHclones/",file))) dir.create(paste0(res_folder,"03_IgBlast/IGHclones/",file)) 
  
  
  # Do not need to adapt from here - path of Igblast quality filtered database
  file_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass.tsv")
  
  # Load airr 
  airr <- read_rearrangement(file_path)
  
  # Open a file connection for writing a log file ##un comment if needed
  sink(paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igh_cloneTree_log.txt"), append = FALSE, type = c("output"))
  
  print(paste("number of unique clone_id :",length(unique(airr$clone_id))))

  # 1. Format
  
  airr$read_count <- as.numeric(airr$read_count)

  ## format sequence_id column in order to shorter it (it was concatenated by all sequences collapsed : choose only one sequence id, the first one ...)
  airr$sequence_id <-gsub(",.*","",airr$sequence_id)
  
  ## format c_imgt_form so that it do not takes to many space in the plot (as label) : either M, S or ?
  airr$simplified_form<-ifelse(airr$c_imgtref_form=="Secreted",yes = "S",
                               no = ifelse(airr$c_imgtref_form=="Membrane",yes = "M",no = "?"))
  #airr$simplified_form_bold<-ifelse(airr$simplified_form=="M",yes = "bold",no = "plain")
  
  print(paste("Start formatClones nproc=10 for file :",file))
  
  warning(paste(file,"1st clones formating"))
  
  t1 <- Sys.time()
  clones <- formatClones(airr, traits=c("isotype","c_imgtref_form","sb","simplified_form"),num_fields=c("read_count","umi_count"),text_fields = c("clonotype_id","sb","sb_umi"), columns=c("d_call"), minseq=2,nproc=10)
  t2 <- Sys.time()
  print(paste("formatClones time to convert IGH repertoire in a clone data table - 10 threads - minseq=2 :"))
  print(t2-t1)
  
  print(paste("Number of IGH clones with min 2 Seq - Polytomy by SB :",length(clones$clone_id)))
  
  # 2. Build the Trees
  # build maximum parsimony trees 
  
  print(paste("Start getTrees nproc=10 for file :",file))
  
  t1 <- Sys.time()
  clones <- getTrees(clones ,nproc = 10)
  clones <- scaleBranches(clones, edge_type = "mutations") # make a scale in mutations number (easier to interprete)
  t2 <- Sys.time()
  print(paste("getTrees time to convert IGH repertoire in a clone data table - 10 threads - Min 2 Seq :"))
  print(t2-t1)
  
  # 3. Visualize
  # Plot the trees. Save them in a list of plots. 
  # Use tip metadata: isotype and collapse_count
  
  color_palette <- c("IGHM"="#253494", "IGHD"="#1D91C0", "IGHG1"="#d62728", "IGHG2"="#ff7f0e",  "IGHG3"="#FF6EC7", "IGHG4"="#FCCF05", "IGHA1"="#2ca02c", "IGHA2"="#36e40a","IGHE"="#8c564b","IGH_NA"="grey")
  
  print(paste("Start plotTrees for file :",file))

  t1 <- Sys.time()
  tree_plots <- plotTrees(clones,tips ="isotype", tipsize="umi_count", tip_palette=c(color_palette, "Germline"="#000000"),layout = "roundrect",scale=1)
  names(tree_plots)<-clones$clone_id
  
  warning(paste(file,"1st clones format - save plot"))
  
  tree_plots <- lapply(clones$clone_id,function(clnID){
    plot<-tree_plots[[clnID]]
    plot<-plot+
      geom_tiplab(aes(label=simplified_form),offset = -1,size=2.5,hjust = +0.5)+ # Add S,M,? information at the tip for the form of IGH
      ggtree::layout_dendrogram() # switch as a dendrogram with germline in the top
    
    ggsave(plot = plot,
           filename = paste0(res_folder,"03_IgBlast/IGHclones/",file,"/","Dowser_MaxParsimonyTree_Iso&Form_PolytomyBySB_IGHclone",clnID,".pdf"),
           width = 8,height = 4)
    
    plot
  })
  
  t2 <- Sys.time()
  print(paste("plotTrees time"))
  print(t2-t1)
  
  #Save the RDS in 8_PostIgBlastFilter
  saveRDS(clones,paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igh_clones_datatable_MaxParsimony_PolytomyBySB.rds"))
  saveRDS(tree_plots, paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igh_clones_treePlot_MaxParsimony_Iso&Form_PolytomyBySB.rds"))

  
  ##### 2nd version of Tree plot - Collapse by clonotype : size of spots in the tree is the number of spots
  warning(paste(file,"2nd clones formating"))
  
  clones <- formatClones(airr, traits=c("isotype","c_imgtref_form","simplified_form"),num_fields=c("read_count","umi_count"),text_fields = c("clonotype_id","sb","sb_umi"), columns=c("d_call"), minseq=2,nproc=10)
  clones <- getTrees(clones ,nproc = 10)
  clones <- scaleBranches(clones, edge_type = "mutations") 
  
  print(paste("Number of IGH clones with min 2 Seq - Collapsed by clonotype :",length(clones$clone_id)))
  
  tree_plots <- plotTrees(clones,tips ="isotype", tipsize="collapse_count", tip_palette=c(color_palette, "Germline"="#000000"),layout = "roundrect",scale=1)
  names(tree_plots)<-clones$clone_id
  
  
  warning(paste(file,"2nd clones format - save plot"))
  tree_plots <- lapply(clones$clone_id,function(clnID){
      plot<-tree_plots[[clnID]]
      plot<-plot+
        geom_tiplab(aes(label=simplified_form),offset = -1,size=2.5,hjust = +0.5)+ # Add S,M,? information at the tip for the form of IGH
        ggtree::layout_dendrogram()+ # switch as a dendrogram with germline in the top
        labs(size="Spot_count")
      
      ggsave(plot = plot,
             filename = paste0(res_folder,"03_IgBlast/IGHclones/",file,"/","Dowser_MaxParsimonyTree_Iso&Form_CollapseByClonotype_IGHclone",clnID,".pdf"),
             width = 8,height = 4)
      
      plot
    })
  
  #Save the RDS in 8_PostIgBlastFilter
  saveRDS(clones,paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igh_clones_datatable_MaxParsimony_CollapseByClonotype.rds"))
  saveRDS(tree_plots, paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igh_clones_treePlot_MaxParsimony_Iso&Form_CollapseByClonotype.rds"))
  
  
  # Close the sink connection to save the output to the file ## un comment if needed
  sink()
  
  rm(list=setdiff(ls(),c("FileID","SampleId","IDmatch","ImmcantationFolder","res_folder")))
  gc()
  
}


```

### New clonal lineage analysis with TRIBAL

TRIBAL is a new method of Tree inference of B cell clonal lineages (TRIBAL) which jointly model the evolutionary processes of somatic hypermutation and class switch recombination (isotype away !).
NB: see github [TRIBAL](https://github.com/elkebir-group/TRIBAL)

It was initially developped for single cell VDJ data with isotype information available. I want to use this software with only IGH clonotypes from SpatialVDJ Long read and Isotype calling in order to improve the lineage tree by adding the class switch information.

From the github ;

The cli has two sub-commands:

preprocess - filter the data and find a multiple sequence alignment and parsimony forsest for each clonotype.
fit - infer a set of optimal B cell lineage trees per clonotype an a shared isotype transition probability matrix.

Preprocess:
The preprocessing command will: 1. filter out clonotypes that are below the minimum number of cells 2. filter out cells which have v alleles that differ from the majority of the clonotype 3. perform a multiple sequence alignment (MSA) for each valid clonotype using mafft 4. infer a parsimony forest for each clonotype given the MSA using dnapars
The --data, --roots and --encoding are required arguments. See data description for more details. The --encoding argument should be that path to a text file that lists the correct isotype ordering as well as the isotype labels that are present in the input data.

Fit:
The fit sub-command will infer a set of B cell lineage trees for each clonotype and a shared isotype transition probability matrix.
The pickled dictionary of clonotypes clonotypes.pkl that is output from tribal preprocess will be the input to tribal fit.
There are four optional outputs from tribal fit:
--transmat-infer: the inferred isotype transition probability matrix.
--pickle : a dictionary with clonotype id as key and value is a LineageTreeList containing the optimal lineage trees for each clonotype.
--score : a csv file containing the SHM parsimony scores and CSR likelihood.
--write-results : a directory where the lineage trees files will be saved including:
a fasta file containing the inferred BCR sequences,
a csv file containing the inferred isotypes,
a text file containing the edge list of the lineage tree,
a png file containing a visualization of the lineage tree.

#### Prepare input files

To run TRIBAL, 3 input files are required:

The sequencing data : a csv file with the following columns :

- cellid : unqiue id or barcode of the sequenced B cell. Here, because it is SpatialVDJ data and in a same spatial barcode (SB) there can be multiple clonotypes (unique sequences from different cells), the cellid will be a concatenate between the SB tag and the clonotype_id tags  
- clonotype : unique clonotype id to which that cell belongs , i.e., groups of cells that descend from the same naive B cell receptor. In this case, it corresponds to the clone groups, ie clone_id.  
- heavy_chain_isotype : c_call IGHG1, IGHA1, IGHE ...  
- heavy_chain_allele : v_call : actually it is the v_gene col : choose only the best v_call allele)  
- heavy_chain_seq : the variable region sequence of the heavy chain. In this case it correspond to the sequence_alignment column. Try with the AIRR format (with the dots), but if necessary, remoove the dots.  

The germline clonotype roots sequences : a csv file with the following columns :

- clonotype : unique clonotype id of the germline root (naive BCR). Here the clone_id  
- heavy_chain_root :  heavy chain variable region germline root sequence. The germline_alignment sequence.  

The encoding file (isotypes.txt) : a txt file with only one column, one isotype per line, ordered according to the genome. The example file contains only 8 isotypes for the human. IGHM, IGHG3, IGHG1, ... IGHE, IGHA2. There is no IGHD isotype in this file. For the first try, I will try to add the IGHD isotype just after the IGHM, as well as the IGH_NA, at the end of the file, to keep the maximum information.

IMPORTANT NOTE : the sequence and germline alignment column from the IgBlast/AIRR format with dot is not working with TRIBAL. It seams to work with the dots replaced as N (should be checked that it is OK).

```{r,eval=FALSE}
SampleId <- Anonym_table$SpatialVDJ_id

IDmatch <- Anonym_table$SpatialGEX_id
names(IDmatch) <- SampleId


for(file in FileID){
  
  #folder to save Input and output files from TRIBAL
  if(!file.exists(paste0(ImmcantationFolder,"8_PostIgBlastFilter/TRIBAL_tree"))) dir.create(paste0(ImmcantationFolder,"8_PostIgBlastFilter/TRIBAL_tree")) 
  
  # PAth of the final IGH clone with germline reconstruct database
  file_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass.tsv")
  
  # Load airr 
  airr <- read_rearrangement(file_path)

  ## Create a cell column by concatenating the SB and the clonotype_id
  airr$cell <- paste(airr$sb,airr$clonotype_id,sep = "-")
  
  ## Add the string "Clonotype_" before the clone_id to match the required format of TRIBAL
  #### NB : do not seams to be required : changed to Clone to avoid confusion
  airr$clone_id <- paste0("Clone_",airr$clone_id)
  
  #replace _ by - because it seems to generate error in TRIBAL fit (key error)
  airr$cell <- gsub("_","-",airr$cell)
  
  ## Sub in 2 data.frame with the required columns

  ## TEST on 2 clones : one with only 1 clonotype, one with 4 clonotypes 
  #### airr <- airr[airr$clone_id %in% c("Clone_1","Clone_140"),]
  
  
  ### First SeqData
  SeqData <- airr[,c("cell","clone_id","isotype","sequence_alignment","v_gene")]
    ### rename column to fit the required input
  colnames(SeqData)<- c("cell","clonotype","heavy_chain_isotype","heavy_chain_seq","heavy_chain_v_allele")
  
  #### TRY replacing all dots by N
  SeqData$heavy_chain_seq <- str_replace_all(SeqData$heavy_chain_seq,"\\.","N")
  
  ### Then RootData
  RootData <- airr[,c("clone_id","germline_alignment_d_mask")]
    ### rename column to fit the required input
  colnames(RootData)<- c("clonotype","heavy_chain_root")
    ### Keep only one unique version per clone_id (all the germline_alignment are duplicated sequences) for one clone_id
  RootData <- RootData[!duplicated(RootData$clonotype),]
  #### TRY replacing all dots by N
  RootData$heavy_chain_root <- str_replace_all(RootData$heavy_chain_root,"\\.","N")
  
  #Write Sequencing data csv
  out_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/TRIBAL_tree/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass_TRIBAL_SeqData.csv")
  write.csv(x = SeqData,file = out_path,quote = F,row.names = F)
  
  #Write Roots sequences csv
  out_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/TRIBAL_tree/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass_TRIBAL_RootData.csv")
  write.csv(x = RootData,file = out_path,quote = F,row.names = F)
  
}  


```


#### Run TRIBAL

```{bash,eval=F}
## Make sure to add the encoding txt file to the TRIBAL_tree folder
mv PATH/20241028_HomoSapiens_isotypes_inSpatialVDJ_order.txt PATH/Processed_3rdRun/Immcantation/8_PostIgBlastFilter/TRIBAL_tree/

#conda activate tribal

###### Test TRIBAL with TEST DATA from Github repo in the folder PATH/test_TRIBAL/
tribal preprocess --data "PATH/test_TRIBAL/data.csv"  \
    --roots "PATH/test_TRIBAL/roots.csv" \
    --encoding "PATH/test_TRIBAL/isotypes.txt" \
    --min-size 2 \
    --cores 3 \
    --out "PATH/test_TRIBAL/clonotypes.pkl" \
    --heavy

tribal fit -c "PATH/test_TRIBAL/clonotypes.pkl" \
  --n_isotypes 10 \
  --mode refinement \
  --cores 10 \
  --transmat-infer "transmat.txt" \
  --pickle "lineage_trees.pkl" \
  --write-results "results" \
  --score "objective.csv"



###### TEST With a subset from my DATA (only 2 clones, one with 1 clonotype, 1 with 4 clonotypes)
tribal preprocess --data "PATH/test_TRIBAL/TestTRIBAL_SeqData.csv"  \
    --roots "PATH/test_TRIBAL/TestTRIBAL_RootData.csv" \
    --encoding "PATH/test_TRIBAL/isotypes.txt" \
    --min-size 2 \
    --cores 3 \
    --out "PATH/test_TRIBAL/TestTRIBAL_clonotypes.pkl" \
    --heavy


tribal fit -c "TestTRIBAL_clonotypes.pkl" \
  --n_isotypes 10 \
  --mode refinement \
  --cores 10 \
  --transmat-infer "TestTRIBALL_transmat.txt" \
  --pickle "TestTRIBAL_lineage_trees.pkl" \
  --write-results "TestTRIBAL_results" \
  --score "TestTRIBAL_objective.csv"



##### NB : There must be no underscore in cell id or it generates an error

# conda deactivate (deactivate base)
#conda activate tribal

LongRead_dir="PATH/"

cd ${LongRead_dir}

#List file 
files=`cat "${LongRead_dir}sampleId_batch1_run3_SMRTbellCRC.txt"`

for file in $files; do

  echo $file 


  echo "run preprocess"
  ## First run the preprocess command
  #### Only on IGH because from SpatialVDJ I do not have paired clonotypes (Maybe can be used with repair inferred pairs ...)
  ### --min-size : minimum clonotype size (default 4). Change it to 2 to be in a similar situation as Dowser package usage (above)
  ### --cores : number of cores
  ### --heavy : only use the heavy chain and ifnore the light chain
  ### --dataframe : path to where the filtered dataframe with additional sequences and isotype encodings should be saved
  ### --out : path to where pickled clonotype dictionary input should be saved (input of fit)
  tribal preprocess --data "${LongRead_dir}Immcantation/8_PostIgBlastFilter/TRIBAL_tree/${file}_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass_TRIBAL_SeqData.csv"  \
    --roots "${LongRead_dir}Immcantation/8_PostIgBlastFilter/TRIBAL_tree/${file}_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass_TRIBAL_RootData.csv" \
    --encoding "PATH/Immcantation/8_PostIgBlastFilter/TRIBAL_tree/20241028_HomoSapiens_isotypes_inSpatialVDJ_order.txt" \
    --min-size 2 \
    --cores 3 \
    --out "${LongRead_dir}Immcantation/8_PostIgBlastFilter/TRIBAL_tree/${file}/clonotypes.pkl" \
    --heavy
  
  echo "run fit"
  ## Then run the fit command
  ### -c : clonotypes.pkl file output from prepaprocess command
  ### --cores : number of thread
  ### --transmat-infer: the inferred isotype transition probability matrix.
  ### --pickle : a dictionary with clonotype id as key and value is a LineageTreeList containing the optimal lineage trees for each clonotype.
  ### --score : a csv file containing the SHM parsimony scores and CSR likelihood.
  ### --write-results : a directory where the lineage trees files
  ### --mode :refinement (recommended) : the full algorithm where the CSR likelihood is optimized by solving the most parsimonious tree refinement problem. score : the input parsimony lineage trees are not refined and isotypes of the internal nodes are inferred using weighted parsimony via the Sankoff algorithm, with the isotype transition probabilities as weights
  ###  --n_isotypes : the number of isotypes states to use
  #  --stay-prob
  #  --transmat
  #  --niter
  #  --thresh
  #  --max-cand
  #  --seed
  #  --restarts
  tribal fit -c "${LongRead_dir}Immcantation/8_PostIgBlastFilter/TRIBAL_tree/${file}/clonotypes.pkl" \
  --n_isotypes 10 \
  --mode refinement \
  --cores 10 \
  --transmat-infer "${LongRead_dir}Immcantation/8_PostIgBlastFilter/TRIBAL_tree/${file}/transmat.txt" \
  --pickle "${LongRead_dir}Immcantation/8_PostIgBlastFilter/TRIBAL_tree/${file}/lineage_trees.pkl" \
  --write-results "${LongRead_dir}Immcantation/8_PostIgBlastFilter/TRIBAL_tree/${file}/results" \
  --score "${LongRead_dir}Immcantation/8_PostIgBlastFilter/TRIBAL_tree/${file}/objective.csv"


done


```



## Final anonymized objects 

### Clonotypes merged airr dataset with clone identifier - Anonymized

```{r,eval=FALSE}

SampleId <- Anonym_table$SpatialVDJ_id

IDmatch <- Anonym_table$SpatialGEX_id
names(IDmatch) <- SampleId

Ano_corresp <- setNames(Anonym_table$Anonym_id,nm = Anonym_table$SpatialVDJ_id) # not shared

airr_list <- list()

for(file in FileID){
    file_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass.tsv")
    ## Load airr 
    tmp <- read_rearrangement(file_path)
    
    ##### anonym id
    ano_id <- Ano_corresp[[file]]

    ## adapt sb to match Seurat combined object
    tmp$sb <- paste(paste0(tmp$sb,"-1"),ano_id,sep = "_")
    
    ## Create a cell column by concatenating the SB and the clonotype_id
    tmp$cell <- paste(tmp$sb,tmp$clonotype_id,sep = "-")
    tmp$cell<-gsub("_","-",tmp$cell)
    ## Add the string "Clone_" before the clone_id to match the required format of TRIBAL
    tmp$clone_id <- paste0("Clone_",tmp$clone_id)
    
    # Add sampleid column with anonymized id
    tmp$sampleid <- ano_id
    
    # store in list
    airr_list[[file]]<-tmp
}    

airr<-Reduce(rbind,airr_list)

# Write merged table
write_rearrangement(airr,file=paste0(res_folder,"20260721_airr_igblast-db_igh_clonal_grouped_full_database.tsv"))   

```


### Clones table statistics - Anonymized

```{r,eval=FALSE}
SampleId <- Anonym_table$SpatialVDJ_id

IDmatch <- Anonym_table$SpatialGEX_id
names(IDmatch) <- SampleId

Ano_corresp <- setNames(Anonym_table$Anonym_id,nm = Anonym_table$SpatialVDJ_id) # not shared

isotype_palette <- c("IGHM"="#253494", "IGHD"="#1D91C0", "IGHG1"="#d62728", "IGHG2"="#ff7f0e",  "IGHG3"="#FF6EC7", "IGHG4"="#FCCF05", "IGHA1"="#2ca02c", "IGHA2"="#36e40a","IGHE"="#8c564b","IGH_NA"="grey")

Order <- c("IGHM","IGHD","IGHG3","IGHG1","IGHA1","IGHG2","IGHG4","IGHE","IGHA2","IGH_NA")

my_colors <- c("#66b803", "#f94144", "#f8961e", "#3574b8", "#DE77AE", "#9970AB", "#D5E4A2", "#71D0F5", "#B1746F", "#ADE2D0", "#20DE8BFF", "#CCDE8BFF", "#f7f31c", "#FFA88BFF", "#FF6A8BFF")

bin<-"No_bin"# MetaTopic No_bin

# Generate descriptive data.frame for each sample at clone level (Nb cell, Nb isotype, Meta topic distribution, isotype ratio, etc). And then merge in one full table for all sample
list_file <- list()
for(file in FileID){
  #Load the Seurat object with the metadata containing the MetaTopic info
  Seurat <- readRDS(paste0("PATH/",file,"_seurat_sct_BayesSpace_Starfysh_TLSsign_SpaTopic_CellTopic_",bin,"_MetaTopicMOVICS",".rds"))
  SpotsMetaTopics <-Seurat$MetaTopic
  ##Palette of MetaTopics
  MetaTopPal <- my_colors[1:length(levels(Seurat$MetaTopic))]
  names(MetaTopPal)<-levels(Seurat$MetaTopic)#unify the color of metatopic
  
   # PAth of the final IGH clone with germline reconstruct database
    file_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass.tsv")
    ## Load airr
    airr <- read_rearrangement(file_path)
    ## Create a cell column by concatenating the SB and the clonotype_id
    airr$cell <- paste(airr$sb,airr$clonotype_id,sep = "-")
    airr$cell<-gsub("_","-",airr$cell)
    ## Add the string "Clone_" before the clone_id to match the required format of TRIBAL
    airr$clone_id <- paste0("Clone_",airr$clone_id)
    CloneID <- unique(airr$clone_id) # get unique id
    
    # Calculate combined R and S mutation frequencies/number for each clonotype
    airr <- shazam::observedMutations(airr, sequenceColumn="sequence_alignment",
                                germlineColumn="germline_alignment_d_mask",
                                regionDefinition=NULL,
                                frequency=TRUE, 
                                combine=TRUE,
                                nproc=6)
    
    airr <- shazam::observedMutations(airr, sequenceColumn="sequence_alignment",
                                germlineColumn="germline_alignment_d_mask",
                                regionDefinition=NULL,
                                frequency=FALSE, 
                                combine=TRUE,
                                nproc=6)
    
    # recompute align seq (VDJ) length
    airr$seq_aligned_length <-stri_count(airr[["sequence_alignment"]],regex="[^-.N]")
    
    ## Create a data frame containing the number of cell per IGH clone
    df <- data.frame(sort(table(airr$clone_id),decreasing = T))
    colnames(df)<-c("clone_id","Nb_cell")
   
    
    ## Compute clone frequency in the tumor
    df$Clone_frequency <- df$Nb_cell/sum(df$Nb_cell)
    
    ## Number of unique clonotype per clone
    Uniq_clonotype_per_clone<-tapply(airr$clonotype_id, airr$clone_id, function(clonotype)length(unique(clonotype)))
    df$Uniq_clonotype <-Uniq_clonotype_per_clone[as.character(df$clone_id)]
    
    ## NB of unique isotype per clone
    ### correct if IGH_NA, then not an isotype :: count 0
    Uniq_isotype_per_clone<-tapply(airr$isotype, airr$clone_id, function(iso){
      iso<-iso[iso!="IGH_NA"] # remove IGH_NA
      length(unique(iso))
    })
    df$Uniq_iso <-Uniq_isotype_per_clone[as.character(df$clone_id)]

    
    ## NB of unique isotype per clone
    ### correct if IGH_NA, then not an isotype :: count 0
    Uniq_isotype_per_clone<-tapply(airr$isotype, airr$clone_id, function(iso){
      iso<-iso[iso!="IGH_NA"] # remove IGH_NA
      length(unique(iso))
    })
    df$Uniq_iso <-Uniq_isotype_per_clone[as.character(df$clone_id)]

    ## attribute mutational 
    # If mutiple mutationnal freq per clonotype_id, then consider the one of the longest vdj alignment (the more accurate ...)
    # For a clone id, count the max mu_freq of all its clonotype / average accross clonotypes
    mutational_freq_cln<-tapply(airr[,c("clonotype_id","mu_freq","mu_count","seq_aligned_length")], airr$clone_id, function(mut_df){
      #select only one value of mu_freq per clonotype
      one_mu_freq_per_clono<-tapply(mut_df,mut_df$clonotype_id,function(clono_df){
        unique_maxLen_mu<-unique(clono_df[clono_df$seq_aligned_length==max(clono_df$seq_aligned_length),]$mu_freq)
        if(length(unique_maxLen_mu)==1){
          return(unique_maxLen_mu)
        }else{
          warning("Multiple value of mu_freq for same seq len (max) in one clonotype - return mean")
          return(mean(unique_maxLen_mu))
        }
      })
      #select only one value of mu_count per clonotype
      one_mu_count_per_clono<-tapply(mut_df,mut_df$clonotype_id,function(clono_df){
        unique_maxLen_mu<-unique(clono_df[clono_df$seq_aligned_length==max(clono_df$seq_aligned_length),]$mu_count)
        if(length(unique_maxLen_mu)==1){
          return(unique_maxLen_mu)
        }else{
          warning("Multiple value of mu_freq for same seq len (max) in one clonotype - return mean")
          return(mean(unique_maxLen_mu))
        }
      })
      
      
      #return max and average
      return(data.frame(max_mu_freq=max(one_mu_freq_per_clono),
             mean_mu_freq=mean(one_mu_freq_per_clono),
             max_mu_count=max(one_mu_count_per_clono),
             mean_mu_count=mean(one_mu_count_per_clono)))
    },simplify = T)
    toname<-names(mutational_freq_cln)
    mutational_freq_cln<-Reduce(rbind,mutational_freq_cln)
    mutational_freq_cln$clone_id<-toname
    
    df<-merge(df,mutational_freq_cln,by="clone_id")
    df<-df[order(df$Nb_cell,decreasing = T),]
    
    ## Get the correspondance between SB and MetaTopics 
    MetaCorresp <- Seurat$MetaTopic
     
    # Apply for each clone, take the list of SB, and return a vector of length levels of Metatopics and count the number of each SB in each MetaTop
    res <- sapply((setNames(levels(df$clone_id),levels(df$clone_id))),function(cln){
      sb_cln <- airr[airr$clone_id %in% cln,]$sb
      sb_cln<-paste0(sb_cln,"-1") # adapt to Visium format
      #get meta in clone
      meta_cln <- MetaCorresp[sb_cln]
      table(meta_cln)
    })
    res<-(t(res))
    res_prop <- res/rowSums(res)
    res <- data.frame(res)
    res$clone_id <- factor(rownames(res),levels = rownames(res))
    res_prop <- data.frame(res_prop)
    res_prop$clone_id <- factor(rownames(res_prop),levels = rownames(res_prop))
    
    ### Merge res_prop with original df to aggregate multiple variable per clone 
    df <- merge(df,res_prop,by="clone_id")
    df <- df[order(df$Nb_cell,decreasing = T),]
    
    #melt res_prop and res
    df_count <- reshape::melt(res,id.vars=c("clone_id"))
    df_prop <- reshape::melt(res_prop,id.vars=c("clone_id"))
    
    # Script provided but not included in final article : plot == distribution of Meta Topic for each clone ordered by Nb of Cell in clone (decreasing)
    # p1 <-ggplot(df_count) +
    #     geom_bar(stat = 'identity', position = 'stack',aes(x = clone_id, y = value,fill=variable))+
    #     theme_bw()+
    #     labs(y="SB count in IGH clone",
    #          x=NULL,
    #          fill="MetaTopics")+
    #     theme(axis.text.x = element_blank())+
    #     scale_fill_manual(values = MetaTopPal)
    # p2 <-ggplot(df_prop) +
    #     geom_bar(stat = 'identity', position = 'stack',aes(x = clone_id, y = value,fill=variable))+
    #     theme_bw()+
    #     labs(y="SB proportion in IGH clone",
    #          x="Unique IGH clone",
    #          fill="MetaTopics")+
    #     theme(axis.text.x = element_blank())+
    #     scale_fill_manual(values = MetaTopPal)
    # 
    # patch<-(p1/p2)+plot_layout(guides = 'collect') + plot_annotation(title = file)
    # ggsave(plot = patch,
    #          filename =  paste0(res_folder,file,"_StackBarplot_MetaTopicMOVICS_noBin_IGH_clone_expansion_AllClones",".pdf"),
    #          width = 10,height = 5)
    # 
  
    ###############################
    # Same but now count each IGH isotypes rather than MetaTopics
    # Apply for each clone, take the list of SB, and return a vector of length levels of Metatopics and count the number of each SB in each MetaTop
    res <- table(airr$clone_id,airr$isotype)
    res_prop <- res/rowSums(res)
    res <- data.frame(res)
    colnames(res)<-c("clone_id","variable","value")
    res$clone_id <- factor(as.character(res$clone_id),levels = levels(df$clone_id))
    res$variable <- factor(as.character(res$variable),levels = Order)
    
    res_prop <- data.frame(res_prop)
    colnames(res_prop)<-c("clone_id","variable","value")
    res_prop$clone_id <- factor(as.character(res_prop$clone_id),levels = levels(df$clone_id))
    res_prop$variable <- factor(as.character(res_prop$variable),levels = Order)
    
    ##### wider res_prop to aggregate with previous df 
    res_prop_wid <- pivot_wider(res_prop,names_from = "variable",values_from = "value")
    res_prop_wid<-as.data.frame(res_prop_wid)
    df <- merge(df,res_prop_wid,by="clone_id")
    df<-df[order(df$Nb_cell,decreasing = T),]
    
    # res_prop and res
    df_count <- res
    df_prop <- res_prop
    
    # similarly , this script is provided but not included : distribution of isotype / clone
    # p1 <-ggplot(df_count) +
    #     geom_bar(stat = 'identity', position = 'stack',aes(x = clone_id, y = value,fill=variable))+
    #     theme_bw()+
    #     labs(y="Isotype count in IGH clone",
    #          x=NULL,
    #          fill="Isotype")+
    #     theme(axis.text.x = element_blank())+
    #     scale_fill_manual(values = isotype_palette)
    # p2 <-ggplot(df_prop) +
    #     geom_bar(stat = 'identity', position = 'stack',aes(x = clone_id, y = value,fill=variable))+
    #     theme_bw()+
    #     labs(y="Isotype proportion in IGH clone",
    #          x="Unique IGH clone",
    #          fill="Isotype")+
    #     theme(axis.text.x = element_blank())+
    #     scale_fill_manual(values = isotype_palette)
    # 
    # patch<-(p1/p2)+plot_layout(guides = 'collect') + plot_annotation(title = file)
    # ggsave(plot = patch,
    #          filename =  paste0(res_folder,file,"_StackBarplot_Isotypes_IGH_clone_expansion_AllClones",".pdf"),
    #          width = 10,height = 5)
    # 
 
    ## Export df for each sample to create a merged plot
     ##### anonym id
    ano_id <- Ano_corresp[[file]]
    df$sampleid<-ano_id
    list_file[[file]] <-df
}  

# merge all df
## add IGH_NA and IGHE column for each and fill with 0
list_file <- lapply(list_file,function(table){
  if(!("IGHE" %in% colnames(table))){
    table$IGHE <- 0
  }
  if(!("IGH_NA" %in% colnames(table))){
    table$IGH_NA <- 0
  }
  table
})

# reorder col and merge
colOrder<-colnames(list_file[[1]])
list_file<-lapply(list_file,function(table)table[,colOrder])

# Create a merged plot
list_file<-lapply(list_file,function(table)table[order(table$Nb_cell),])
Merged_df <- Reduce(rbind,list_file)

## count number of uniq metatopic per clone/ presence in MetaTopic M3 (TLS)
Merged_df$Uniq_metaTop <-rowSums(Merged_df[,c("M1",  "M2",  "M3", "M4",  "M5",  "M6", "M7" )]!=0)
Merged_df$M3_metaTop_pos <-ifelse(Merged_df[,c("M3")]!=0,yes = "M3_pos",no = "M3_neg")
Merged_df$M3_metaTop_pos<-factor(Merged_df$M3_metaTop_pos,levels = c("M3_neg","M3_pos"))

# correct NB iso ; if 0, then only "isotype" is IGH_NA, then count 1 (because there is at least 1 isotype ...). But do not consider IGH_NA as an added isotype (not included for CSR) if there are other
Merged_df$Uniq_iso[Merged_df$Uniq_iso==0]<-1

## Reorganize column to have first sampleid, clone_id, etc
Merged_df <- Merged_df[,c("sampleid",setdiff(colnames(Merged_df),"sampleid"))]

write.csv(Merged_df,file = paste0(res_folder,"20260722_SpatialVDJ_CloneStatistics_reproduce_Figure3.csv"),row.names = F)
```


# Reload merged dataset

Reload data

```{r}
# Seurat
Srt_combined <- readRDS(paste0("PATH/Figure2_S6_S7/","20260721_Seurat_merged_SpatialGEX_annotated_metadata.rds")) # Seurat object generated in previous script


# ORder IGH iso
Order <- c("IGHM","IGHD","IGHG3","IGHG1","IGHA1","IGHG2","IGHG4","IGHE","IGHA2","IGH_NA")


# airr dataset for SpatialVDJ containing clone identifier. Dataset merged for all samples (and anonymized)
airr_merged <- read_rearrangement(paste0(res_folder,"20260721_airr_igblast-db_igh_clonal_grouped_full_database.tsv"))
## Level isotypes
airr_merged$isotype <- factor(airr_merged$isotype,levels = c(Order,setdiff(unique(airr_merged$isotype),Order)))
## level sample id
airr_merged$sampleid<-factor(airr_merged$sampleid,levels = Anonym_table$Anonym_id)  


# Clone_Merged_df is the clones level statistics
Clone_Merged_df <- read.csv(paste0(res_folder,"20260722_SpatialVDJ_CloneStatistics_reproduce_Figure3.csv"))


```


# Figures

## Figure 3A - Experimental scheme (biorender)

**Figure 3A compressed** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"Figure3A_to_render_compressed.pdf"),density = 300)
plot(img)
```

## Figure 3B - Clonotype distribution for one selected clone

```{r}
isotype_palette <- c("IGHM"="#253494", "IGHD"="#1D91C0", "IGHG1"="#d62728", "IGHG2"="#ff7f0e",  "IGHG3"="#FF6EC7", "IGHG4"="#FCCF05", "IGHA1"="#2ca02c", "IGHA2"="#36e40a","IGHE"="#8c564b","IGH_NA"="grey")

Order <- c("IGHM","IGHD","IGHG3","IGHG1","IGHA1","IGHG2","IGHG4","IGHE","IGHA2","IGH_NA")

my_colors <- c("#66b803", "#f94144", "#f8961e", "#3574b8", "#DE77AE", "#9970AB", "#D5E4A2", "#71D0F5", "#B1746F", "#ADE2D0", "#20DE8BFF", "#CCDE8BFF", "#f7f31c", "#FFA88BFF", "#FF6A8BFF")

# bin<-"No_bin" # from previous optimization ...

# Following script initially supposed to loop on all file and all clone but here it will be used only to reproduce the Main Figure 3B (tumor SP_LHD6US and clone == 490)

file<-"SP_LHD6US"
# for(file in Anonym_table$Anonym_id){
  #Airr path and load airr
  # PAth of the final IGH clone with germline reconstruct database
  # file_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass.tsv")
  # ## Load airr 
  # airr <- read_rearrangement(file_path)
  
  airr <- airr_merged[airr_merged$sampleid %in% file,] # get back the sample level table by subsetting on anonymized sample id
  
  CloneID <- unique(airr$clone_id) # get unique id
  
  # Number of cell per clone_id
  SortCln<-sort(table(airr$clone_id),decreasing = T)
  
  #Load the Seurat object with the metadata containing the MetaTopic info
  # Seurat <- readRDS(paste0("PATH/",file,"_seurat_sct_BayesSpace_Starfysh_TLSsign_SpaTopic_CellTopic_",bin,"_MetaTopicMOVICS",".rds"))
  Seurat <-  suppressWarnings(subset(Srt_combined,subset = file_id == file)) # retrieve the sample level seurat by subsetting with anonymized id metadata
  Seurat@images <- Seurat@images[file] # remvove all other images
  SpotsMetaTopics <-Seurat$MetaClusters
  ##Palette of MetaTopics
  MetaTopPal <- my_colors[1:length(levels(Seurat$MetaClusters))]
  names(MetaTopPal)<-levels(Seurat$MetaClusters)#unify the color of metatopic
  
  ggplot_list<-list()
  
  #iterate on each clone ID
  #### adapt to only run relevent clone
  id<-"Clone_490"
  # for(id in names(head(SortCln,n=100))){
    ## get the information on clonotype for the corresponding clone
    airr_sub_clone<-data.frame(airr[airr$clone_id %in% id,])
    rownames(airr_sub_clone)<-airr_sub_clone$cell #names according to cell unique id
    
    #number of sequences/cells
    n<-nrow(airr_sub_clone)
    #########  Plot as in Meylan et al : spatial pie plot
    CellID_list <- unique(airr_sub_clone$clonotype_id) #get the list of cell Id from one clonal family
  
    #Concatenate all Cell id within one spot - Generate a spot data.frame
    ClonalFamily<-tapply(airr_sub_clone$clonotype_id,airr_sub_clone$sb,paste,collapse=",")
    # names(ClonalFamily)<-paste0(names(ClonalFamily),"-1")
    ClonalFamily<-as.data.frame(ClonalFamily)

    colnames(ClonalFamily) <- id
    
    Seurat_Cln <- AddMetaData(object= Seurat,
                          metadata = ClonalFamily)
    SeuratSub <- AddMetaData(Seurat_Cln,!is.na(Seurat_Cln@meta.data[,id]),col.name="Sub")
    SeuratSub <- suppressWarnings(subset(SeuratSub,subset = Sub == "TRUE"))
    
    #Change colors according to previous plot
    ##Get the color of the ggplot (stack overflow :https://stackoverflow.com/questions/8197559/emulate-ggplot2-default-color-palette)
    gg_color_hue <- function(n) {
      hues = seq(15, 375, length = n + 1)
      hcl(h = hues, l = 65, c = 100)[1:n]
    }
    
    CheckPlot <- SpatialDimPlot(SeuratSub, group.by = id,crop = F,pt.size.factor = 1.5,image.alpha = 0.5)+theme(plot.margin = margin(0, 0, 0, 0, "cm"),legend.key.size = unit(0.4, 'cm'),legend.direction = "vertical",legend.title = element_text(size = 9),legend.text = element_text(size = 8))
    
    
    #### Reproduce plot from MaxMey Immunity 
    TissueCoordinates <- GetTissueCoordinates(object = Seurat_Cln)
     #transform 
              y <- TissueCoordinates$imagerow
              y_mid <- min(y)+0.5*(max(y)-min(y))
              coords_y <- y-y_mid # centre sur le milieu
              coords_y<- -coords_y # on inverse tout
              coords_y <- coords_y+y_mid #on repasse sur une echelle 
              TissueCoordinates$imagerow <-coords_y 
    
    #Create a matrix of spot X clonotype (unique clonotype) 
    scatter_df <-table(airr_sub_clone$sb,airr_sub_clone$clonotype_id)
    # rownames(scatter_df)<-paste0(rownames(scatter_df),"-1")
    scatter_df<-data.frame(pivot_wider(data.frame(scatter_df),names_from = "Var2",values_from = "Freq"))
    rownames(scatter_df)<-scatter_df$Var1
    scatter_df<-(scatter_df[,setdiff(colnames(scatter_df),"Var1"),drop=F])
    
     ## If there is only 1 clonotype within the clone (multiple cell accross the slide) : Add a "Unselected" column to scatter_df or it won't work with geom_scatterpie...
    if(ncol(scatter_df)==1){
      scatter_df$Unselected <- 0
    }
    
    #Combine coordinates, in the right order, and clonotypes count in each spot
    scatter_df <- cbind(TissueCoordinates[rownames(scatter_df),],scatter_df)
    scatter_df$region <- factor(1:nrow(scatter_df))
    
    transparent_col <- "#00000000"
    #list of metadata to highlight with corresponding barcodes
   ## Create a dummy list with names being the clonotypes and content being one valid barcode in the dataset but no matter which one 
    ##### In case the barcode for which multiple clonotypes are foudn (ie 1 spot with mutiple clonotypes), It leads to issue in the label. Knowing to_select is just used to create a good labeling, I modify to_select so that there is no overlapping barcode ... 

    to_select <- lapply(1:length(CellID_list),function(j){
      colnames(Seurat)[j]
    })
    names(to_select)<-CellID_list
    
    p <- SpatialPlot(Seurat_Cln,
                     image.alpha = 1,
                     pt.size.factor =0,
                     alpha=1,
                     cells.highlight = to_select,
                     label.size = 3,
                     cols.highlight = c(rainbow(length(to_select)),transparent_col))
    p <- p  + geom_scatterpie_edit(data=scatter_df, aes(y=imagerow, x=imagecol,group=region),
                              #pie_scale = 0.8, # to increase or decrease the spot size
                              cols=colnames(scatter_df[,3:(ncol(scatter_df)-1)]))
    p <- p + ggtitle(paste0(id))+labs(fill="Clonotypes")

    ### THIS IS MAIN PLOT 3B
    ggsave(plot = p,
           filename = paste0(res_folder,"Figure3B_",file,"_",id,"_Spatial_PiePlot_clonotypes_n",n,"cells",".pdf"),
           width = 6,height = 4)
    
 # } # loop clone
#} # loop file  
```


**Figure 3B** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"Figure3B_SP_LHD6US_Clone_490_Spatial_PiePlot_clonotypes_n54cells.pdf"),density = 300)
plot(img)
```


## Figure 3C - Clonal expansion in all SpatialVDJ

Count the number of cells (combination of a unique clonotype in a unique SB) per IGH clone as well as the number of cells per isotype per IGH clones. 
Represent as a stack barplot for number and proportion. 
And add a tag information for TLS positivity.

```{r}
## Expansion basic plot
Merged_df_melt<-Clone_Merged_df[,c("sampleid","clone_id","Nb_cell")]
Merged_df_melt <- reshape::melt.data.frame(Merged_df_melt,id.vars = c("sampleid","clone_id"))

# Merged_df_melt$sampleid<- factor(Merged_df_melt$sampleid,levels = Anonym_table$Anonym_id)
Merged_df_melt$sampleid<- factor(Merged_df_melt$sampleid,levels = c( "SP_LHD6US",
                                                                     "SP_0V2O6G",
                                                                     "SP_0O0B5S",
                                                                     "SP_R9MS02")) # specify directly level to get same order as initially ...


plot<-ggplot(Merged_df_melt,aes(x=sampleid,y=value,fill=value))+
  geom_bar(position="stack", stat="identity",color="black",linewidth = 0.02)+
  scale_fill_viridis_c(trans = "log", direction = +1,labels = prettyNum,breaks=base_breaksLog10())+
  theme_bw()+
  labs(x=NULL,
       y="Number of cell (unique clonotype/spot)",
       title = "Clonal expansion as number of cells in IGH clone",
       fill="Cell/clone")

ggsave(plot = plot,
             filename =  paste0(res_folder,"Figure3C_Barplot_ClonalExpansion_CellNumber_In_IGHclone",".pdf"),
             width = 8,height = 6)
```

**Figure 3C** 
```{r,fig.height=8,fig.width=8}
img <- magick::image_read_pdf(paste0(res_folder,"Figure3C_Barplot_ClonalExpansion_CellNumber_In_IGHclone.pdf"),density = 300)
plot(img)
```



## Figure 3D - Experimental scheme (biorender)

**Figure 3D compressed** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"Figure3D_to_render_compressed.pdf"),density = 300)
plot(img)
```


## Figure 3E, 3F, 3G - Clonal expansion SHM, CSR vs TLS cluster

```{r}
## Clone present in TLS at least one time vs other
p1<-ggplot(Clone_Merged_df,aes(x=M3_metaTop_pos,y=Clone_frequency))+
  geom_violin(aes(fill=M3_metaTop_pos),trim = T)+
  geom_boxplot(color="black",fill="white",width=0.1)+
  geom_jitter(size=0.5)+
  # scale_color_manual(values = c("M3_pos"="#f8961e","M3_neg"="black"))+
  scale_y_continuous(trans = "log10",labels = prettyNum,breaks = base_breaksLog10())+
  #scale_fill_viridis_c()+
  theme_bw()+
  theme(legend.position = "none")+
  labs(x=NULL,
       y="IGH clone frequency\n(IGH unique cells in clone/total IGH cell)",
       title = "Clonal expansion")+
  stat_compare_means()

p2<-ggplot(Clone_Merged_df,aes(x=M3_metaTop_pos,y=mean_mu_freq))+
  geom_violin(aes(fill=M3_metaTop_pos),trim = T)+
  geom_boxplot(color="black",fill="white",width=0.1)+
  # geom_jitter()+
  # scale_color_manual(values = c("M3_pos"="#f8961e","M3_neg"="black"))+
  # scale_y_continuous(trans = "log10",labels = prettyNum,breaks = base_breaksLog10())+
  #scale_fill_viridis_c()+
  theme_bw()+
  theme(legend.position = "none")+
  labs(x=NULL,
       y="IGH clone mean mutational frequency",
       title = "SHM")+
  stat_compare_means()


Clone_Merged_df$CSR<-ifelse(Clone_Merged_df$Uniq_iso>1,yes = "> 1 isotype",no = "only 1 isotype")
Clone_Merged_df$CSR <- factor(Clone_Merged_df$CSR,levels = c("only 1 isotype","> 1 isotype"))

p3<-plot_contingency(Clone_Merged_df,
                     variables = c("CSR","M3_metaTop_pos"),
                     create_labels = F,
                     test_name = "fisher.test",
                     title = "CSR")
p3<-p3[[1]]

tabl_M3_IGH<-table(Clone_Merged_df$M3_metaTop_pos)

patch<-((p1|p2|p3))+plot_annotation(title = "IGH clones with at least one cell in TLS (MetaTopic M3) : TLS-positive (M3_pos)",
                                    subtitle = paste(paste("Number of IGH clones: ",names(tabl_M3_IGH),tabl_M3_IGH,sep = "=",collapse = ";")))


ggsave(plot = patch,
             filename =  paste0(res_folder,"Figure3EFG_MetaTopic_violin_ClonalExpansion_and_mutation_and_CSR_inTLS",".pdf"),
             width = 10,height = 5)

```

**Figure 3E, 3F, 3G** 
```{r,fig.height=7,fig.width=10}
img <- magick::image_read_pdf(paste0(res_folder,"Figure3EFG_MetaTopic_violin_ClonalExpansion_and_mutation_and_CSR_inTLS.pdf"),density = 300)
plot(img)
```



## Figure S8A - Experimental scheme (biorender)

**Figure S8A compressed** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"S8A_to_render_compressed.pdf"),density = 300)
plot(img)
```


## Figure S8B and S8C - IGH clonotype forms (BCR or antibody)


```{r}
SampleId <- Anonym_table$SpatialVDJ_id

IDmatch <- Anonym_table$SpatialGEX_id
names(IDmatch) <- SampleId

IDmatchAnonym <- Anonym_table$Anonym_id
names(IDmatchAnonym) <- SampleId

# this was the initial script / here airr will be obtained from the merged clonotypes table shared
# airr<-list()
# 
# for(file in FileID){
#   
#   print(file)
#   
#   # Do not need to adapt from here - path of Igblast quality filtered database
#   file_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI.tsv")
#   
#   # Load airr 
#   tmp<-read_rearrangement(file_path)
#   tmp$sampleid<-file
#   airr[[file]] <- tmp
#   
# }  
# airr<-Reduce(rbind,airr)
# airr$sampleid <- setNames(IDmatchAnonym[airr$sampleid],NULL)
# airr$sampleid<-factor(airr$sampleid,levels = Anonym_table$Anonym_id)


airr <- airr_merged

###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### 
###### Number & proportion of unique clonotype in each isotype / form
###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### 

## Form proportion for each isotype (only for IGH) - per patient
#### Simplify form by grouping ambiguous and incomplete in ambiguous
airr$c_imgtref_form <- ifelse(airr$c_imgtref_form %in% c("Incomplete","Form_NA"),
                              yes = "Ambiguous" ,
                              no = airr$c_imgtref_form) # simplify form

Sub <- tapply(airr$clonotype_id,airr[,c("c_imgtref_form","isotype","sampleid")],function(x)length(unique(x)))
Sub <- apply(Sub,3,function(matrix){
    tmp<-data.frame(matrix)
    tmp[is.na(tmp)] <- 0 # if count is NA it means 0 counts in this case... (for IGHE)
    tmp
})
Sub <- lapply(names(Sub),function(id){
  tmp<-Sub[[id]]
  tmp$c_imgtref_form <- rownames(tmp)
  rownames(tmp)<-NULL
  tmp$sampleid<-id
  tmp
})
Sub <- Reduce(rbind,Sub)
Sub <- reshape::melt.data.frame(Sub,id.vars = c("sampleid","c_imgtref_form"))
# Sub$sampleid <- factor(Sub$sampleid,levels =Anonym_table$Anonym_id )
Sub$sampleid <- factor(Sub$sampleid,levels =c( "SP_LHD6US",
                                               "SP_0V2O6G",
                                               "SP_0O0B5S",
                                               "SP_R9MS02") ) # specify directly level to get same order as initially ...
Sub <- Sub[grepl("IGH",Sub$variable),]#remove all non IGH
Order <- c("IGHM","IGHD","IGHG3","IGHG1","IGHA1","IGHG2","IGHG4","IGHE","IGHA2","IGH_NA")# sort according to order of IGH on the genome
Sub$variable <- factor(as.character(Sub$variable),levels = Order)
      pal <- c("IGHM"="#253494", "IGHD"="#1D91C0", "IGHG1"="#d62728", "IGHG2"="#ff7f0e",  "IGHG3"="#FF6EC7", "IGHG4"="#FCCF05", "IGHA1"="#2ca02c", "IGHA2"="#36e40a","IGHE"="#8c564b","IGH_NA"="grey")

Sub$c_imgtref_form<- factor(Sub$c_imgtref_form,levels = c("Secreted","Membrane","Ambiguous"))

Sub <- pivot_wider(Sub,names_from = c_imgtref_form,values_from = value,id_cols = c("sampleid","variable")) 


## PLot number of unique clonotype in membrane form for each isotype
PlotMembrane <- ggplot(Sub,aes(x=variable,y=Membrane))+
  geom_boxplot(fill="red")+
  geom_point()+
  geom_line(aes(group=sampleid,color=sampleid))+
  theme_bw()+
  labs(x=NULL,
       y="Membrane form count")

## Convert in numeric according to order of factor for the scatterpie function
Sub$sampleid <-as.numeric(Sub$sampleid)
Sub$variable <-as.numeric(Sub$variable)

PlotScatter<-ggplot(Sub,aes(y=sampleid, x=variable))+
  scatterpie::geom_scatterpie(data = Sub,mapping = aes(y=sampleid, x=variable),cols=c("Secreted","Membrane","Ambiguous"),
                  pie_scale = 2)+ # to increase or decrease the spot size)
  #scale_y_continuous(trans = "log10",breaks = base_breaksLog10(),labels = prettyNum)+
  theme_bw()+
  scale_fill_manual(values = c(Ambiguous="grey",Secreted="#A6CEE3",Membrane="red"))+
  scale_x_continuous(breaks=seq(1,length(Order),1),labels=Order)+
  # scale_y_continuous(breaks=seq(1,length(Anonym_table$Anonym_id),1),labels=Anonym_table$Anonym_id)+
  scale_y_continuous(breaks=seq(1,length(Anonym_table$Anonym_id),1),labels=c( "SP_LHD6US",
                                               "SP_0V2O6G",
                                               "SP_0O0B5S",
                                               "SP_R9MS02"))+
  labs(x=NULL,
       y="Sample id",
       fill="IGH form")

Patch<-(PlotScatter/PlotMembrane)+plot_layout(heights = c(2,1))

ggsave(plot = Patch,
       filename = paste0(res_folder,"S8BC_IGH_Isotype_scatterPlot_and_Count_MembraneSecretedForm.pdf"),
       width = 8,
       height = 5)


```



**Figure S8B and S8C** 
```{r,fig.height=7,fig.width=10}
img <- magick::image_read_pdf(paste0(res_folder,"S8BC_IGH_Isotype_scatterPlot_and_Count_MembraneSecretedForm.pdf"),density = 300)
plot(img)
```


## Figure 3H and S8D - Spatial plot with meta topic and clone distribution

Plot each IGH clone in spatial plot with MetaTopic in background

```{r}
isotype_palette <- c("IGHM"="#253494", "IGHD"="#1D91C0", "IGHG1"="#d62728", "IGHG2"="#ff7f0e",  "IGHG3"="#FF6EC7", "IGHG4"="#FCCF05", "IGHA1"="#2ca02c", "IGHA2"="#36e40a","IGHE"="#8c564b","IGH_NA"="grey")

Order <- c("IGHM","IGHD","IGHG3","IGHG1","IGHA1","IGHG2","IGHG4","IGHE","IGHA2","IGH_NA")

my_colors <- c("#66b803", "#f94144", "#f8961e", "#3574b8", "#DE77AE", "#9970AB", "#D5E4A2", "#71D0F5", "#B1746F", "#ADE2D0", "#20DE8BFF", "#CCDE8BFF", "#f7f31c", "#FFA88BFF", "#FF6A8BFF")

# bin<-"No_bin" # from previous optimization ...

# Following script initially supposed to loop on all file and all clone but here it will be used only to reproduce :
### The Main Figure 3H (tumor SP_LHD6US and clone == 370)
### The Supplementary Figure S8D (tumor SP_R9MS02 and clone == 9)

for(file in c("SP_LHD6US","SP_R9MS02")){
  #Airr path and load airr
  # PAth of the final IGH clone with germline reconstruct database
  # file_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass.tsv")
  # ## Load airr 
  # airr <- read_rearrangement(file_path)
  
  airr <- airr_merged[airr_merged$sampleid %in% file,] # get back the sample level table by subsetting on anonymized sample id
  
  CloneID <- unique(airr$clone_id) # get unique id
  
  # Number of cell per clone_id
  SortCln<-sort(table(airr$clone_id),decreasing = T)
  
  #Load the Seurat object with the metadata containing the MetaTopic info
  # Seurat <- readRDS(paste0("PATH/",file,"_seurat_sct_BayesSpace_Starfysh_TLSsign_SpaTopic_CellTopic_",bin,"_MetaTopicMOVICS",".rds"))
  Seurat <-  suppressWarnings(subset(Srt_combined,subset = file_id == file)) # retrieve the sample level seurat by subsetting with anonymized id metadata
  Seurat@images <- Seurat@images[file] # remvove all other images
  SpotsMetaTopics <-Seurat$MetaClusters
  ##Palette of MetaTopics
  MetaTopPal <- my_colors[1:length(levels(Seurat$MetaClusters))]
  names(MetaTopPal)<-levels(Seurat$MetaClusters)#unify the color of metatopic
  
  ggplot_list<-list()
  #iterate on each clone ID
  #### here adapted to run only on selected clone
  
  id <- ifelse(file=="SP_LHD6US",yes = "Clone_370",no = "Clone_9") # Clone370 for SP_LHD6US / Clone_9 for SP_R9MS02
  
  # for(id in names(head(SortCln,n=100))){
    ## get the information on clonotype for the corresponding clone
    airr_sub_clone<-data.frame(airr[airr$clone_id %in% id,])
    rownames(airr_sub_clone)<-airr_sub_clone$cell #names according to cell unique id
    
    #number of sequences/cells
    n<-nrow(airr_sub_clone)
    #########  Plot as in Meylan et al : spatial pie plot
    CellID_list <- unique(airr_sub_clone$clonotype_id) #get the list of cell Id from one clonal family
  
    #Concatenate all Cell id within one spot - Generate a spot data.frame
    ClonalFamily<-tapply(airr_sub_clone$clonotype_id,airr_sub_clone$sb,paste,collapse=",")
    # names(ClonalFamily)<-paste0(names(ClonalFamily),"-1")
    ClonalFamily<-as.data.frame(ClonalFamily)

    colnames(ClonalFamily) <- id
    
    Seurat_Cln <- AddMetaData(object= Seurat,
                          metadata = ClonalFamily)
    SeuratSub <- AddMetaData(Seurat_Cln,!is.na(Seurat_Cln@meta.data[,id]),col.name="Sub")
    SeuratSub <- suppressWarnings(subset(SeuratSub,subset = Sub == "TRUE"))
    
    #Change colors according to previous plot
    ##Get the color of the ggplot (stack overflow :https://stackoverflow.com/questions/8197559/emulate-ggplot2-default-color-palette)
    gg_color_hue <- function(n) {
      hues = seq(15, 375, length = n + 1)
      hcl(h = hues, l = 65, c = 100)[1:n]
    }
    
    CheckPlot <- SpatialDimPlot(SeuratSub, group.by = id,crop = F,pt.size.factor = 1.5,image.alpha = 0.5)+theme(plot.margin = margin(0, 0, 0, 0, "cm"),legend.key.size = unit(0.4, 'cm'),legend.direction = "vertical",legend.title = element_text(size = 9),legend.text = element_text(size = 8))

    
    #### Reproduce plot from MaxMey Immunity 
    TissueCoordinates <- GetTissueCoordinates(object = Seurat_Cln)
     #transform 
              y <- TissueCoordinates$imagerow
              y_mid <- min(y)+0.5*(max(y)-min(y))
              coords_y <- y-y_mid # centre sur le milieu
              coords_y<- -coords_y # on inverse tout
              coords_y <- coords_y+y_mid #on repasse sur une echelle 
              TissueCoordinates$imagerow <-coords_y 
    
    #Create a matrix of spot X clonotype (unique clonotype) 
    scatter_df <-table(airr_sub_clone$sb,airr_sub_clone$clonotype_id)
    # rownames(scatter_df)<-paste0(rownames(scatter_df),"-1")
    scatter_df<-data.frame(pivot_wider(data.frame(scatter_df),names_from = "Var2",values_from = "Freq"))
    rownames(scatter_df)<-scatter_df$Var1
    scatter_df<-(scatter_df[,setdiff(colnames(scatter_df),"Var1"),drop=F])
    
     ## If there is only 1 clonotype within the clone (multiple cell accross the slide) : Add a "Unselected" column to scatter_df or it won't work with geom_scatterpie...
    if(ncol(scatter_df)==1){
      scatter_df$Unselected <- 0
    }
    
    #Combine coordinates, in the right order, and clonotypes count in each spot
    scatter_df <- cbind(TissueCoordinates[rownames(scatter_df),],scatter_df)
    scatter_df$region <- factor(1:nrow(scatter_df))
    
    transparent_col <- "#00000000"
    #list of metadata to highlight with corresponding barcodes
   ## Create a dummy list with names being the clonotypes and content being one valid barcode in the dataset but no matter which one 
    ##### In case the barcode for which multiple clonotypes are foudn (ie 1 spot with mutiple clonotypes), It leads to issue in the label. Knowing to_select is just used to create a good labeling, I modify to_select so that there is no overlapping barcode ... 

    to_select <- lapply(1:length(CellID_list),function(j){
      colnames(Seurat)[j]
    })
    names(to_select)<-CellID_list
    
    p <- SpatialPlot(Seurat_Cln,
                     image.alpha = 1,
                     pt.size.factor =0,
                     alpha=1,
                     cells.highlight = to_select,
                     label.size = 3,
                     cols.highlight = c(rainbow(length(to_select)),transparent_col))
    p <- p  + geom_scatterpie_edit(data=scatter_df, aes(y=imagerow, x=imagecol,group=region),
                              #pie_scale = 0.8, # to increase or decrease the spot size
                              cols=colnames(scatter_df[,3:(ncol(scatter_df)-1)]))
    p <- p + ggtitle(paste0(id))+labs(fill="Clonotypes")


    # Plot Isotypes
    
    CorrespIso <- airr[airr$clonotype_id  %in% CellID_list,c("clonotype_id","isotype")] # Get isotypes
    CorrespIso <- CorrespIso[!duplicated(CorrespIso$clonotype_id),] # remove duplicated cells
    
    # isotype as character otherwise it do not work properly
    CorrespIso$isotype <- as.character(CorrespIso$isotype)
    
    ## apply per isotype, retrieve all clonotypes for corresponding isotypes :
    ### if the clonotypes is/are in the spot, then 1 for the isotype / else 0
    scatter_iso<-tapply(CorrespIso$clonotype_id,CorrespIso$isotype,function(Clonotypes){
              tmp<-scatter_df[,Clonotypes,drop=F]
              tmp<-rowSums(tmp)
              bc<-names(tmp)
              setNames(as.numeric(tmp>0),nm = bc)
          },simplify = F)
    IsoColNames <- names(scatter_iso)
    scatter_iso <- data.frame(Reduce(cbind,scatter_iso)) #data.frame in case there is only 1 iso, else it does nothing
    colnames(scatter_iso)<-IsoColNames
    
    #First try
    ## If there is only 1 isotype : Add a "Unselected" column to scatter_iso or it won't work with geom_scatterpie...
    if(ncol(scatter_iso)==1){
      scatter_iso$Unselected <- 0
    }
    
    scatter_iso <- cbind(scatter_df[,c("imagerow","imagecol","region")],scatter_iso[rownames(scatter_df),])
    
    # change the levels of MetaTopic
    Seurat_Cln$MetaClusters<-factor(Seurat_Cln$MetaClusters,levels = c(levels(Seurat_Cln$MetaClusters),names(isotype_palette)))
    
    p <- SpatialPlot(Seurat_Cln,
                     image.alpha = 0,
                     pt.size.factor =2,
                     alpha = 0.6,
                     group.by = "MetaClusters")
    p <- p  + geom_scatterpie_edit(data=scatter_iso, aes(y=imagerow, x=imagecol,group=region),
                              #pie_scale = 2, # to increase or decrease the spot size
                              cols=colnames(scatter_iso)[4:(ncol(scatter_iso))])+
      scale_fill_manual(values = c(MetaTopPal,isotype_palette))
    p <- p + ggtitle(paste0(id))+labs(fill="Isotypes\n&\nMetaClusters")

    
  ### These  are Main 3H or Sup S8D depending on sample/clone  
    prefixe <- ifelse(file=="SP_LHD6US",yes = "Figure3H_",no = "S8D_")
      
    ggsave(plot = p,
           filename =  paste0(res_folder,prefixe,file,"_",id,"_Spatial_PiePlot_isotypes_and_MetaTopics_n",n,"cells",".pdf"),
           width = 6,height = 4)

  # } # end of clone loop
  
} #end of file loop
  
```



**Figure 3H** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"Figure3H_SP_LHD6US_Clone_370_Spatial_PiePlot_isotypes_and_MetaTopics_n67cells.pdf"),density = 300)
plot(img)
```


**Figure S8D** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"S8D_SP_R9MS02_Clone_9_Spatial_PiePlot_isotypes_and_MetaTopics_n42cells.pdf"),density = 300)
plot(img)
```



## Figure 3I and S8E - iGraph lineage tree reconstruction and shaping

Reconstruct the graph built in python with R package iGraph in order to adapt the size, shape and color of nodes.

Note : the script was initially built to run on all samples and all graph. It has been adapted to run on the 2 main TRIBAL graph included in the article (but everything is still available if one want to reproduce all graph ...).

```{r,eval=FALSE}
library(ggraph)
library(tidygraph)
SampleId <- Anonym_table$SpatialVDJ_id

IDmatch <- Anonym_table$SpatialGEX_id
names(IDmatch) <- SampleId


isotype_palette <- c("IGHM"="#253494", "IGHD"="#1D91C0", "IGHG1"="#d62728", "IGHG2"="#ff7f0e",  "IGHG3"="#FF6EC7", "IGHG4"="#FCCF05", "IGHA1"="#2ca02c", "IGHA2"="#36e40a","IGHE"="#8c564b","IGH_NA"="grey")

Order <- c("IGHM","IGHD","IGHG3","IGHG1","IGHA1","IGHG2","IGHG4","IGHE","IGHA2","IGH_NA")

my_colors <- c("#66b803", "#f94144", "#f8961e", "#3574b8", "#DE77AE", "#9970AB", "#D5E4A2", "#71D0F5", "#B1746F", "#ADE2D0", "#20DE8BFF", "#CCDE8BFF", "#f7f31c", "#FFA88BFF", "#FF6A8BFF")

#V1 - to generate all plot in all condition tested and all samples ...

##### Choose No_Bin CellTopics and the MOVICS consensus clustering MetaTopics with k=7
# run for both No_bin and Bin for consistency in plot (pipeline already run, overide the Bin plot from another clustering strategy ...) but finally consider only No_hin
# BinOrNot<-c(No_bin="No_bin",Bin="Bin") 
# 
# for(bin in BinOrNot){
#   TribalPlot<-list()
#   for(file in FileID){
#     #Airr path and load airr
#     # PAth of the final IGH clone with germline reconstruct database
#     file_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass.tsv")
#     ## Load airr 
#     airr <- read_rearrangement(file_path)
#     ## Create a cell column by concatenating the SB and the clonotype_id
#     airr$cell <- paste(airr$sb,airr$clonotype_id,sep = "-")
#     airr$cell<-gsub("_","-",airr$cell)
#     ## Add the string "Clone_" before the clone_id to match the required format of TRIBAL
#     airr$clone_id <- paste0("Clone_",airr$clone_id)
#     
#     #Load the Seurat object with the metadata containing the MetaTopic info
#     Seurat <- readRDS(paste0("PATH/",file,"_seurat_sct_BayesSpace_Starfysh_TLSsign_SpaTopic_CellTopic_",bin,"_MetaTopicMOVICS",".rds"))
#     SpotsMetaTopics <-Seurat$MetaTopic
#     ##Palette of MetaTopics
#     MetaTopPal <- my_colors[1:length(levels(Seurat$MetaTopic))]
#     names(MetaTopPal)<-levels(Seurat$MetaTopic)#unify the color of metatopic
#     
#     # Path of results of TRIBAL
#     tribal_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/TRIBAL_tree/",file,"/results/")
#     ## Get names of Clones 
#     CloneID <- list.files(tribal_path) # get all files in the results folders
#     CloneID <- gsub("Clone_","Clone-",CloneID) # change the first _ to -
#     CloneID <- gsub("_.*","",CloneID) # remove everything after the _
#     CloneID <- gsub("-","_",CloneID) # get back to - in _
#     CloneID <- unique(CloneID) # get unique id
#     
#     ggplot<-list()
#     #iterate on each clone ID
#     for(id in CloneID){
#       #Load edges
#       edges <- read.delim(paste0(tribal_path,id,"_edge_list0.txt"),header = F,sep = ",")
#       colnames(edges)<-c("from","to")
#       
#       #Load nodes
#       ##Load nodes and isotypes
#       nodes <- read.delim(paste0(tribal_path,id,"_isotypes0.csv"),header = F,sep = ",")
#       colnames(nodes)<-c("id","isotype")
#       nodes$isotype<- factor(nodes$isotype,levels = Order) #order according to genome
#       ## get the information on clonotype for the corresponding clone
#       airr_sub_clone<-data.frame(airr[airr$clone_id %in% id,])
#       rownames(airr_sub_clone)<-airr_sub_clone$cell #names according to cell unique id
#       ## Add umi_count for each node
#       nodes$umi_count <-airr_sub_clone[nodes$id,"umi_count"]
#       ## Add secreted or membrane form for each node
#       nodes$form <-airr_sub_clone[nodes$id,"c_imgtref_form"]
#       nodes$form <- ifelse(is.na(nodes$form),yes = NA,
#                           no = ifelse(nodes$form %in% "Secreted",yes = "S",
#                         no = ifelse(nodes$form %in% "Membrane",yes = "M",no = "?")))
#       nodes$form<-factor(nodes$form,levels=c("S","M","?"))
#       
#       #specified the form scale
#       Form_scale <- c(S=16,M=17,"?"=15)
#       
#       ## Retrieve spot Name and add the MetaTopic info to node data.frame
#       nodes$spot_id <- gsub("-Group.*","-1",nodes$id)
#       nodes$MetaTopic<-SpotsMetaTopics[nodes$spot_id]
# 
#       
#       #Create the graph
#       mygraph<-tidygraph::tbl_graph(nodes = nodes, edges = edges,directed = T)
#       
#       #number of sequences/cells
#       n<-sum(!is.na(nodes$umi_count))
#       
#       plot<-ggraph(mygraph, layout = "tree") + 
#         geom_edge_diagonal()+
#         geom_node_point(aes(fill=isotype),shape=23,size=2)+
#         geom_node_point(aes(color=MetaTopic,shape=form),size=4)+
#         #geom_node_point(aes(color=nodes$BayesSpace),size = 5) + 
#         #geom_node_text(aes(label = form), vjust = 1.5) + 
#         theme_graph(base_family = "")+
#         #theme_void()+
#         # theme(panel.grid = element_blank(),
#         #         axis.text = element_blank(),
#         #         axis.ticks  = element_blank(),
#         #         axis.title = element_blank())+
#         scale_fill_manual(values = isotype_palette)+
#         scale_color_manual(values = MetaTopPal)+
#         scale_shape_manual(values = Form_scale)+
#         labs(title = paste("TRIBAL isotype aware tree"),subtitle = paste(file,"-",id,"\n","n","=",n,"cells (clonotypes/spots)"),
#              fill="Isotype")
#       #Plot the graph
#       ggplot[[id]]<-list(plot=plot,
#         n=n)
#       
#       ## Save the plot
#       ggsave(plot = ggplot[[id]]$plot,
#            filename =paste0(res_folder,"04_Clone_IgBlast_in_MetaTopics/TRIBAL_tree/MetaTopic_",bin,"/",file,"/","TRIBAL_Tree0_iGraph_n",n,"cells_",id,".pdf"),
#            width = 15,height = 6)
#     }
#     TribalPlot[[file]]<-ggplot
#   }
#   
#   #Order by number of seq by clone
#   Reorder<-lapply(TribalPlot,function(id_list){
#       names(sort(sapply(id_list,function(clone)clone$n),decreasing = T))
#   })
#   TribalPlot<-lapply(FileID,function(id_tum){
#     tmp<-TribalPlot[[id_tum]]
#     ord<-Reorder[[id_tum]]
#     lapply(tmp[ord],function(newlist)newlist$plot)
#   })
# 
# }

#V2 : only to reproduce the selected plot
bin<-"No_bin"

# Following script initially supposed to loop on all file and all clone but here it will be used only to reproduce :
### The Main Figure 3H (tumor SP_LHD6US and clone == 370)
### The Supplementary Figure S8D (tumor SP_R9MS02 and clone == 9)

for(file in c("SP_LHD6US","SP_R9MS02")){
  #Airr path and load airr
  # PAth of the final IGH clone with germline reconstruct database
  # file_path <- paste0(ImmcantationFolder,"8_PostIgBlastFilter/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_igh-form_Collapse-clonotype-split-SB_Min-2-UMI_IGH-dist_clone-pass_germ-pass.tsv")
  # ## Load airr 
  # airr <- read_rearrangement(file_path)
  # ## Create a cell column by concatenating the SB and the clonotype_id
  # airr$cell <- paste(airr$sb,airr$clonotype_id,sep = "-")
  # airr$cell<-gsub("_","-",airr$cell)
  # ## Add the string "Clone_" before the clone_id to match the required format of TRIBAL
  # airr$clone_id <- paste0("Clone_",airr$clone_id)
  
  airr <- airr_merged[airr_merged$sampleid %in% file,] # get back the sample level table by subsetting on anonymized sample id
  
  #Load the Seurat object with the metadata containing the MetaTopic info
  # Seurat <- readRDS(paste0("PATH/",file,"_seurat_sct_BayesSpace_Starfysh_TLSsign_SpaTopic_CellTopic_",bin,"_MetaTopicMOVICS",".rds"))
  Seurat <-  suppressWarnings(subset(Srt_combined,subset = file_id == file)) # retrieve the sample level seurat by subsetting with anonymized id metadata
  Seurat@images <- Seurat@images[file] # remvove all other images
  SpotsMetaTopics <-Seurat$MetaClusters
  ##Palette of MetaTopics
  MetaTopPal <- my_colors[1:length(levels(Seurat$MetaClusters))]
  names(MetaTopPal)<-levels(Seurat$MetaClusters)#unify the color of metatopic
  
  # Path of results of TRIBAL
  tribal_path <- paste0(res_folder,"TRIBAL_to_reproduce_tree/",file,"/")

  ggplot<-list()
  #iterate on each clone ID
  #### here adapted to run only on selected clone
  id <- ifelse(file=="SP_LHD6US",yes = "Clone_370",no = "Clone_9") # Clone370 for SP_LHD6US / Clone_9 for SP_R9MS02
  
  # for(id in CloneID){
    #Load edges
    edges <- read.delim(paste0(tribal_path,id,"_edge_list0.txt"),header = F,sep = ",")
    colnames(edges)<-c("from","to")
    
    #Load nodes
    ##Load nodes and isotypes
    nodes <- read.delim(paste0(tribal_path,id,"_isotypes0.csv"),header = F,sep = ",")
    colnames(nodes)<-c("id","isotype")
    nodes$isotype<- factor(nodes$isotype,levels = Order) #order according to genome
    ## get the information on clonotype for the corresponding clone
    airr_sub_clone<-data.frame(airr[airr$clone_id %in% id,])
    
    airr_sub_clone$cell <- gsub("-1-..-......-","-",airr_sub_clone$cell) # adjust cell_id so that it is similar to what was used in input of TRIBAL (individual, not merged for all samples, so it do not contain the name of the sample in the cell id ...)
    
    rownames(airr_sub_clone)<-airr_sub_clone$cell #names according to cell unique id
    ## Add umi_count for each node
    nodes$umi_count <-airr_sub_clone[nodes$id,"umi_count"]
    ## Add secreted or membrane form for each node
    nodes$form <-airr_sub_clone[nodes$id,"c_imgtref_form"]
    nodes$form <- ifelse(is.na(nodes$form),yes = NA,
                        no = ifelse(nodes$form %in% "Secreted",yes = "S",
                      no = ifelse(nodes$form %in% "Membrane",yes = "M",no = "?")))
    nodes$form<-factor(nodes$form,levels=c("S","M","?"))
    
    #specified the form scale
    Form_scale <- c(S=16,M=17,"?"=15)
    
    ## Retrieve spot Name and add the MetaTopic info to node data.frame
    nodes$spot_id <- gsub("-Group.*","-1",nodes$id)
    
    nodes$spot_id <- paste(nodes$spot_id,file,sep = "_") # add this to match name of spot in the merged file ... (with name of the file ...)
    
    nodes$MetaClusters<-SpotsMetaTopics[nodes$spot_id]

    
    #Create the graph
    mygraph<-tidygraph::tbl_graph(nodes = nodes, edges = edges,directed = T)
    
    #number of sequences/cells
    n<-sum(!is.na(nodes$umi_count))
    
    plot<-ggraph(mygraph, layout = "tree") + 
      geom_edge_diagonal()+
      geom_node_point(aes(fill=isotype),shape=23,size=2)+
      geom_node_point(aes(color=MetaClusters,shape=form),size=4)+
      #geom_node_point(aes(color=nodes$BayesSpace),size = 5) + 
      #geom_node_text(aes(label = form), vjust = 1.5) + 
      theme_graph(base_family = "")+
      #theme_void()+
      # theme(panel.grid = element_blank(),
      #         axis.text = element_blank(),
      #         axis.ticks  = element_blank(),
      #         axis.title = element_blank())+
      scale_fill_manual(values = isotype_palette)+
      scale_color_manual(values = MetaTopPal)+
      scale_shape_manual(values = Form_scale)+
      labs(title = paste("TRIBAL isotype aware tree"),subtitle = paste(file,"-",id,"\n","n","=",n,"cells (clonotypes/spots)"),
           fill="Isotype")

    ## Save the plot
      ### These  are Main 3I or Sup S8E depending on sample/clone  
    prefixe <- ifelse(file=="SP_LHD6US",yes = "Figure3I_",no = "S8E_")
    ggsave(plot = plot,
         filename =paste0(res_folder,prefixe,file,"_",id,"_TRIBAL_Tree0_iGraph_n",n,"cells",".pdf"),
         width = 15,height = 6)
    
  # }

}  
```



**Figure 3I** 
```{r,fig.height=7,fig.width=15}
img <- magick::image_read_pdf(paste0(res_folder,"Figure3I_SP_LHD6US_Clone_370_TRIBAL_Tree0_iGraph_n67cells.pdf"),density = 300)
plot(img)
```


**Figure S8E** 
```{r,fig.height=7,fig.width=15}
img <- magick::image_read_pdf(paste0(res_folder,"S8E_SP_R9MS02_Clone_9_TRIBAL_Tree0_iGraph_n42cells.pdf"),density = 300)
plot(img)
```



# SessionInfo

```{r}
sessionInfo()
```
