---
title: "Immunoverse Immcantation Clonal analysis - Ig repertoire, IgG and IgA tumor cells and response to ICI"
author: "Guilhem Pupier"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmdformats::readthedown:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Paired **Tumor**, **PBMC before ICI treatment (T0)** and **after 6 weeks of ICI treatment (T6)**, stored frozen, were processed to extract RNA, generate Immunoverse libraries, and compare the total Ig repertoire between these samples and according to **ICI treatment response (R= responders, NR= non responders)**. A total of 90 paired samples from 30 patients (12NR + 18R) were analyzed together with an additional tumor from a NR patient lacking the paired blood samples, as well as 5 blood samples from **healthy donors (HD)** used as reference.

Paired information of IgG and IgA coverage on tumor cells was available for a subset of 20 patients (60 samples) + 1 tumor only (without PBMC).

Goal :

- To analyze Ig repertoire correlation with IgG/IgA coverage on tumor cells  and response to ICI.  

Sections :

- **Preprocess** :  
All pre-process steps are aailables in previous script "Figure4".    

- **Reload merged dataset** :  
Reload data to generate plots.

- **Figures** : 
Script to reproduce Figure 5 and S11.  

# Load required packages


```{r,eval=T}
library(airr)
library(alakazam)
library(shazam) # airr data packages

library(data.table) # function fread() for efficient and fast read of tab file

library(stringi)
library(stringr) # string manipulations

library(dplyr) 
library(tidyr)
library(reshape2) # table manipulations

library(MESS) # compute AUC

library(pastecs)
library(RColorBrewer)
library(viridis) # color palettes
library(patchwork)
library(ggplot2)
library(ggpubr)
library(grid)
library(gridExtra)
library(scales) # plot functions
#from Heather Turner @stackoverflow ## Adapt y scale in log scale 
base_breaksLog10 <- function(n = 10){
  function(x) {
    axisTicks(log10(range(x, na.rm = TRUE)), log = T, n = n)
  }
}
base_breaksLog1p <- function(n = 10){
  function(x) {
    axisTicks(log1p(range(x, na.rm = TRUE)), log = T, n = n)
  }
}

# library(qs) # optimize (faster than rds) writing and reading of large dataset in R

# plot contingency (fro Maxime Meylan)
plot_contingency <- function(data,
           variables = c("TLS_pos", "HBV"),
           hide_legend = F,
           text_size=6,
           text_position_side=T,
           create_labels = T,
           inverse_factor2=F,
           order_factor=NULL,
           palette = NA,
           geom_label=F,
           test_name = c("chisq.test", "fisher.test")[1],
           labs = c(xlab = "",
                    ylab = "",
                    legend_title = ""),
           hide_test = F,
           title = ""){
    if(inverse_factor2){
      data[,variables[2]] <- factor(data[,variables[2]],levels=rev(levels(data[,variables[2]])))
    }
    cont_table <- table(data[,variables[1]],data[,variables[2]])
    #clean data table to avoid empty rows
    non_empty_rows <- rowSums(cont_table)!=0
    cont_table <- cont_table[non_empty_rows,]

    if(create_labels){
      #sapply(levels(factor(data[,variables[2]])), function(x) paste(variables[2],x,sep="_"))
      colnames(cont_table) <- sapply(levels(factor(data[,variables[2]])), function(x) paste(variables[2],x,sep="_"))
      rownames(cont_table) <- sapply(levels(factor(data[,variables[1]])), function(x) paste(variables[1],x,sep="_"))
    }
  if(test_name != "fisher.test" & test_name != "chisq.test"){
    stop("incorrect test, valid options are fisher.test or chisq.test")
  }
  if(test_name == "fisher.test"){
    error_test <-  if(class(try(fisher.test(cont_table),silent=T)) == "try-error"){
      test_res <- fisher.test(cont_table,simulate.p.value = T)
    }else{
      test_res <- fisher.test(cont_table)
    }
  }
  if(test_name == "chisq.test"){
    test_res <- chisq.test(cont_table)
  }
  #make sure there is a matrix (enough values in both sides)
  if(!is.null(dim(cont_table))){
    cont_table_perc <-apply(cont_table,1,function(x) {
      res <- x / sum(x) *100
    })
    cont_table_m <- melt(cont_table_perc)
    cont_table_m$Var1 <- factor(cont_table_m$Var1)
    p <- ggplot(data=cont_table_m, aes(x=Var2,y=value,fill=Var1,label=value))
    p <- p + geom_bar(stat="identity",width=0.95)
    #to reverse order of bars
    #p <- p + geom_bar(stat="identity",width=0.95,position = position_stack(reverse = TRUE))
    
    if(geom_label){
      if(text_position_side){
        p <- p + geom_label(color="white",label = paste0(as.vector(t(cont_table))," (",format(cont_table_m$value,digits = 1),"%)"),
                       position = position_stack(vjust = .5),size=text_size)
      }else{
        p <- p + geom_label(color="white",label = paste0(as.vector(t(cont_table)),"\n(",format(cont_table_m$value,digits = 1),"%)"),
                       position = position_stack(vjust = .5),size=text_size)
      }
    }else{
      if(text_position_side){
        p <- p + geom_text(label = paste0(as.vector(t(cont_table))," (",format(cont_table_m$value,digits = 1),"%)"),
                       position = position_stack(vjust = .5),size=text_size)
      }else{
        p <- p + geom_text(label = paste0(as.vector(t(cont_table)),"\n(",format(cont_table_m$value,digits = 1),"%)"),
                       position = position_stack(vjust = .5),size=text_size)
      }
    }
    
    p <- p + theme_minimal()

    if(any(!is.na(palette))){
      p <- p + scale_fill_manual(values = palette)
    }
    #p <- p + theme(legend.title = element_blank())#,axis.text.x = element_text(angle = 25, hjust = 1))
    p <- p + guides(fill=guide_legend(title=labs[3]))
    if(hide_test){
      p <- p + labs(subtitle=paste("p = ",format(test_res$p.value,digits=3),sep=" "),title=title,x=labs[1],y=labs[2])
      if(test_res$p.value < 0.05 ) p <- p + theme(plot.subtitle=element_text(face="bold"))
    }else{
      p <- p + labs(subtitle=paste(test_name, " p = ",format(test_res$p.value,digits=3),sep=" "),title=title,x=labs[1],y=labs[2])
      if(test_res$p.value < 0.05 ) p <- p + theme(plot.subtitle=element_text(face="bold"))
    }

    if(hide_legend){
      p <- p + guides(fill=FALSE)
    }
  }else{
    p <- ggplot()
    test_res <-1
  }
  return(list(p,test_res,cont_table))

}

# Color palettes for consistency on plot
##Get the color of the ggplot (stack overflow :https://stackoverflow.com/questions/8197559/emulate-ggplot2-default-color-palette)
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

## palette of R_NR factor
pal_R_NR_HD <- gg_color_hue(3)
names(pal_R_NR_HD)<-c("NR","R","HD")
## palette used for heatmap will be used for all plots
ColorPal<- list(Arm=c(`Nivo-Ipi`="#6A3D9A",
                      Nivo="#1F78B4",
                      TKI="#FFD92F"),
                R_NR=c(R="#00BA38",
                           NR="#F8766D",
                       HD="#619CFF"),
                sample_storage=c(FFPE="darkgrey",
                                 Frozen="#E7298A"),
                sample_type=c(`Tumor.T0`="black",
                         `PBMC.T0`="#1F78B4",
                         `PBMC.T6`="#E31A1C",
                         `Tonsil.T0`="darkgrey"),
                sample_type2=c(`Tumor.T0`="black",
                         `PBMC.T0`="#1F78B4",
                         `PBMC.T6`="#E31A1C",
                         `PBMC.HD`="#6A3D9A"),
                IgG_cat=c(`high IgG+ tumor cells`="#FF7F00",
                          `low IgG+ tumor cells`="#1F78B4" ),
                IgA_cat=c(`high IgA+ tumor cells`="#FDBF6F",
                          `low IgA+ tumor cells`="#A6CEE3"),
                IgG_IgA_var=c(`IgG_positive_tumor.cells`= "#FF7F00", 
                              `IgA_positive_tumor.cells`="#619CFF" ))



# Give the path of the Immcantation folder. With a back / at the end

# ImmcantationFolder <- "PATH/"

# Output folder to retrieve the files and plot
res_folder <-"PATH/"
```

# Preprocess - Not rendered

See previous script : Figure 4. For preprocess steps necessary to identify clonotypes and clones statistics.   


# Reload merged dataset

Reload data

```{r}
# Metadata ANONYMIZED - NB : Response to ICI is not shared due to patient privacy requirement but can be asked to the study sponsor unpon reasonable request (see article for contact)
Immcantation_meta_ICI_unique <-readRDS(paste0("PATH/","20260727_Immunoverse_Metadata_AnonymId_ICIresponse_IgTumorCells.rds")) # edit : study sponsor accepted to share the information


#reload clone abundance and diversity for all IGH clones
## Same as in previous script (Figure 4)
abund_list <- readRDS(paste0("PATH/","20260724_Immcantation_IGH_clonal_alakazamEstimateAbundance_Unweighted_Uniform_anonymized.rds"))
div <- readRDS(paste0("PATH/","20260724_Immcantation_IGH_clonal_alakazamAlphaDiversity_Unweighted_Uniform_anonymized.rds"))


```



# Figures

## Figure 5A and 5B - Experimental scheme (biorender) and IF stainings

**Figure 5A and 5B compressed > see paper ...** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"Figure5AB_to_render_compressed.pdf"),density = 300)
plot(img)
```

## Figure 5C and S11A - Ig diversity vs IgG or IgA-coverage on tumor cells


```{r}
# Plot diversity R/NR or sample_type for all kind of parameters
## NB : initially a loop on all parameters. Here only the unweighted and uniform=T parameter are selected
# var_to_iterate<-setNames(names(div),nm = names(div)) # modify the vector to iterate on

var<-"Unweighted_UniformlyResampled"

# for(var in var_to_iterate){
  # hill_div <- div[[var]]@diversity
  hill_div <- div
  
  ## Merge back with metadata info
  hill_div<-merge(hill_div,Immcantation_meta_ICI_unique[,c("sample_id","patient_id","sample_origin","sample_timepoint","sample_type","sample_type2","R_NR","IgG_positive_tumor.cells", "IgA_positive_tumor.cells")],by="sample_id")
  
  hill_div<-hill_div[!is.na(hill_div$IgG_positive_tumor.cells),] #remove sample with NA in IHC
  
  ## convert q in a factor rather than a numeric for plot
  hill_div$q<-factor(hill_div$q)
  
 
  # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
  # correlation at specific q
  # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
  
  alpha_q0 <- filter(hill_div, q == 0)
  
  
  # cor and test
  cor_results <- alpha_q0 %>%
    group_by(sample_type,R_NR) %>%
    summarise(
      cor_test = list(cor.test(IgG_positive_tumor.cells, d, method = "spearman")),
      .groups = "drop"
    ) %>%
    mutate(
      rho = sapply(cor_test, function(x) x$estimate),
      p_value = sapply(cor_test, function(x) x$p.value)
    )
  
  ## format label
  cor_results$label <- paste0("rho = ",
                              formatC(cor_results$rho, digits = 3, format = "f"),
                              "\n", # "\n" if I want a line return
                              "p = "
                              ,ifelse(cor_results$p_value < 0.001,
                                       formatC(cor_results$p_value, format = "e", digits = 2),
                                       formatC(cor_results$p_value, digits = 3, format = "f"))
                              )
  
  ## positions
  cor_results$IgG_positive_tumor.cells<-rep(c(25,75),times=3)
  cor_results$d<-max(alpha_q0$d)+(IQR(alpha_q0$d)/3)
  
  
  ##plot
  p<-ggplot(alpha_q0, aes(x = IgG_positive_tumor.cells, y = d)) +
    # geom_vline(xintercept = 60)+
    # geom_hline(yintercept = threshold_div)+
    geom_smooth(aes(color=R_NR),method = "lm")+
    geom_label(data = cor_results,mapping = aes(label=label,color=R_NR),fill="white")+
    geom_point(aes(color=R_NR),size=2)+
    facet_grid(.~sample_type,scales = "free_x",space = "free_x")+
    # annotate("text",x = 50,y =max(alpha_q0$d),label=paste("fisher.test\np-value:",round(res_test$p.value,2)))+
    scale_x_continuous(expand = expansion(mult = 0.1)) +  # ← marge en X
    scale_y_continuous(expand = expansion(mult = 0.1)) +  # ← marge en Y
    labs(title = "Richness (q=0) - Hill numbers", x = "IgG-labelled tumor cells proportion (%)", y = "IGH Richness\n(Hill number with q=0)",
         subtitle = "Tumor treated with ICI with both repertoire (frozen rna) and ihc (matching ffpe block) data",
         caption = var)+  
    scale_color_manual(values = ColorPal$R_NR)+
    theme_minimal()
  
  #### THIS IS FIGURE 5C
  ggsave(plot = p,
         filename = paste0(res_folder,"Figure5C_Immcantation_AlphaDiversity_",var,"_subsetTumorWithIHCdata_RichnessIGHclone_vs_IgGtumorProportion_correlationByResponse.pdf"),
         height = 4,width = 10)
  
  ###################################
  # same but vs IgA_tumor cells
  ###################################
  # cor and test
  cor_results <- alpha_q0 %>%
    group_by(sample_type,R_NR) %>%
    summarise(
      cor_test = list(cor.test(IgA_positive_tumor.cells, d, method = "spearman")),
      .groups = "drop"
    ) %>%
    mutate(
      rho = sapply(cor_test, function(x) x$estimate),
      p_value = sapply(cor_test, function(x) x$p.value)
    )
  
  ## format label
  cor_results$label <- paste0("rho = ",
                              formatC(cor_results$rho, digits = 3, format = "f"),
                              "\n",
                              "p = "
                              ,ifelse(cor_results$p_value < 0.001,
                                       formatC(cor_results$p_value, format = "e", digits = 2),
                                       formatC(cor_results$p_value, digits = 3, format = "f"))
                              )
  
  ## positions
  cor_results$IgA_positive_tumor.cells<-rep(c(25,60),times=3)
  cor_results$d<-max(alpha_q0$d)+(IQR(alpha_q0$d)/3)
  
  
  p<-ggplot(alpha_q0, aes(x = IgA_positive_tumor.cells, y = d)) +
    # geom_vline(xintercept = 60)+
    # geom_hline(yintercept = threshold_div)+
    geom_smooth(aes(color=R_NR),method = "lm")+
    geom_label(data = cor_results,mapping = aes(label=label,color=R_NR),fill="white")+
    geom_point(aes(color=R_NR),size=2)+
    facet_grid(.~sample_type,scales = "free_x",space = "free_x")+
    # annotate("text",x = 50,y =max(alpha_q0$d),label=paste("fisher.test\np-value:",round(res_test$p.value,2)))+
    # scale_x_continuous(expand = expansion(mult = 0.1)) +  # ← marge en X
    scale_y_continuous(expand = expansion(mult = 0.1)) +  # ← marge en Y
    labs(title = "Richness (q=0) - Hill numbers", x = "IgA-labelled tumor cells proportion (%)", y = "IGH Richness\n(Hill number with q=0)",
         subtitle = "Tumor treated with ICI with both repertoire (frozen rna) and ihc (matching ffpe block) data",
         caption = var)+  
    scale_color_manual(values = ColorPal$R_NR)+
    theme_minimal()
  
  ### This is Figure supplementary S11A
  
  ggsave(plot = p,
         filename = paste0(res_folder,"S11A_Immcantation_AlphaDiversity_",var,"_subsetTumorWithIHCdata_RichnessIGHclone_vs_IgAtumorProportion_correlationByResponse.pdf"),
         height = 4,width = 10)
  
  ########################################################################
  ## Same but for all values of q >>>> Not included in the paper but mentionned in the text (data not shared)
  ########################################################################

  # cor and test
  cor_results <- hill_div %>%
    group_by(sample_type,q,R_NR) %>%
    summarise(
      cor_test = list(cor.test(IgG_positive_tumor.cells, d, method = "spearman")),
      .groups = "drop"
    ) %>%
    mutate(
      rho = sapply(cor_test, function(x) x$estimate),
      p_value = sapply(cor_test, function(x) x$p.value)
    )
  
  ## format label
  cor_results$label <- paste0("rho=",
                              formatC(cor_results$rho, digits = 3, format = "f"),
                              "\n",
                              "p="
                              ,ifelse(cor_results$p_value < 0.001,
                                       formatC(cor_results$p_value, format = "e", digits = 2),
                                       formatC(cor_results$p_value, digits = 3, format = "f"))
                              )
  
  ## positions
  cor_results$IgG_positive_tumor.cells<-rep(rep(c(25,75),times=11),times=3)
  # max_per_plot <-
  cor_results$d<-0
  
  
  ## Replace the lbel by a "" if it is not significant / remove p_value from label
  cor_results$label<-ifelse(test =cor_results$p_value>=0.05 ,yes = "",no = gsub("\\np=.*","",cor_results$label))
  

  ##plot
  p<-ggplot(hill_div, aes(x = IgG_positive_tumor.cells, y = d)) +
    # geom_vline(xintercept = 60)+
    # geom_hline(yintercept = threshold_div)+
    geom_smooth(aes(color=R_NR),method = "lm")+
    geom_text(data = cor_results,mapping = aes(label=label,color=R_NR),size=3,nudge_y =-100)+
        # geom_label(data = cor_results,mapping = aes(label=label,color=R_NR),fill="white",size=2,nudge_y =+100)+
    geom_point(aes(color=R_NR),size=2)+
    scale_x_continuous(expand = expansion(mult = 0.1)) +  # ← marge en X
    scale_y_continuous(expand = expansion(mult = 0.1)) +  # ← marge en Y
    labs(title = "", x = "IgG-labelled tumor cells proportion (%)", y = "Clone diversity (Hill number)"
         # subtitle = "Tumor treated with ICI with both repertoire (frozen rna) and ihc (matching ffpe block) data"
         )+  
    scale_color_manual(values = ColorPal$R_NR)+
    facet_grid(q~sample_type,scales = "free_y")+
    theme_minimal()
  
    # ggsave(plot = p,
    #      filename = paste0(res_folder,"Immcantation_AlphaDiversity_",var,"_subsetTumorWithIHCdata_AllQValues_vs_IgGtumorProportion_correlationByResponseV1.pdf"),
    #      height = 16,width = 8)
  
  
#}

```


**Figure 5C** 
```{r,fig.height=10,fig.width=10}
img <- magick::image_read_pdf(paste0(res_folder,"Figure5C_Immcantation_AlphaDiversity_Unweighted_UniformlyResampled_subsetTumorWithIHCdata_RichnessIGHclone_vs_IgGtumorProportion_correlationByResponse.pdf"),density = 300)
plot(img)
```


**Figure S11A** 
```{r,fig.height=10,fig.width=10}
img <- magick::image_read_pdf(paste0(res_folder,"S11A_Immcantation_AlphaDiversity_Unweighted_UniformlyResampled_subsetTumorWithIHCdata_RichnessIGHclone_vs_IgAtumorProportion_correlationByResponse.pdf"),density = 300)
plot(img)
```



## Figure 5D and S11B - Top 5 Ig clone frequency vs IgG or IgA-coverage on tumor cells


```{r}
# Plot diversity R/NR or sample_type for all kind of parameters
## NB : initially a loop on all parameters. Here only the unweighted and uniform=T parameter are selected
# var_to_iterate<-setNames(names(div),nm = names(div)) # modify the vector to iterate on

var<-"Unweighted_UniformlyResampled"

# for(var in var_to_iterate){
#   clone_frequency <- abund_list[[var]]@abundance

  clone_frequency <- abund_list
  
  ## Merge back with metadata info
  clone_frequency<-merge(clone_frequency,Immcantation_meta_ICI_unique[,c("sample_id","patient_id","sample_origin","sample_timepoint","sample_type","sample_type2","R_NR","IgG_positive_tumor.cells", "IgA_positive_tumor.cells")],by="sample_id")
  
  clone_frequency<-clone_frequency[!is.na(clone_frequency$IgG_positive_tumor.cells),] #remove sample with NA in IHC

  ###########################################################################
  # Top clones
  ###########################################################################
  ## top 5
  #### actually initially top 10 but adapt to top5 later ...
  clone_frequency_top10<-clone_frequency
  clone_frequency_top10$rank_character<-as.character(clone_frequency_top10$rank)
  clone_frequency_top10 <- clone_frequency_top10[clone_frequency_top10$rank_character %in% as.character(1:5),]#filter top 10
  clone_frequency_top10$rank_character <- factor(clone_frequency_top10$rank_character,levels = as.character(1:5)) # as factor
  
  ## Correaltion IgG proportion and clone frequency
  
  # cor and test
  cor_results <- clone_frequency_top10 %>%
    group_by(rank_character,sample_type,R_NR) %>%
    summarise(
      cor_test = list(cor.test(IgG_positive_tumor.cells, p, method = "spearman")),
      .groups = "drop"
    ) %>%
    mutate(
      rho = sapply(cor_test, function(x) x$estimate),
      p_value = sapply(cor_test, function(x) x$p.value)
    )
  
  ## format label
  cor_results$label <- paste0("rho = ",
                              formatC(cor_results$rho, digits = 3, format = "f"),
                              "\n",
                              "p = "
                              ,ifelse(cor_results$p_value < 0.001,
                                       formatC(cor_results$p_value, format = "e", digits = 2),
                                       formatC(cor_results$p_value, digits = 3, format = "f"))
                              )
  
  ## positions
  cor_results$IgG_positive_tumor.cells<-ifelse(cor_results$R_NR=="NR",yes=25,no=75)
  cor_results$p<-0.5
  
  p<-ggplot(clone_frequency_top10,aes(x =IgG_positive_tumor.cells,y= p,color=R_NR))+
    geom_point(size=2)+
    geom_smooth(aes(color=R_NR),method = "lm")+
    geom_label(data = cor_results,mapping = aes(label=label,color=R_NR),fill="white",size=3)+
    scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum)+
    # facet_grid(rank_character~.,scales = "free_x",space = "free_x")+
    facet_grid(sample_type~rank_character)+
    theme_minimal()+
    scale_color_manual(values = ColorPal$R_NR)+
    # theme(legend.position = "none")+
    labs(y="Clone frequency",title = "Rank clone abundance - Top 10 clones",
         x="IgG-labelled tumor cells proportion (%)",
         subtitle = "Tumor treated with ICI with both repertoire (frozen rna) and ihc (matching ffpe block) data",
         caption = var)
  
  #### THIS IS FIGURE 5D
  
  ggsave(plot = p,
         filename = paste0(res_folder,"Figure5D_Immcantation_ClonalFrequency_",var,"_Top10clone_SplitTop10_correlation_response_and_IgGproportion.pdf"),
         height = 6,width = 12)
  
  
  
  
  
    ## Correaltion IgA proportion and clone frequency
  
  # cor and test
  cor_results <- clone_frequency_top10 %>%
    group_by(rank_character,sample_type,R_NR) %>%
    summarise(
      cor_test = list(cor.test(IgA_positive_tumor.cells, p, method = "spearman")),
      .groups = "drop"
    ) %>%
    mutate(
      rho = sapply(cor_test, function(x) x$estimate),
      p_value = sapply(cor_test, function(x) x$p.value)
    )
  
  ## format label
  cor_results$label <- paste0("rho = ",
                              formatC(cor_results$rho, digits = 3, format = "f"),
                              "\n",
                              "p = "
                              ,ifelse(cor_results$p_value < 0.001,
                                       formatC(cor_results$p_value, format = "e", digits = 2),
                                       formatC(cor_results$p_value, digits = 3, format = "f"))
                              )
  
  ## positions
  cor_results$IgA_positive_tumor.cells<-ifelse(cor_results$R_NR=="NR",yes=25,no=60)
  cor_results$p<-0.5
  
  p<-ggplot(clone_frequency_top10,aes(x =IgA_positive_tumor.cells,y= p,color=R_NR))+
    geom_point(size=2)+
    geom_smooth(aes(color=R_NR),method = "lm")+
    geom_label(data = cor_results,mapping = aes(label=label,color=R_NR),fill="white",size=3)+
    scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum)+
    # facet_grid(rank_character~.,scales = "free_x",space = "free_x")+
    facet_grid(sample_type~rank_character)+
    theme_minimal()+
    scale_color_manual(values = ColorPal$R_NR)+
    # theme(legend.position = "none")+
    labs(y="Clone frequency",title = "Rank clone abundance - Top 10 clones",
         x="IgA-labelled tumor cells proportion (%)",
         subtitle = "Tumor treated with ICI with both repertoire (frozen rna) and ihc (matching ffpe block) data",
         caption = var)
  
  ## THIS IS SUPPLEMENTARY FIGURE S11B
  ggsave(plot = p,
         filename = paste0(res_folder,"S11B_Immcantation_ClonalFrequency_",var,"_Top10clone_SplitTop10_correlation_response_and_IgAproportion.pdf"),
         height = 6,width = 12)


#}
  
  
```



**Figure 5D** 
```{r,fig.height=10,fig.width=12}
img <- magick::image_read_pdf(paste0(res_folder,"Figure5D_Immcantation_ClonalFrequency_Unweighted_UniformlyResampled_Top10clone_SplitTop10_correlation_response_and_IgGproportion.pdf"),density = 300)
plot(img)
```


**Figure S11B** 
```{r,fig.height=10,fig.width=12}
img <- magick::image_read_pdf(paste0(res_folder,"S11B_Immcantation_ClonalFrequency_Unweighted_UniformlyResampled_Top10clone_SplitTop10_correlation_response_and_IgAproportion.pdf"),density = 300)
plot(img)
```




# SessionInfo

```{r}
sessionInfo()
```
