---
title: "Immunoverse Immcantation Clonal analysis - Tumor, Blood, response to ICI"
author: "Guilhem Pupier"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmdformats::readthedown:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction


Immunoverse bulk Ig (heavy and light chains) targeted capture libraries are sequenced with MiSeq 2\*300 bp :

-   R1 : 8 bases UMI + 13 bases template switch (TS=AACCGCCAGGAGT) + Insert (V\>D\>J\>C)  
-   R2 : Constant primer around 20 bp (gene specific primer = GSP2) + Insert (C\>J\>D\>V)  

Paired **Tumor**, **PBMC before ICI treatment (T0)** and **after 6 weeks of ICI treatment (T6)**, stored frozen, were processed to extract RNA, generate Immunoverse libraries, and compare the total Ig repertoire between these samples and according to **ICI treatment response (R= responders, NR= non responders)**. A total of 90 paired samples from 30 patients (12NR + 18R) were analyzed together with an additional tumor from a NR patient lacking the paired blood samples, as well as 5 blood samples from **healthy donors (HD)** used as reference.

Goal :

- To analyze Tumor clonotypes vs Blood clonotypes before and after treatment (R vs NR)  
- To analyze Tumor clonotypes between R vs NR  

Sections :

- **Preprocess** :  
All pre-process step of **Immunoverse** data following Immcantation pipeline with raw reads preparation, IgBlast alignment, clonotype table clustering, clonal grouping step and lineage tree reconstruction.    

- **Reload merged dataset** :  
Reload data to generate plots.

- **Figures** : 
Script to reproduce Figure 4, S9 and S10.  

# Load required packages


```{r,eval=T}
library(airr)
library(alakazam)
library(shazam) # airr data packages

library(data.table) # function fread() for efficient and fast read of tab file

library(stringi)
library(stringr) # string manipulations

library(dplyr) 
library(tidyr)
library(reshape2) # table manipulations

library(MESS) # compute AUC

library(pastecs)
library(RColorBrewer)
library(viridis) # color palettes
library(patchwork)
library(ggplot2)
library(ggpubr)
library(grid)
library(gridExtra)
library(scales) # plot functions
#from Heather Turner @stackoverflow ## Adapt y scale in log scale 
base_breaksLog10 <- function(n = 10){
  function(x) {
    axisTicks(log10(range(x, na.rm = TRUE)), log = T, n = n)
  }
}
base_breaksLog1p <- function(n = 10){
  function(x) {
    axisTicks(log1p(range(x, na.rm = TRUE)), log = T, n = n)
  }
}

# library(qs) # optimize (faster than rds) writing and reading of large dataset in R

# plot contingency (fro Maxime Meylan)
plot_contingency <- function(data,
           variables = c("TLS_pos", "HBV"),
           hide_legend = F,
           text_size=6,
           text_position_side=T,
           create_labels = T,
           inverse_factor2=F,
           order_factor=NULL,
           palette = NA,
           geom_label=F,
           test_name = c("chisq.test", "fisher.test")[1],
           labs = c(xlab = "",
                    ylab = "",
                    legend_title = ""),
           hide_test = F,
           title = ""){
    if(inverse_factor2){
      data[,variables[2]] <- factor(data[,variables[2]],levels=rev(levels(data[,variables[2]])))
    }
    cont_table <- table(data[,variables[1]],data[,variables[2]])
    #clean data table to avoid empty rows
    non_empty_rows <- rowSums(cont_table)!=0
    cont_table <- cont_table[non_empty_rows,]

    if(create_labels){
      #sapply(levels(factor(data[,variables[2]])), function(x) paste(variables[2],x,sep="_"))
      colnames(cont_table) <- sapply(levels(factor(data[,variables[2]])), function(x) paste(variables[2],x,sep="_"))
      rownames(cont_table) <- sapply(levels(factor(data[,variables[1]])), function(x) paste(variables[1],x,sep="_"))
    }
  if(test_name != "fisher.test" & test_name != "chisq.test"){
    stop("incorrect test, valid options are fisher.test or chisq.test")
  }
  if(test_name == "fisher.test"){
    error_test <-  if(class(try(fisher.test(cont_table),silent=T)) == "try-error"){
      test_res <- fisher.test(cont_table,simulate.p.value = T)
    }else{
      test_res <- fisher.test(cont_table)
    }
  }
  if(test_name == "chisq.test"){
    test_res <- chisq.test(cont_table)
  }
  #make sure there is a matrix (enough values in both sides)
  if(!is.null(dim(cont_table))){
    cont_table_perc <-apply(cont_table,1,function(x) {
      res <- x / sum(x) *100
    })
    cont_table_m <- melt(cont_table_perc)
    cont_table_m$Var1 <- factor(cont_table_m$Var1)
    p <- ggplot(data=cont_table_m, aes(x=Var2,y=value,fill=Var1,label=value))
    p <- p + geom_bar(stat="identity",width=0.95)
    #to reverse order of bars
    #p <- p + geom_bar(stat="identity",width=0.95,position = position_stack(reverse = TRUE))
    
    if(geom_label){
      if(text_position_side){
        p <- p + geom_label(color="white",label = paste0(as.vector(t(cont_table))," (",format(cont_table_m$value,digits = 1),"%)"),
                       position = position_stack(vjust = .5),size=text_size)
      }else{
        p <- p + geom_label(color="white",label = paste0(as.vector(t(cont_table)),"\n(",format(cont_table_m$value,digits = 1),"%)"),
                       position = position_stack(vjust = .5),size=text_size)
      }
    }else{
      if(text_position_side){
        p <- p + geom_text(label = paste0(as.vector(t(cont_table))," (",format(cont_table_m$value,digits = 1),"%)"),
                       position = position_stack(vjust = .5),size=text_size)
      }else{
        p <- p + geom_text(label = paste0(as.vector(t(cont_table)),"\n(",format(cont_table_m$value,digits = 1),"%)"),
                       position = position_stack(vjust = .5),size=text_size)
      }
    }
    
    p <- p + theme_minimal()

    if(any(!is.na(palette))){
      p <- p + scale_fill_manual(values = palette)
    }
    #p <- p + theme(legend.title = element_blank())#,axis.text.x = element_text(angle = 25, hjust = 1))
    p <- p + guides(fill=guide_legend(title=labs[3]))
    if(hide_test){
      p <- p + labs(subtitle=paste("p = ",format(test_res$p.value,digits=3),sep=" "),title=title,x=labs[1],y=labs[2])
      if(test_res$p.value < 0.05 ) p <- p + theme(plot.subtitle=element_text(face="bold"))
    }else{
      p <- p + labs(subtitle=paste(test_name, " p = ",format(test_res$p.value,digits=3),sep=" "),title=title,x=labs[1],y=labs[2])
      if(test_res$p.value < 0.05 ) p <- p + theme(plot.subtitle=element_text(face="bold"))
    }

    if(hide_legend){
      p <- p + guides(fill=FALSE)
    }
  }else{
    p <- ggplot()
    test_res <-1
  }
  return(list(p,test_res,cont_table))

}

# Color palettes for consistency on plot
##Get the color of the ggplot (stack overflow :https://stackoverflow.com/questions/8197559/emulate-ggplot2-default-color-palette)
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

## palette of R_NR factor
pal_R_NR_HD <- gg_color_hue(3)
names(pal_R_NR_HD)<-c("NR","R","HD")
## palette used for heatmap will be used for all plots
ColorPal<- list(Arm=c(`Nivo-Ipi`="#6A3D9A",
                      Nivo="#1F78B4",
                      TKI="#FFD92F"),
                R_NR=c(R="#00BA38",
                           NR="#F8766D",
                       HD="#619CFF"),
                sample_storage=c(FFPE="darkgrey",
                                 Frozen="#E7298A"),
                sample_type=c(`Tumor.T0`="black",
                         `PBMC.T0`="#1F78B4",
                         `PBMC.T6`="#E31A1C",
                         `Tonsil.T0`="darkgrey"),
                sample_type2=c(`Tumor.T0`="black",
                         `PBMC.T0`="#1F78B4",
                         `PBMC.T6`="#E31A1C",
                         `PBMC.HD`="#6A3D9A"),
                IgG_cat=c(`high IgG+ tumor cells`="#FF7F00",
                          `low IgG+ tumor cells`="#1F78B4" ),
                IgA_cat=c(`high IgA+ tumor cells`="#FDBF6F",
                          `low IgA+ tumor cells`="#A6CEE3"),
                IgG_IgA_var=c(`IgG_positive_tumor.cells`= "#FF7F00", 
                              `IgA_positive_tumor.cells`="#619CFF" ))



# Give the path of the Immcantation folder. With a back / at the end

# ImmcantationFolder <- "PATH/"


# Output folder to get the files and plot
res_folder <-"PATH/"
```

# Preprocess - Not rendered

All script bellow are run only one time - not re run for plot/render markdown. 

## Reads preprocess

### Fastp preprocess

-   **Fastp** : Trim adapters + Correct overlapping base (R1+R2) + length filter 60 bp (around 20 bases primers + min 40 bases insert) + Filter low quality reads (Q20)  

```{bash,eval=F}
# in a conda environment : conda activate fastp

cd PATH

# read all the file of 104 lines in array
readarray -t input_fileR1 < input_file_R1paths.txt
readarray -t input_fileR2 < input_file_R2paths.txt
readarray -t output_file < uniq_id_n104samples.txt

#Moove to out_dir : Immcantation_dir
Immcantation_dir="PATH/"

mkdir pipeline_log
mkdir 01_Fastp

#iterate on all 104 samples by positions
## NB : initially 104 samples because some FFPE samples were also included // not in the final article ! 

for i in $(seq 0 103); do
  # get the corresponding paths and output name by position
    file_inR1="${input_fileR1[${i}]}"
    file_inR2="${input_fileR2[${i}]}"
    file_out="${output_file[${i}]}"
    echo "files: $file_inR1 and $file_inR2 and $file_out"
    
    echo "run fastp"
    fastp --in1 $file_inR1 \
    --in2 $file_inR2 \
    --out1 "${Immcantation_dir}01_Fastp/${file_out}_fastp_processed.R1.fastq" \
    --out2 "${Immcantation_dir}01_Fastp/${file_out}_fastp_processed.R2.fastq" \
    --json "${Immcantation_dir}01_Fastp/${file_out}_fastp.json" \
    --html "${Immcantation_dir}01_Fastp/${file_out}_fastp.html" \
    --dont_eval_duplication \
    --correction \
    --overlap_len_require=30 \
    --overlap_diff_limit=5 \
    --overlap_diff_percent_limit=20 \
    --average_qual=20 \
    --length_required=60 \
    --adapter_sequence=AGATCGGAAGAGCACACGTCTGAACTCCAGTCA \
    --adapter_sequence_r2=AGATCGGAAGAGCGTCGTGTAGGGAAAGAGTGT \
    --thread=30 > "${Immcantation_dir}01_Fastp/${file_out}_fastp.log" 2>&1

    echo "complete"
done

# conda deactivate

```

### pRESTO

-   MaskPrimers.py : Identify UMI+TS in R1 and Constant primer in R2  
-   PairSeq.py : pair back R1 and R2 after MaskPrimers.py  
-   AssemblePairs.py sequential : assemble R1 and R2 in one full read to align with IgBlast  
    NB : UMI are not considered at this step because of too low diversity (only 8 bases ...). Collapse duplicates sequences is also not done at this step but rather with Alakazam package, after IgBlast alignment.  
-   Re run Maskprimer.py score to extract Template switch + UMI which can have been re-inserted if R1 and R2 are totally overlapping.  

```{bash,eval=F}

mkdir 02_pRESTO
mkdir 02_pRESTO/01_MaskPrimers
mkdir 02_pRESTO/02_PairSeq
mkdir 02_pRESTO/03_AssemblePairs
mkdir 02_pRESTO/04_MaskPrimers2

# put the script in a sh executable

#!/bin/bash

# Activate conda environment
source PATH/conda.sh  # Adjust path if needed
conda activate pRESTO

Immcantation_dir="PATH/"

cd ${Immcantation_dir}

files=$(cat PATH/uniq_id_n104samples.txt)

# Start timing
start_time=$(date +%s)
start_date=$(date +"%Y-%m-%d %H:%M:%S")
echo ">>> Pipeline started at: $start_date"


for file in $files; do

  echo $file

## Identify UMI, template switch and C primers

### UMI and template switch in R1
echo "Runing MaskPrimers.py score on R1 to identify Barcode-UMI"
    MaskPrimers.py score -s "${Immcantation_dir}01_Fastp/${file}_fastp_processed.R1.fastq" \
    -p PATH/02_pRESTO/TS_R1.fasta \
    --barcode --mode cut \
    --start 8 \
    --maxerror 0.16 \
    --outdir "${Immcantation_dir}02_pRESTO/01_MaskPrimers/" \
    --outname "${file}_R1" \
    --log "${Immcantation_dir}02_pRESTO/01_MaskPrimers/${file}_MP_TS_R1.log" > "${Immcantation_dir}pipeline_log/${file}_pRESTO_perf.log" 2>&1

echo "Runing ParseLog.py on R1 MaskPrimer log to identify Barcode-UMI"
    ParseLog.py -l "${Immcantation_dir}02_pRESTO/01_MaskPrimers/${file}_MP_TS_R1.log" \
    -f ID PRIMER PRSTART BARCODE INSEQ ALIGN OUTSEQ ERROR >> "${Immcantation_dir}pipeline_log/${file}_pRESTO_perf.log" 2>&1

### C primers in R2

echo "Runing MaskPrimers.py score on R2 to identify C primer"
    MaskPrimers.py score -s "${Immcantation_dir}01_Fastp/${file}_fastp_processed.R2.fastq" \
    -p PATH/02_pRESTO/Cprimers.fasta \
    --mode tag \
    --start 0 \
    --maxerror 0.12 \
    --pf ArcherPrimer \
    --outdir "${Immcantation_dir}02_pRESTO/01_MaskPrimers/" \
    --outname "${file}_R2" \
    --log "${Immcantation_dir}02_pRESTO/01_MaskPrimers/${file}_MP_Cregion_R2.log" >> "${Immcantation_dir}pipeline_log/${file}_pRESTO_perf.log" 2>&1

echo "Runing ParseLog.py on R2 MaskPrimer log - C primer"
    ParseLog.py -l "${Immcantation_dir}02_pRESTO/01_MaskPrimers/${file}_MP_Cregion_R2.log" \
    -f ID PRIMER INSEQ ALIGN OUTSEQ ERROR >> "${Immcantation_dir}pipeline_log/${file}_pRESTO_perf.log" 2>&1


## Copy UMI and Primer annotation and synchronize R1 and R2

echo "Runing PairSeq.py on R1 and R2 after MaskPrimer"
    PairSeq.py -1 "${Immcantation_dir}02_pRESTO/01_MaskPrimers/${file}_R1_primers-pass.fastq" \
    -2 "${Immcantation_dir}02_pRESTO/01_MaskPrimers/${file}_R2_primers-pass.fastq" \
    --coord illumina \
    --outdir "${Immcantation_dir}02_pRESTO/02_PairSeq/" >> "${Immcantation_dir}pipeline_log/${file}_pRESTO_perf.log" 2>&1

## Assemble R1 and R2 pairs in one full length read

echo "Runing AssemblePairs.py sequential on consensus R1 and consensus R2"
    AssemblePairs.py sequential -1 "${Immcantation_dir}02_pRESTO/02_PairSeq/${file}_R1_primers-pass_pair-pass.fastq" \
    -2 "${Immcantation_dir}02_pRESTO/02_PairSeq/${file}_R2_primers-pass_pair-pass.fastq" \
    -r PATH/igblast/fasta/imgt_human_ig_v.fasta \
        --coord illumina --rc tail --scanrev --1f BARCODE --2f ARCHERPRIMER \
        --aligner blastn \
        --nproc 30 \
        --outname $file \
        --outdir "${Immcantation_dir}02_pRESTO/03_AssemblePairs/" \
        --log "${Immcantation_dir}02_pRESTO/03_AssemblePairs/${file}_AP.log" >> "${Immcantation_dir}pipeline_log/${file}_pRESTO_perf.log" 2>&1

# Re run Maskprimer.py score to extract Template switch + UMI which can have been re-inserted if R1 and R2 are totally overlapping.

# --failed to keep reads with no UMI-TS ... will be concatenated back with trimmed reads after maskprimers

### UMI and template switch in assembled reads
echo "Runing MaskPrimers.py score on assembled reads to identify Barcode-UMI"
    MaskPrimers.py score -s "${Immcantation_dir}02_pRESTO/03_AssemblePairs/${file}_assemble-pass.fastq" \
    -p PATH/02_pRESTO/TS_R1.fasta \
    --mode cut \
    --start 8 \
    --maxerror 0.16 \
    --failed \
    --outdir "${Immcantation_dir}02_pRESTO/04_MaskPrimers2/" \
    --outname "${file}_clean" \
    --log "${Immcantation_dir}02_pRESTO/04_MaskPrimers2/${file}_MP_TS_assemble_reads.log" >> "${Immcantation_dir}pipeline_log/${file}_pRESTO_perf.log" 2>&1

echo "Runing ParseLog.py on R1 MaskPrimer log to identify Barcode-UMI"
  ParseLog.py -l "${Immcantation_dir}02_pRESTO/04_MaskPrimers2/${file}_MP_TS_assemble_reads.log" \
  -f ID PRIMER PRSTART BARCODE INSEQ ALIGN OUTSEQ ERROR >> "${Immcantation_dir}pipeline_log/${file}_pRESTO_perf.log" 2>&1

# concatenate back maskprimer.py pass reads and fail (pass = trimmed / failed = without TS, not trimmed= good read ...)
  cat "${Immcantation_dir}02_pRESTO/04_MaskPrimers2/${file}_clean_primers-pass.fastq" "${Immcantation_dir}02_pRESTO/04_MaskPrimers2/${file}_clean_primers-fail.fastq" > "${Immcantation_dir}02_pRESTO/04_MaskPrimers2/${file}_clean_primers-all_concatenate.fastq"
done

# End time and date
end_time=$(date +%s)
end_date=$(date +"%Y-%m-%d %H:%M:%S")
duration=$((end_time - start_time))
hours=$((duration / 3600))
minutes=$(( (duration % 3600) / 60 ))
seconds=$((duration % 60))

# Final report
echo ">>> Pipeline ended at: $end_date"
echo ">>> Total runtime: ${hours}h ${minutes}m ${seconds}s (${duration} seconds)"


# nano > Copy paste the script from the shebang to the final report > pRESTO_loop_all_samples.sh

# As executable file
chmod +x pRESTO_loop_all_samples.sh

# run the loop as nohup so that it do not keyboad interupt when the session disconnect
nohup ./pRESTO_loop_all_samples.sh > pipeline_log/presto_loop_nohup__all_samples_log.txt 2>&1 &

```

### Fastp before alignment

-   Fastp to filter reads lower than 40 bp and get quality of reads before alignment.  

```{bash,eval=F}
# in a conda environment : conda activate fastp

mkdir 02_pRESTO/05_Fastp_ready_to_align

# put the script in a sh executable

#!/bin/bash

# Activate conda environment
source PATH/conda.sh  # Adjust path if needed
conda activate fastp

Immcantation_dir="PATH/"

cd ${Immcantation_dir}

files=$(cat PATH/uniq_id_n104samples.txt)

# Start timing
start_time=$(date +%s)
start_date=$(date +"%Y-%m-%d %H:%M:%S")
echo ">>> Pipeline started at: $start_date"

for file in $files; do

  echo $file

    echo "run fastp"
    fastp --in1 "${Immcantation_dir}02_pRESTO/04_MaskPrimers2/${file}_clean_primers-all_concatenate.fastq" \
    --out1 "${Immcantation_dir}02_pRESTO/05_Fastp_ready_to_align/${file}_ready_to_align.fastq" \
    --json "${Immcantation_dir}02_pRESTO/05_Fastp_ready_to_align/${file}_ready_to_align.json" \
    --html "${Immcantation_dir}02_pRESTO/05_Fastp_ready_to_align/${file}_ready_to_align.html" \
    --dont_eval_duplication --correction \
    --overlap_len_require=30 --overlap_diff_limit=5 --overlap_diff_percent_limit=20 \
    --average_qual=20 \
    --length_required=40 \
    --adapter_sequence=AGATCGGAAGAGCACACGTCTGAACTCCAGTCA \
    --adapter_sequence_r2=AGATCGGAAGAGCGTCGTGTAGGGAAAGAGTGT \
    --thread=16 > "${Immcantation_dir}pipeline_log/${file}_ready_to_align.log" 2>&1

    # Ajoutez votre traitement pour chaque itération de la boucle
    echo "complete"

done

# End time and date
end_time=$(date +%s)
end_date=$(date +"%Y-%m-%d %H:%M:%S")
duration=$((end_time - start_time))
hours=$((duration / 3600))
minutes=$(( (duration % 3600) / 60 ))
seconds=$((duration % 60))

# Final report
echo ">>> Pipeline ended at: $end_date"
echo ">>> Total runtime: ${hours}h ${minutes}m ${seconds}s (${duration} seconds)"

# nano > Copy paste the script from the shebang to the done of the loop > fastp_beforeIgBlast_loop_all_samples.sh

# As executable file
chmod +x fastp_beforeIgBlast_loop_all_samples.sh

# run the loop as nohup so that it do not keyboad interupt when the session disconnect
nohup ./fastp_beforeIgBlast_loop_all_samples.sh > pipeline_log/fastp_beforeIgBlast_loop_nohup__all_samples_log.txt 2>&1 &




####################
# conda deactivate 

```


## IgBlast alignment

Align against Ig and Tr. Tr is used then to filter incorrect alignment

```{bash,eval=F}
# in a conda environment : conda activate pRESTO

Immcantation_dir="PATH/"

cd ${Immcantation_dir}

files=`cat uniq_id_n104samples.txt`

mkdir 03_ChangeO
mkdir 03_ChangeO/01_IgBlast

# put the script in a sh executable

#!/bin/bash

# Activate conda environment
source PATH/conda.sh  # Adjust path if needed
conda activate pRESTO

Immcantation_dir="PATH/"

cd ${Immcantation_dir}

files=$(cat PATH/uniq_id_n104samples.txt)

# Start timing
start_time=$(date +%s)
start_date=$(date +"%Y-%m-%d %H:%M:%S")
echo ">>> Pipeline started at: $start_date"

for file in $files; do

  echo $file

## IgBlast 1.22 alignment - IG

echo "Runing AssignGenes.py igblast 1.22 against IMGT database - For IG"

# Input file WITHOUT any collapse duplicate but on Maskqual.fastq file
## IgBlast alignment
    AssignGenes.py igblast -s "${Immcantation_dir}02_pRESTO/05_Fastp_ready_to_align/${file}_ready_to_align.fastq" \
    -b PATH/igblast \
    --organism human \
    --loci ig \
    --format blast \
    --outname "${file}_ig" \
    --outdir "${Immcantation_dir}03_ChangeO/01_IgBlast/" \
    --nproc 32 > "${Immcantation_dir}pipeline_log/${file}_ChangeO_and_PostProcess_perf.log" 2>&1

## Convert IgBlast result in a database
echo "Runing MakeDb.py igblast to convert IgBlast IG result in a database"
    MakeDb.py igblast -i "${Immcantation_dir}03_ChangeO/01_IgBlast/${file}_ig_igblast.fmt7" \
    -s "${Immcantation_dir}03_ChangeO/01_IgBlast/${file}_ready_to_align.fasta" \
    -r PATH/germlines/imgt/human/vdj PATH/germlines/imgt/human/constant \
    --extended \
    --log "${Immcantation_dir}03_ChangeO/01_IgBlast/${file}_igblast_ig.log" \
    --outdir "${Immcantation_dir}03_ChangeO/01_IgBlast/" >> "${Immcantation_dir}pipeline_log/${file}_ChangeO_and_PostProcess_perf.log" 2>&1


## IgBlast 1.22 alignment - TR

echo "Runing AssignGenes.py igblast 1.22 against IMGT database - For TR"

# Input file WITHOUT any collapse duplicate but on Maskqual.fastq file
## IgBlast alignment
    AssignGenes.py igblast -s "${Immcantation_dir}02_pRESTO/05_Fastp_ready_to_align/${file}_ready_to_align.fastq" \
    -b PATH/igblast \
    --organism human \
    --loci tr \
    --format blast \
    --outname "${file}_tr" \
    --outdir "${Immcantation_dir}03_ChangeO/01_IgBlast/" \
    --nproc 32 >> "${Immcantation_dir}pipeline_log/${file}_ChangeO_and_PostProcess_perf.log" 2>&1

## Convert IgBlast result in a database
echo "Runing MakeDb.py igblast to convert IgBlast TR result in a database"
    MakeDb.py igblast -i "${Immcantation_dir}03_ChangeO/01_IgBlast/${file}_tr_igblast.fmt7" \
    -s "${Immcantation_dir}03_ChangeO/01_IgBlast/${file}_ready_to_align.fasta" \
    -r PATH/germlines/imgt/human/vdj PATH/germlines/imgt/human/constant \
    --extended \
    --log "${Immcantation_dir}03_ChangeO/01_IgBlast/${file}_igblast_tr.log" \
    --outdir "${Immcantation_dir}03_ChangeO/01_IgBlast/" >> "${Immcantation_dir}pipeline_log/${file}_ChangeO_and_PostProcess_perf.log" 2>&1

done

# End time and date
end_time=$(date +%s)
end_date=$(date +"%Y-%m-%d %H:%M:%S")
duration=$((end_time - start_time))
hours=$((duration / 3600))
minutes=$(( (duration % 3600) / 60 ))
seconds=$((duration % 60))

# Final report
echo ">>> Pipeline ended at: $end_date"
echo ">>> Total runtime: ${hours}h ${minutes}m ${seconds}s (${duration} seconds)"

# nano > Copy paste the script from the shebang to the done of the loop > IgBlast_loop_all_samples.sh

# As executable file
chmod +x IgBlast_loop_all_samples.sh

# run the loop as nohup so that it do not keyboad interupt when the session disconnect
nohup ./IgBlast_loop_all_samples.sh > pipeline_log/IgBlast_loop_nohup__all_samples_log.txt 2>&1 &

```

## Post alignment processing

All reads were aligned against IMGT referece with IgBlast 2 times, one to create an Ig DB, one to create a TR DB.

Verify that reads are either assigned to IG or TR but not both ... Keep only Ig because Immunoverse experiment was performed only with Ig primers.

Post process IgBlast alignment and create a log file.

### Quality filters

```{r,eval=FALSE}
##Un comment it to save

#dir.create(paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter"))

```

```{r,eval=FALSE}
# Create a metadata data.frame to store the process pipeline results and create integrated plots

# The ordered QC filters
Var_to_collect <- c("IgBlast_input_reads",
                    "Not_aligned_reads",
                    "Ig_aligned_reverse_incorrect",
                    "Ambiguous_Ig_Tr_aligned",
                    "Too_many_N_in_VDJ",
                    "Short_VDJ_alignment",
                    "Unconsistent_VDJ_genes",
                    "Unproductive_VDJ",
                    "Junction_not_multiple_of_3",
                    "Pass_Alignment_QC")

Metadata_QC <- data.frame(matrix(NA,nrow = length(FileID),ncol = length(Var_to_collect)))
rownames(Metadata_QC)<-FileID
colnames(Metadata_QC)<-Var_to_collect

# Open a file connection for writing a log file ##un comment if needed
sink(paste0(ImmcantationFolder,"pipeline_log/","R_Immunoverse_IgBlast_Alignment_QualityFilter.log"), append = FALSE, type = c("output"))

  t1<-Sys.time()
  cat(paste("Start Alignment QC pipeline:",as.character(t1)))
  #add space in the log
  cat("\n")
  cat("\n")

for(file in FileID){

  cat(file,"\n")
  #create directory to store results per file
  if(!file.exists(paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file))) dir.create(paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file)) 

  # Do not need to adapt from here - path of R1 after clustering and renaming
  file_path <- paste0(ImmcantationFolder,"03_ChangeO/01_IgBlast/",file)
  
  # Load airr from IgBlast alignment - Not collapsed file ...
  options(readr.show_progress = FALSE)
  DB_IG <- read_rearrangement(paste0(file_path,"_ig_igblast_db-pass.tsv"))
  DB_TR <- read_rearrangement(paste0(file_path,"_tr_igblast_db-pass.tsv"))
  
  # Compute total number of reads aligned to Ig and to TR from input
  ## Retrieve the number of seq used in input of MiXCR for the pipeline log
  ## USe the bash command wc -l within R
  
  Input <- paste0(ImmcantationFolder,"03_ChangeO/01_IgBlast/",file,"_ready_to_align.fasta")
  Input <- system(paste("wc -l", shQuote(Input)), intern = TRUE)
  Input <- as.numeric(unlist(base::strsplit(Input,split = " "))[[1]] )
  Input <- Input/2 #the command above retrieve the number of line in fasta. Divide by 2 to get the number of seq
  cat(paste0("Reads in input for IgBlast (Insert only) : ",Input))
  cat("\n")
  Metadata_QC[file,"IgBlast_input_reads"]<-Input  # fill Metadata

  ## Reads aligned to Ig with IgBlast
  Num <- nrow(DB_IG)
  Prop <- round((Num/Input)*100,2)
  cat(paste0("Reads aligned to Ig with IgBlast : ",Num," (",Prop,"%)"))
  cat("\n")
  Metadata_QC[file,"Not_aligned_reads"]<-(Input-Num)  # fill Metadata
  
  ## Reads aligned to TR with IgBlast
  Num <- nrow(DB_TR)
  Prop <- round((Num/Input)*100,2)
  cat(paste0("Reads aligned to TR with IgBlast : ",Num," (",Prop,"%)"))
  cat("\n")
  
  # Reads should be aligned in forward orientation only
  ## orientation in rev comp suggest an incorrect pre-processing of reads and thus an incorrect UMI attribution
  ### Number of seq with incorrect Ig alignment orientation
  Num <- sum((DB_IG$rev_comp))
  PropIg <- round((Num/Input)*100,2)
  cat(paste0("Reads aligned to Ig in Incorrect orientation (reverse) : ",Num," (",PropIg," %Ig)"))
  cat("\n")

  Metadata_QC[file,"Ig_aligned_reverse_incorrect"]<-Num  # fill Metadata
  
  ### Number of seq with incorrect TR alignment orientation
  #Num <- sum((DB_TR$rev_comp))
  #PropTR <- round((Num/nrow(DB_TR))*100,2)
  #cat(paste0("Reads aligned to TR in Incorrect orientation (reverse) : ",Num," (",PropTR," %TR)"))
  
  
  # Filter only reads aligned in right orientation (forward : inverse or rev_comp) and only for Ig or Tr
  DB_IG <- DB_IG[!DB_IG$rev_comp,]
  DB_TR <- DB_TR[!DB_TR$rev_comp,]
  
  # Compute number of reads aligned to both Ig and TR
  if(sum(DB_IG$sequence_id %in% DB_TR$sequence_id)>0){
    DB_IG_Intersect <- DB_IG[DB_IG$sequence_id %in% DB_TR$sequence_id,]
    DB_IG_Intersect <- data.frame(DB_IG_Intersect)
    rownames(DB_IG_Intersect)<-DB_IG_Intersect$sequence_id
    DB_TR_Intersect <- DB_TR[DB_TR$sequence_id %in% DB_IG$sequence_id,]
    DB_TR_Intersect <- data.frame(DB_TR_Intersect)
    rownames(DB_TR_Intersect)<-DB_TR_Intersect$sequence_id
    DB_TR_Intersect <- DB_TR_Intersect[rownames(DB_IG_Intersect),] # check that the order of row for both subset are the same
    
    ## check v score of Ig and Tr alignment (top score), same for j
    TopVscore <- ifelse(DB_IG_Intersect$v_score < DB_TR_Intersect$v_score,
           yes = "V-TR",
           no = ifelse(DB_IG_Intersect$v_score == DB_TR_Intersect$v_score,
                       yes ="V-ambiguous" ,
                       no = "V-IG"))
    TopJscore <- ifelse(DB_IG_Intersect$j_score < DB_TR_Intersect$j_score,
           yes = "J-TR",
           no = ifelse(DB_IG_Intersect$j_score == DB_TR_Intersect$j_score,
                       yes ="J-ambiguous" ,
                       no = "J-IG"))
    ## Same for v/j_support. Lower value means better probability of correct match
    ## check v score of Ig and Tr alignment (top score), same for j
    LowVsupport <- ifelse(DB_IG_Intersect$v_support > DB_TR_Intersect$v_support,
           yes = "Vsupport-TR",
           no = ifelse(DB_IG_Intersect$v_support == DB_TR_Intersect$v_support,
                       yes ="Vsupport-ambiguous" ,
                       no = "Vsupport-IG"))
    LowJsupport <- ifelse(DB_IG_Intersect$j_support > DB_TR_Intersect$j_support,
           yes = "Jsupport-TR",
           no = ifelse(DB_IG_Intersect$j_support == DB_TR_Intersect$j_support,
                       yes ="Jsupport-ambiguous" ,
                       no = "Jsupport-IG"))
    
    
    ## Check Ig and Tr alignement length
    ## SeqLen could be commented to keep memory/time
    IGseq_len <- stri_count(DB_IG_Intersect[["sequence_alignment"]],regex="[^-.N]")
    TRseq_len <- stri_count(DB_TR_Intersect[["sequence_alignment"]],regex="[^-.N]")
    TopSeqLen <- ifelse(IGseq_len < TRseq_len,
           yes = "SeqLen-TR",
           no = ifelse(IGseq_len == TRseq_len,
                       yes ="SeqLen-ambiguous" ,
                       no = "SeqLen-IG"))
    
    
    
    Ig_Or_Tr_df <- data.frame(TopVscore=TopVscore,
                              TopJscore=TopJscore,
                              TopSeqLen=TopSeqLen,
                              LowVsupport=LowVsupport,
                              LowJsupport=LowJsupport,
                              IgProd=ifelse(DB_IG_Intersect$productive,yes ="IG-Prod" ,no = "IG-NotProd"),
                              IgInFrame=ifelse(DB_IG_Intersect$vj_in_frame,yes = "IG-InFrame",no = "IG-NotInFrame"),
                              TrProd=ifelse(DB_TR_Intersect$productive,yes ="TR-Prod",no = "TR-NotProd"),
                              TrInFrame=ifelse(DB_TR_Intersect$vj_in_frame,yes ="TR-InFrame",no = "TR-NotInFrame"))
    
    
    Ig_Or_Tr_df$Collapse_only4Top <-paste(Ig_Or_Tr_df$TopVscore,Ig_Or_Tr_df$LowVsupport,Ig_Or_Tr_df$TopJscore,Ig_Or_Tr_df$LowJsupport,sep  = "_")
    
    # cat nb of correct Ambiguous / correct IG / correct TR
    ## Total read with both alignment
    Num <- nrow(Ig_Or_Tr_df)
    PropIg <- round((Num/nrow(DB_IG))*100,2)
    PropTR <- round((Num/nrow(DB_TR))*100,2)
    cat(paste0("Reads aligned to TR as well as Ig with IgBlast : ",Num," (",PropIg," %Ig, ",PropTR, " %TR)"))
    cat("\n")
    
    ## Ambiguous vs saved read
    AmbiguousTable <- table(Ig_Or_Tr_df$Collapse_only4Top)
    NumIg <- AmbiguousTable[["V-IG_Vsupport-IG_J-IG_Jsupport-IG"]]
    ## There should be no true TR so this is commented (from SpatialVDJ approach)
    #NumTr <- AmbiguousTable[["V-TR_Vsupport-TR_J-TR_Jsupport-TR"]]
    NumAmbig <- sum(AmbiguousTable)-NumIg
    PropIg <- round((NumIg/sum(AmbiguousTable))*100,2)
    #PropTR <- round((NumTr/sum(AmbiguousTable))*100,2)
    PropAmbig <- round((NumAmbig/sum(AmbiguousTable))*100,2)

    cat(paste0("Reads aligned to TR as well as Ig with IgBlast - TRUE Ig : ",NumIg," (",PropIg,"%)"))
    cat("\n")
    #cat(paste0("Reads aligned to TR as well as Ig with IgBlast - TRUE Tr : ",NumTr," (",PropTR,"%)"))
    cat(paste0("Reads aligned to TR as well as Ig with IgBlast - Ambiguous Ig-Tr : ",NumAmbig," (",PropAmbig,"%)"))
    cat("\n")
    
     Metadata_QC[file,"Ambiguous_Ig_Tr_aligned"]<-NumAmbig  # fill Metadata

    #DB_intersecet_retained
    id_seq_true_ig <- DB_IG_Intersect$sequence_id[Ig_Or_Tr_df$Collapse_only4Top=="V-IG_Vsupport-IG_J-IG_Jsupport-IG"]
    #id_seq_true_tr <- DB_TR_Intersect$sequence_id[Ig_Or_Tr_df$Collapse_only4Top=="V-TR_Vsupport-TR_J-TR_Jsupport-TR"]

    DB_intersect_retained <- DB_IG[DB_IG$sequence_id %in% id_seq_true_ig,]
    
    # Concatenate DB_IG and DB_intersect_retained
    ## For reads both Ig and Tr aligned, select top V score aligned to conclude as Ig or Tr. Remove ambiguous 
    airr <- rbind(DB_IG[DB_IG$sequence_id %in% base::setdiff(DB_IG$sequence_id, DB_TR$sequence_id),], #Ig specific (not aligned to TR)
                  # DB_TR[DB_TR$sequence_id %in% base::setdiff(DB_TR$sequence_id,DB_IG$sequence_id),], #Tr specific (not aligned to Ig)
                  DB_intersect_retained)
    
  }else{
    cat(paste0("Reads aligned to TR as well as Ig with IgBlast : ","0"," (","0"," %Ig, ","0", " %TR)"))
    cat("\n")
    cat(paste0("Reads aligned to TR as well as Ig with IgBlast - TRUE Ig : ","NA"," (","NA","%)"))
    cat("\n")
    #cat(paste0("Reads aligned to TR as well as Ig with IgBlast - TRUE Tr : ","NA"," (","NA","%)"))
    cat(paste0("Reads aligned to TR as well as Ig with IgBlast - Ambiguous Ig-Tr : ","NA"," (","NA","%)"))
    cat("\n")
    
         Metadata_QC[file,"Ambiguous_Ig_Tr_aligned"]<-0  # fill Metadata

  }
  
  
  Num <- nrow(airr)
  cat(paste0("Sequences kept after filtering incorrect and ambiguous alignment : ",Num))
  cat("\n")
  
  ######################################################################
  # **Identify Reads with an Acceptable Number of Ambiguous Nucleotides - Max 10% N**
  ## Identify Reads with an Acceptable Number of Ambiguous Nucleotides
  ## Max 10% N
  airr$num_n_inVDJ <- stri_count(airr[["sequence_alignment"]],fixed="N") 
  
  ## Plot the Num N nucleotide distribution
  Plot1 <- ggplot(airr, aes(x = num_n_inVDJ)) +
    geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
    labs(title = paste("Number of N per VDJ"), x = "Number of N in aligned sequence", y = "Frequency")+
    theme_bw()
  
  ## Same in ratio
  airr$seqVDJ_aligned_length <- stri_count(airr[["sequence_alignment"]],regex="[^-.]") # compute seq_len with N
  airr$ratio_n_inVDJ <- airr$num_n_inVDJ/airr$seqVDJ_aligned_length
  
  ## Plot the ratio N nucleotide distribution
  Plot2 <- ggplot(airr, aes(x = ratio_n_inVDJ)) +
    geom_histogram(binwidth = 0.01, fill = "blue", color = "black", alpha = 0.7) +
    labs(title = paste("Ratio of N per VDJ"), x = "Ratio of N in aligned sequence", y = "Frequency") +
    geom_vline(xintercept = 0.1, linetype = "dotted", color = "blue", linewidth = 1) +
    annotate("text", x = 0.1, y = Inf, label = "Quality filter\nLow N", color = "blue", vjust = +2.5,hjust = -0.1)+
    scale_x_continuous(limits = c(-0.01,1.01))+
    theme_bw()
  
  ##store results
  ggsave(filename = paste0(res_folder,file,"_AlignedSeq_Num_and_Ratio_N.pdf"),plot = (Plot1|Plot2),width = 8,height = 4)

  ##Filter and save table results
  low_n <- airr$ratio_n_inVDJ <= 0.10
  # no_n <- ratio_n==0
  
  Num <- sum(!low_n)
  Prop <- round((Num/Input)*100,2)
  cat(paste0("Filter vdj with more than 10% N : ",Num," (",Prop,"%)"))
  cat("\n")

  Metadata_QC[file,"Too_many_N_in_VDJ"]<-Num  # fill Metadata
  
  airr <- airr[low_n,] ## Keep only seq with ratio N <= 0.1 >> low_n TRUE
  
  ######################################################################
  # **Identify Sequences with a Min length of 200 nt**
  ### Identify Short Sequences
  ### Compute sequence alignment length excluding dots
  ## Min length 200 nt
  airr$read_length <- stri_count(airr[["sequence"]],regex="[^-.]")
  summary_length <- summary(airr$read_length)
  
  # cat("Long read length in input to IgBlast (only informative positions)")
  # cat(summary_length) 
   
  Plot1 <- ggplot(airr, aes(x = read_length)) +
    geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
    scale_x_continuous(limits = c(0,610))+
    labs(title = paste("Read length in input to IgBlast"), x = "Input Sequence Length", y = "Frequency") +
    # scale_y_log10()+
     geom_vline(xintercept = summary_length["Mean"], linetype = "dashed", color = "red", linewidth = 1) +
    annotate("label", x = summary_length["Mean"], y = Inf, label = paste("Average Length:", round(summary_length["Mean"], 2)), color = "red", vjust = +1,hjust = 0.5)+
    theme_bw()

  ## Same for aligned sequence without N (only informative bases)
  airr$seqVDJ_aligned_length_noN <- stri_count(airr[["sequence_alignment"]],regex="[^-.N]")
  summary_length <- summary(airr$seqVDJ_aligned_length_noN)
  
  ## Plot the sequence length distribution
  Plot2 <- ggplot(airr, aes(x = seqVDJ_aligned_length_noN)) +
    geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
    scale_x_continuous(limits = c(0,610))+
    labs(title = paste("Aligned sequence (VDJ) length"),subtitle = "Gap and N excluded - Only informative", x = "Aligned Sequence Length", y = "Frequency") +
    # scale_y_log10()+
    geom_vline(xintercept = 200, linetype = "dotted", color = "blue", linewidth = 1) +
    annotate("label", x = 200, y = Inf, label = "Quality filter\nlength>200b", color = "blue", vjust = +2.5,hjust = +1)+
     geom_vline(xintercept = summary_length["Mean"], linetype = "dashed", color = "red", linewidth = 1) +
    annotate("label", x = summary_length["Mean"], y = Inf, label = paste("Average Length:", round(summary_length["Mean"], 2)), color = "red", vjust = +1,hjust = 0.5)+
    theme_bw()
  
  ##store results
  ggsave(filename = paste0(res_folder,file,"_InputRead_and_AlignedSeqVDJnoN_length.pdf"),plot = Plot1/Plot2,width = 5,height = 7)
  
  #####  #####  Store an intermediate object containing only read and vdj no N length to create an integrated plot after the loop
  Store_length<- airr[,c("read_length","seqVDJ_aligned_length_noN")]
  Store_length$sampleid <- file
  saveRDS(object = Store_length,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_not_filtered_read_and_vdj_seqlen.rds"))
    #####   ##### 
  
  ##create a filter
  long_seq <- airr$seqVDJ_aligned_length >= 200
  
  ##store results
  Num <- sum(!long_seq)
  Prop <- round((Num/Input)*100,2)
  cat(paste0("Filter vdj alignment < 200 bases (no N or gap) : ",Num," (",Prop,"%)"))
  cat("\n")

  Metadata_QC[file,"Short_VDJ_alignment"]<-Num  # fill Metadata

  airr <- airr[long_seq,] ## Keep only seq with vdj informative len >=200 >> long_seq TRUE

  ######################################################################
  # **Identify disagreements between the VDJC IgBLAST alignment**
  ## Identify sequences with a disagreements between the V/J gene alignment and and the C gene IgBLAST alignment
  ### Identify sequences with no C_call identify by IgBlast and remove them or tag chain and isotype as chain_NA
  ## Keep reads with coherent gene, primer and isotype calls 

  ## Compute disagreement
  V_disagree_locus <-ifelse(getLocus(airr[["v_call"]]) == airr[["locus"]],yes = "same_locus",no = "V_different_from_locus")
  
  J_disagree_locus <-ifelse(getLocus(airr[["j_call"]]) == airr[["locus"]],yes = "same_locus",no = "J_different_from_locus")
  
  ##replace C_Call NA as paste(locus,NA)
  airr$isotype <- gsub("\\*","",substr(airr[["c_call"]],1,5))
  airr$isotype <- ifelse(test = is.na((airr[["isotype"]])),
                                      yes = paste(airr[["locus"]],"NA",sep = "_"),
                                      no = airr[["isotype"]])
  
  Isotype_disagree_locus <- ifelse(getLocus(airr[["isotype"]]) == airr[["locus"]],yes = "same_locus",no = "C_different_from_locus")
  
  same_locus <- (V_disagree_locus=="same_locus")  & (J_disagree_locus=="same_locus") & (Isotype_disagree_locus=="same_locus")
  
  ##store results
  Num <- sum(!same_locus)
  Prop <- round((Num/Input)*100,2)
  cat(paste0("Filter inconsistent VDJ (e.g. VH and JK) : ",Num," (",Prop,"%)"))
  cat("\n")
  
  Metadata_QC[file,"Unconsistent_VDJ_genes"]<-Num  # fill Metadata

  airr <- airr[same_locus,] ## Keep only seq with consistent vdj >> same_locus TRUE

  ######################################################################
  # **Only productive sequences are kept**
  ##If not already filtered, only productive sequences are kept
  prod <- airr[["productive"]]

  ##store results
  Num <- sum(!prod)
  Prop <- round((Num/Input)*100,2)
  cat(paste0("Filter unproductive vdj : ",Num," (",Prop,"%)"))
  cat("\n")
  
  Metadata_QC[file,"Unproductive_VDJ"]<-Num  # fill Metadata

  airr <- airr[prod,] ## Keep only seq with productive vdj >> prod TRUE
  
  ######################################################################
  # **Identify sequences with junction length multiple of three**
  ### Identify sequences with junction length multiple of three
  ## %% modulo operator
  
  m3 <- airr[["junction_length"]] %% 3 == 0
  
  Num <- sum(!m3)
  Prop <- round((Num/Input)*100,2)
  cat(paste0("Filter junction with length not modulo 3 : ",Num," (",Prop,"%)"))
  cat("\n")
  
  Metadata_QC[file,"Junction_not_multiple_of_3"]<-Num  # fill Metadata
  
  airr <- airr[m3,] ## Keep only seq with junction modulo 3 >> m3 TRUE
  
  
  ######################################################################
  ### Count and save repertoire
  Num <- nrow(airr)
  Prop <- round((Num/Input)*100,2)
  cat(paste0("Sequences kept after all QC filter : ",Num," (",Prop,"%)"))
  cat("\n")
  
  Metadata_QC[file,"Pass_Alignment_QC"]<-Num  # fill Metadata
  
  # plot chain and proportion of reads (not deduplicated)

  TableChain <- table(airr$locus)
  Total <- sum(TableChain)

  cat(paste0("Total READS identifying clone after quality filter (no deduplication): ",Total))
  cat("\n")

  Num<-ifelse(test = "IGH" %in% names(TableChain),yes = TableChain["IGH"],no = 0)
  Prop <- round((Num/Total)*100,2)
  cat(paste0("Total READS after all quality filter - IGH : ",Num," (",Prop,"%)"))
  cat("\n")
  
  Num<-ifelse(test = "IGK" %in% names(TableChain),yes = TableChain["IGK"],no = 0)
  Prop <- round((Num/Total)*100,2)
  cat(paste0("Total READS after all quality filter - IGK : ",Num," (",Prop,"%)"))
  cat("\n")
  
  Num<-ifelse(test = "IGL" %in% names(TableChain),yes = TableChain["IGL"],no = 0)
  Prop <- round((Num/Total)*100,2)
  cat(paste0("Total READS after all quality filter - IGL : ",Num," (",Prop,"%)"))
  cat("\n")

  
  pal <- RColorBrewer::brewer.pal(n = 11,name = "Paired")
  pal <- c(IGL=pal[[1]],
           IGK=pal[[2]],
           IGH=pal[[6]])
  
  
  ## Plot chain percentage

  TablePlot <- c("IGH"=0,"IGK"=0,"IGL"=0)
  TablePlot[names(TableChain)] <-TableChain 

  
  pdf(paste0(res_folder,file,"_READS_after_all_quality_filter_Chain_proportion.pdf"),width = 5, height = 5) # Save a pie plot in pdf
    pie(TablePlot, 
        labels = paste0(names(TablePlot)," n=",TablePlot), 
        main = "Chain proportion in READS after all quality filter", 
        col = pal[names(TablePlot)])
  invisible(dev.off()) # close the pdf device without printing the message in the sink  
    
  ##Un comment it to save
  # 
  write_rearrangement(airr,file=paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter.tsv"))
  
  cat("\n")

  #keep some memory
  ## remove all object but c("FileID","ImmcantationFolder","RawReadFolder","res_folder")
  rm(list=setdiff(ls(),c("FileID","ImmcantationFolder","res_folder","t1","Metadata_QC")))
  gc()
  
}

  t2<-Sys.time()
  cat(paste("Pipeline complete for all samples:",as.character(t2)))
  cat("\n")
  t2-t1
  cat("\n")
sink()

# Save the Metadata in RDS to generate plot later
## run on all sample
saveRDS(object = Metadata_QC,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","Metadata_Step1_alignment_QC_All_samples.rds"))

```

### Chimeric reads

#### Create Germlines

```{bash,eval=F}
# in a environment : conda activate pRESTO

# put the script in a sh executable

#!/bin/bash

# Activate conda environment
source PATH/conda.sh  # Adjust path if needed
conda activate pRESTO

Immcantation_dir="PATH/"

cd ${Immcantation_dir}

files=$(cat PATH/uniq_id_n104samples.txt)

# Start timing
start_time=$(date +%s)
start_date=$(date +"%Y-%m-%d %H:%M:%S")
echo ">>> Pipeline started at: $start_date"

for file in $files; do

  echo $file

## Create Germlines : Input file WITHOUT any collapse duplicate 

echo "Runing CreateGermlines.py"
CreateGermlines.py -d "${Immcantation_dir}03_ChangeO/02_PostIgBlastFilter/${file}/${file}_igblast-db_quality_filter-allFilter.tsv" -r PATH/germlines/imgt/human/vdj -g dmask --format airr --outdir "${Immcantation_dir}03_ChangeO/02_PostIgBlastFilter/${file}/" >> "${Immcantation_dir}pipeline_log/${file}_ChangeO_and_PostProcess_perf.log" 2>&1

done


# End time and date
end_time=$(date +%s)
end_date=$(date +"%Y-%m-%d %H:%M:%S")
duration=$((end_time - start_time))
hours=$((duration / 3600))
minutes=$(( (duration % 3600) / 60 ))
seconds=$((duration % 60))

# Final report
echo ">>> Pipeline ended at: $end_date"
echo ">>> Total runtime: ${hours}h ${minutes}m ${seconds}s (${duration} seconds)"

# nano > Copy paste the script from the shebang to the done of the loop > CreateGermline_reads_loop_all_samples.sh

# As executable file
chmod +x CreateGermline_reads_loop_all_samples.sh

# run the loop as nohup so that it do not keyboad interupt when the session disconnect
nohup ./CreateGermline_reads_loop_all_samples.sh > pipeline_log/CreateGermline_reads_loop_nohup__all_samples_log.txt 2>&1 &

```

#### Chimeric read filtering

```{r,eval=F}
# Load a modified version of slideWindowDb to close the cluster at the end when nproc=12 (numeric) ...
## if (stop_cluster & !is.numeric(nproc)) =>>> replaced by : if (stop_cluster). if nproc=12; then the statement was evaluated as False (!is.numeric(nproc)) and the cluster was not stropped ...
#clear the session before running the following script. With 10 cores, it takes aroun 20-30 Go memory per iterations

slideWindowDb_modif <- function(db, sequenceColumn="sequence_alignment",
                          germlineColumn="germline_alignment_d_mask",
                          mutThresh=6, windowSize=10, nproc=1){
    # Hack for visibility of foreach index variables
    i <- NULL
    
    # Check input
    check <- checkColumns(db, c(sequenceColumn, germlineColumn))
    if (check != TRUE) { stop(check) }

    db <- db[,c(sequenceColumn, germlineColumn)]
    # If the user has previously set the cluster and does not wish to reset it
    if(!is.numeric(nproc)){
        stop_cluster <- FALSE
        cluster <- nproc
        nproc <- 0
    } else {
        stop_cluster <- TRUE
    }

    # Ensure that the nproc does not exceed the number of cores/CPUs available
    nproc <- min(nproc, cpuCount())

    # If user wants to parallelize this function and specifies nproc > 1, then
    # initialize and register slave R processes/clusters &
    # export all necessary environment variables, functions and packages.
    if (nproc == 1) {
        # If needed to run on a single core/cpu then, register DoSEQ
        # (needed for 'foreach' in non-parallel mode)
        registerDoSEQ()
    } else {
        cluster_type <- "FORK"
        if (.Platform$OS.type == "windows") {
            cluster_type <- "PSOCK"
        }
        if (nproc != 0) {
            #cluster <- makeCluster(nproc, type="SOCK")
            cluster <- parallel::makeCluster(nproc, type= cluster_type)
        }
        parallel::clusterExport(cluster,
                                    list('db', 'sequenceColumn', 'germlineColumn',
                                         'mutThresh', 'windowSize','slideWindowSeq'),
                                    envir=environment() )
        registerDoParallel(cluster,cores=nproc)
    }

    filter <- unlist(foreach(i=1:nrow(db),
                   .verbose=FALSE, .errorhandling='stop') %dopar% {
                       slideWindowSeq(inputSeq = db[i, sequenceColumn],
                                      germlineSeq = db[i, germlineColumn],
                                      mutThresh = mutThresh,
                                      windowSize = windowSize)
                   })
    if (stop_cluster) {
        parallel::stopCluster(cluster)
    }
    filter

}


#Reload metadata
## reload
Metadata_QC <- readRDS(paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","Metadata_Step1_alignment_QC_All_samples.rds"))
#  

library(doParallel)
```


```{r,eval=F}
# Open a file connection for writing a log file ##un comment if needed
sink(paste0(ImmcantationFolder,"pipeline_log/","R_Immunoverse_IgBlast_Alignment_ChimericReadFilter.log"), append = FALSE, type = c("output"))

  t1<-Sys.time()
  cat(paste("Start chimeric filtering pipeline:",as.character(t1)))
  #add space in the log
  cat("\n")
  cat("\n")


for(file in FileID){
  
  cat(file)
  cat("\n")
  
  # Do not need to adapt from here - path of Igblast quality filtered database
  file_path <- paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass.tsv")
  
  # Load airr 
  options(readr.show_progress = FALSE)
  airr <- read_rearrangement(file_path)
  
  cat(paste("Start slideWindowDb_modif for file :",file))
  cat("\n")
  t11 <- Sys.time()
  
  is_chimeric <- slideWindowDb_modif(airr,sequenceColumn = "sequence_alignment", germlineColumn = "germline_alignment_d_mask", mutThresh=6,windowSize=10,nproc=30)
  
  t22 <- Sys.time()
  t22-t11
  
  #filter out chimeric reads
  Total <- nrow(airr)
  airr <- airr[!is_chimeric,]
  
  Num<-nrow(airr)
  Prop <- round((Num/Total)*100,2)
  cat(paste0("Total READS after all Chimeric reads filtering : ",Num," (",Prop,"%)"))
  cat("\n")
   
  ##### Fill metadata
  ## first compute the numnber of reads which fail create germline.
  Metadata_QC[file,"Fail_CreateGermline_reads"]<-Metadata_QC[file,"Pass_Alignment_QC"]-Total  # fill Metadata
  ## then the number of reads which fail chimeric filter
  Metadata_QC[file,"Ig_chimeric_reads"]<-sum(is_chimeric)  # fill Metadata

  #####
  
  ##Un comment it to save
  
  write_rearrangement(airr,file=paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter.tsv"))

  #keep some memory
  ## remove all object but c("FileID","ImmcantationFolder","RawReadFolder","res_folder")
  rm(list=setdiff(ls(),c("FileID","ImmcantationFolder","res_folder","slideWindowDb_modif","Metadata_QC","t1")))
  gc()
  
  cat("\n")

}
  t2<-Sys.time()
  cat(paste("Pipeline complete for all samples:",as.character(t2)))
  cat("\n")
  t2-t1
  cat("\n")
sink()

# Save the Metadata in RDS to generate plot later
# saveRDS(object = Metadata_QC,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","Metadata_Step2_ChimericFilter_All_samples.rds"))

# compute chimeric_filter_pass (forgot to add it in the pipeline before)
Metadata_QC$Pass_Chimeric_QC<-NA

for(file in FileID){
  airr <-read_rearrangement(paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter.tsv"))
  Metadata_QC[file,"Pass_Chimeric_QC"]<-nrow(airr)
}  

# Save the Metadata in RDS to generate plot later
saveRDS(object = Metadata_QC,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","Metadata_Step2_ChimericFilter_All_samples.rds"))

```


### Collapse duplicates

```{r,eval=FALSE}
#Reload metadata
## reload from the step 2 (chimeric filter) for all 105 samples
Metadata_QC <- readRDS( paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","Metadata_Step2_ChimericFilter_All_105samples.rds"))
```


```{r,eval=F}

# Open a file connection for writing a log file ##un comment if needed
sink(paste0(ImmcantationFolder,"pipeline_log/","R_Immunoverse_IgBlast_Alignment_CollapseDuplicates.log"), append = FALSE, type = c("output"))

  t1<-Sys.time()
  cat(paste(">>> Pipeline started :",t1))
  cat("\n")
  cat("Collapse duplicates group :",paste(c("v_gene","j_gene","cdr3","isotype","productive"),collapse = ","))
  cat("\n")
  cat("\n")
  
for(file in FileID){
  
  cat(file)
  cat("\n")
  
  # Do not need to adapt from here - path of Igblast quality filtered database
  file_path <- paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter.tsv")
  
  # Load airr 
  options(readr.show_progress = FALSE)
  airr <- read_rearrangement(file_path)
  
  Input <- nrow(airr)
  cat(paste0("Total reads in input (before collapse duplicates) : ",Input))
  cat("\n")
  
  # Start Collapse Duplicates
  
  cat(paste("Start collapseDuplicates for file :",file))
  cat("\n")
  
  t11 <- Sys.time()

    ## Collapse group : same v_gene, j_gene, CDR3, isotype, productive 
  ### WITHOUT UMI to collapse group because I want to define clonotype based on identical CDR3 on all the repertoire (quite similar to MiXCR ...). 
  collapse_groups <- c("v_gene","j_gene","cdr3","isotype","productive")
  
  airr <- airr %>% 
      mutate(v_gene=getGene(v_call), j_gene=getGene(j_call)) %>% 
      group_by(.dots=collapse_groups) 
  
  CollapseGroup=group_indices(airr) # retrieve id of all collapse group
  
  airr <- airr %>%
    ungroup %>%
    mutate(CollapseGroupId=paste0("GroupId",CollapseGroup)) ## Add a groupID column
  
  
  ## Collapse duplicate with dry=TRUE to tag collapsing group and notably add a collapse_id column but do not deleted the sequences
  pbapply::pboptions(type = "none")

  airr <- airr %>%
  group_by(across(all_of(collapse_groups))) %>%
  group_modify(~ {
    collapsed_data <- collapseDuplicates(.x,  id = "sequence_id", seq = "sequence_alignment", text_fields = NULL, num_fields = NULL, seq_fields = NULL, add_count = FALSE, ignore = c("N", "-", ".", "?"), sep = ",", dry = TRUE, verbose = FALSE)
    collapsed_data$collapse_id <- paste0("CollapseId", as.character(collapsed_data$collapse_id))# Convert collapse_id to character type
    collapsed_data
  }) %>%
  ungroup()
  
  # Create a clonotype_id column by concatenating GroupId and CollapseId
  airr$clonotype_id <- paste(airr$CollapseGroupId,airr$collapse_id,sep = "_")

  Num <- sum(airr$collapse_class != "ambiguous")
  Prop <- round((Num/nrow(airr))*100,2)
  cat(paste0("Total reads with not ambiguous clonotype defined : ",Num," (",Prop,"%)"))
  cat("\n")
  
  ##### Fill metadata
  Metadata_QC[file,"Ambiguous_clonotype_reads"]<-sum(airr$collapse_class == "ambiguous")  # fill Metadata
  
  ##### Correct PCR error leading to false diversity :
  ##### Same UMI in different clonotypes of same group (same V, J, cdr3, Isotype) suggest of PCR error
  airr <- airr[airr$collapse_class != "ambiguous",]

  # Count Nb of unique umi per clonotype_id 
  ## Because of low diversity : the same UMI can be found in different clonotypes.
  ## BUT : 2 Different clonotypes of the same group  (same V, J, cdr3, Isotype) suggest PCR error
  Clone_PerUMI <- tapply(airr$clonotype_id,airr$barcode,table)
  
  Read_PerClone <- table(airr$clonotype_id)
  
  Group_PerUMI <- lapply(Clone_PerUMI,function(cln){
    GroupIdDuplicated <- gsub("_.*","",names(cln))
    tapply(names(cln),GroupIdDuplicated,function(cln)cln,simplify = F)
  })
  
  ## search the ID of the cln for which the count is the highest and thus should be kept, the reads in other cln should be filtered

  library(parallel)
  clust <- makeCluster(32)
  clusterExport(clust, "Read_PerClone")
  clusterExport(clust, "Group_PerUMI")
  
  airr$PCRbias <- parApply(cl = clust,airr,1,function(read){
    cln <- read[["clonotype_id"]]
    umi <- read[["barcode"]]
    group <- read[["CollapseGroupId"]]
    
    ListClonoToTest <- Group_PerUMI[[umi]][[group]]
    
    L<-length(ListClonoToTest)
      if(L==1){
        res <- "To_keep"
      }else{
        #The clonotype with the highest number of read is probably the true one ...
        Count <- Read_PerClone[ListClonoToTest]
        Max <- max(Count)
        if(sum(Count==Max)>1){
          res <-"MultipleMax"
        }else{
          res <- ifelse(Read_PerClone[[cln]] == Max,yes = "To_keep",no = "To_filter")
        }
      }
      res
  })
  

  stopCluster(clust)
  
  
  Num <- sum(airr$PCRbias == "To_keep")
  Prop <- round((Num/nrow(airr))*100,2)
  cat(paste0("Total reads after filter PCR bias : ",Num," (",Prop,"%)"))
  cat("\n")
  
  ##### Fill metadata
  Metadata_QC[file,"PCR_bias_filtered_reads"]<-sum(airr$PCRbias != "To_keep")  # fill Metadata
  
  airr <- airr[airr$PCRbias == "To_keep",]
 
  BeforeCollapse <- nrow(airr)
  
  airr <- airr %>%
    group_by(across(all_of(collapse_groups))) %>%
    group_modify(~ collapseDuplicates(
      .x,
      id = "sequence_id",
      seq = "sequence_alignment",
      text_fields = c("barcode", "clonotype_id"),
      num_fields = NULL,
      seq_fields = NULL,
      add_count = TRUE,
      ignore = c("N", "-", ".", "?"),
      sep = ",",
      dry = FALSE,
      verbose = FALSE
    )) %>%
    ungroup() %>%
    mutate(read_count = as.numeric(collapse_count)) %>%  # convert in numeric if necessary and rename in read_count
    mutate(unique_umi = barcode) %>%                      # rename
    select(-CollapseGroupId, -collapse_id, -collapse_class,
           -collapse_pass, -collapse_count, -PCRbias, -barcode)  # remove unecessary columns

  
  t22 <- Sys.time()
  TimeDiff<-t22-t11
  cat("Collapse duplicate sequences run time",paste(round(as.numeric(TimeDiff),2), attr(TimeDiff, "units")))
  cat("\n")

  # Add Nb umi to final repertoire
  airr$umi_count <- 1+str_count(airr$unique_umi,",")
  
  # Re-order the final repertoire table by umi_count per clone to have a repertoire ordered in the same way as MiXCR (top counted clone is first, etc)
  ## And rename clonotype_id accordingly (clonotype_id = chain + clone + n°)
  airr <- airr[order(airr$umi_count,decreasing = T),]
  
  airr$clonotype_id <- paste0(airr$locus,"clonotype",0:(nrow(airr)-1)) #adapt clonotype_id

  # Save the repertoire data after collapse duplicates
  AfterCollapse <- nrow(airr)
  Num<-AfterCollapse
  Prop <- round((Num/Input)*100,2)
  cat(paste0("Total CLONOTYPE after collapse duplicated reads : ",Num," (",Prop,"% input reads)"))
  cat("\n")

  ##### Fill metadata
  Metadata_QC[file,"Duplicate_reads"]<-BeforeCollapse-AfterCollapse  # fill Metadata
  
  ## Plot distribution of read count before filtering and summary 
  
  Plot <- ggplot(airr, aes(x = read_count)) +
            geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
            labs(title = paste("Read count distribution per unique sequence","-","Before filtering" ), x = "Read count", y = "Nb of unique sequences")+
            scale_y_continuous(trans = "log1p", breaks=c(0,1,10,100,1000,5000,10000)) + theme(panel.grid.minor = element_blank())
  theme_bw()

  ##store results
  ggsave(filename = paste0(res_folder,file,"_CollapseSeq_ReadCount_distrib_beforeFilter.pdf"),plot = Plot,width = 7,height = 5)

  Plot <- ggplot(airr, aes(x = umi_count)) +
            geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
            labs(title = paste("UMI count distribution per unique sequence","-","Before filtering" ), x = "UMI count", y = "Nb of unique sequences")+
            scale_y_continuous(trans = "log1p", breaks=c(0,1,10,100,1000,5000,10000)) + theme(panel.grid.minor = element_blank())+
  theme_bw()

  ##store results
  ggsave(filename = paste0(res_folder,file,"_CollapseSeq_UMICount_distrib_beforeFilter.pdf"),plot = Plot,width = 7,height = 5)
  
   #####  #####  Store an intermediate object containing only read and umi count to create an integrated plot after the loop
  Store_count<- airr[,c("read_count","umi_count")]
  Store_count$sampleid <- file
  saveRDS(object = Store_count,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_Collapse_duplicates_clonotypes_read_umi_count.rds"))
    #####   ##### 
  
  
#Same save but after filtering umi count 2
  AfterCollapseFilter2 <- airr[airr$umi_count > 1,]
  Num<-nrow(AfterCollapseFilter2)
  Prop <- round((Num/Input)*100,2)
  cat(paste0("Total CLONOTYPE after filter min 2 UMI counts : ",Num," (",Prop,"%)"))
  cat("\n")
  
  ##### Fill metadata
  Metadata_QC[file,"Only_1umi_clonotypes"]<-AfterCollapse-Num  # fill Metadata
  Metadata_QC[file,"Final_clonotypes"]<-Num  # fill Metadata

  # Print number of unique final clone per chain
  
  ### ON FILTERED TABLE 
  TableChain <- table(AfterCollapseFilter2$locus) #### AfterCollapseFilter2 or airr
  Total <- sum(TableChain)
  
  cat(paste0("Total CLONOTYPE after all quality filter : ",Total))
  cat("\n")
  
  Num<-ifelse(test = "IGH" %in% names(TableChain),yes = TableChain["IGH"],no = 0)
  Prop <- round((Num/Total)*100,2)
  cat(paste0("Total CLONOTYPE after all quality filter - IGH : ",Num," (",Prop,"%)"))
  cat("\n")
  Metadata_QC[file,"Final_clonotypes_IGH"]<-Num  # fill Metadata
  
  Num<-ifelse(test = "IGK" %in% names(TableChain),yes = TableChain["IGK"],no = 0)
  Prop <- round((Num/Total)*100,2)
  cat(paste0("Total CLONOTYPE after all quality filter - IGK : ",Num," (",Prop,"%)"))
  cat("\n")
  Metadata_QC[file,"Final_clonotypes_IGK"]<-Num  # fill Metadata
  
  Num<-ifelse(test = "IGL" %in% names(TableChain),yes = TableChain["IGL"],no = 0)
  Prop <- round((Num/Total)*100,2)
  cat(paste0("Total CLONOTYPE after all quality filter - IGL : ",Num," (",Prop,"%)"))
  cat("\n")
  Metadata_QC[file,"Final_clonotypes_IGL"]<-Num  # fill Metadata
  
  pal <- RColorBrewer::brewer.pal(n = 11,name = "Paired")
  pal <- c(IGL=pal[[1]],
           IGK=pal[[2]],
           IGH=pal[[6]])
  
  
  ## Plot chain percentage

  TablePlot <- c("IGH"=0,"IGK"=0,"IGL"=0)
  TablePlot[names(TableChain)] <-TableChain 
  TablePlotProp<-round((TablePlot/sum(TablePlot))*100,0)
  
  pdf(paste0(res_folder,file,"_CLONOTYPE_after_all_quality_filter_Chain_proportion.pdf"),width = 6, height = 5) # Save a pie plot in pdf
    pie(TablePlot, 
        labels = paste0(names(TablePlot)," n=",TablePlot,"(",TablePlotProp,"%)"), 
        main = "Chain proportion after all quality filter", 
        col = pal[names(TablePlot)])
  invisible(dev.off()) # close the pdf device without printing the message in the sink  
    
  

  ## Write final repertoire table 
  write_rearrangement(airr,file=paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_Collapse-unique_All-kept.tsv"))

  write_rearrangement(AfterCollapseFilter2,file=paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_Collapse-unique_at-least2-count.tsv"))


  #keep some memory
  ## remove all object but c("FileID","ImmcantationFolder","RawReadFolder","res_folder")
  rm(list=setdiff(ls(),c("FileID","ImmcantationFolder","res_folder","collapse_groups","t1","Metadata_QC")))
  gc()
  
  cat("\n")
}

  t2<-Sys.time()
  cat(paste(">>> Pipeline complete for all samples:",t2))
  cat("\n")
  DiffGlobal<-t2-t1
  cat(">>> Total runtime:",paste(round(as.numeric(DiffGlobal),2), attr(DiffGlobal, "units")))
  cat("\n")
  cat("\n")
  
sink()


# Save the Metadata in RDS to generate plot later
saveRDS(object = Metadata_QC,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","Metadata_Step3_CollapseDuplicates_All_105samples.rds"))
```

```{r,eval=FALSE}
# EDIT :: Reload Step3 metadata to add IGH isotypes counts per samples > to add in the report
Metadata_QC <- readRDS( paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","Metadata_Step3_CollapseDuplicates_All_105samples.rds"))

# Loop to reload the last table with final clonotypes and compute the number of clonotypes for each IGH isotypes
for(file in FileID){
  print(file)
  
  airr<-read_rearrangement(paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_Collapse-unique_at-least2-count.tsv"))
  
  # order IGH
  Order <- c("IGHM","IGHD","IGHG3","IGHG1","IGHA1","IGHG2","IGHG4","IGHE","IGHA2","IGH_NA")
  
  #Table isotypes
  TableIso <- table(airr$isotype)

  #fill all isotypes in the order on chromosomes
  for(iso in Order){
    Num<-ifelse(test = iso %in% names(TableIso),yes = TableIso[iso],no = 0)
    Metadata_QC[file,paste("Final_clonotypes",iso,sep = "_")]<-Num  # fill Metadata
  }
  
  #keep some memory
  ## remove all object but c("FileID","ImmcantationFolder","RawReadFolder","res_folder")
  rm(list=setdiff(ls(),c("FileID","ImmcantationFolder","res_folder","Metadata_QC")))
  gc()
}

# Save the Metadata in RDS to generate plot later
saveRDS(object = Metadata_QC,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","Metadata_Step3_CollapseDuplicates_All_105samples_WithIGHisotypes.rds"))

```


### Clustering sequences into clonal groups

For the file collapsed unique and after filtering 2 UMI Count

#### Calculate the Distance to the Nearest Distribution and find a threshold

```{r,eval=F}
  #create directory to store results per file
  if(!file.exists(paste0(res_folder,"Clone_Dist_Threshold"))) dir.create(paste0(res_folder,"Clone_Dist_Threshold")) 

# Open a file connection for writing a log file ##un comment if needed
sink(paste0(ImmcantationFolder,"pipeline_log/","R_Immunoverse_IgBlast_Alignment_NearestNeighbor.log"), append = FALSE, type = c("output"))

  t1<-Sys.time()
  cat(paste(">>> Pipeline started :",t1))
  cat("\n")

  list_plot<-list()
  
for(file in FileID){
  
  cat(file)
  cat("\n")
  
  # Do not need to adapt from here - path of Igblast quality filtered database
  file_path <- paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_Collapse-unique_at-least2-count.tsv")
  
  # Load airr 
  options(readr.show_progress = FALSE)
  airr <- read_rearrangement(file_path)
  
  airr_heavy <- airr %>%
  filter(locus == "IGH")
  
  cat("Start")
  cat("\n")
  t11 <- Sys.time()
  
  # Open a file connection for writing a log file ##un comment if needed
  sink(paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igh_dist_nearest_and_threshold_log.txt"), append = FALSE, type = c("output"))
  
  cat(paste("Number of unique IGH clonotypes :",nrow(airr_heavy)))
  cat("\n")
  
  cat(paste("Start distToNearest for file :",file))
  cat("\n")
  airr_heavy <- distToNearest(airr_heavy, sequenceColumn="junction",vCallColumn="v_call", jCallColumn="j_call", model="ham", first=FALSE, normalize="len", nproc=20)

  cat(paste("Start findThreshold for file :",file))
  cat("\n")
  
  threshold <- try(findThreshold(airr_heavy[["dist_nearest"]], method="density"),silent = T)
  if (inherits(threshold, "try-error")){
    clone_threshold <- NA
  }else{
    clone_threshold <- round(threshold@threshold,digits = 2) 
  }
  
  cat(paste("findThreshold - density method - round 2 digits :",clone_threshold))
  cat("\n")
  
  # Close the sink connection to save the output to the file ## un comment if needed
  sink()
  
  #Save the plot
  if (inherits(threshold, "try-error")){
    error_message <- conditionMessage(attr(threshold, "condition"))  # Extract error message
    # Create a text-based ggplot with the error message
    p <- ggplot() +
      annotate("text", x = 0.5, y = 0.5, label = error_message, size = 2, hjust = 0.5) +
      theme_void()+ # Remove axes and background
      labs(title = file) 
    
    ggsave(filename = paste0(res_folder,"Clone_Dist_Threshold/",file,"_Hamming_length_normalized_igh_threshold.pdf"),
           plot = p,
           width =3 ,height =2 )
  }else{
    p<-plot(threshold, binwidth=0.02, silent=T)+
      labs(title = file)
  
    ggsave(filename = paste0(res_folder,"Clone_Dist_Threshold/",file,"_Hamming_length_normalized_igh_threshold.pdf"),
           plot = p,
           width =3 ,height =2 )
  }

  list_plot[[file]]<-p
  # Save the repertoire data after collapse duplicates
  write_rearrangement(airr_heavy, file=paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_Collapse-unique_at-least2-count_IGH-dist.tsv"))
  
  t22 <- Sys.time()
  Diff<-t22-t11
  cat(file,">>> Total runtime:",paste(round(as.numeric(Diff),2), attr(Diff, "units")))
  cat("\n")
  
  cat("End")
  cat("\n")
  
  rm(list=setdiff(ls(),c("FileID","ImmcantationFolder","res_folder","t1","list_plot")))
  gc()
  
}

  t2<-Sys.time()
  cat(paste(">>> Pipeline complete for all samples:",t2))
  cat("\n")
  DiffGlobal<-t2-t1
  cat(">>> Total runtime:",paste(round(as.numeric(DiffGlobal),2), attr(DiffGlobal, "units")))
  cat("\n")
  cat("\n")
  
sink()

#global plot on 105 samples
patch<-patchwork::wrap_plots(list_plot)
ggsave(filename = paste0(res_folder,"Clone_Dist_Threshold/","Global_Patchork_allSamples_Hamming_length_normalized_igh_threshold.pdf"),
     plot = patch,
     width =25 ,height =17)
  
```

#### Define clones

Define clone according to previously defined threshold. Create germline according to clone information.

```{bash,eval=F}
# put the script in a sh executable

#!/bin/bash

# Activate conda environment
source PATH/conda.sh  # Adjust path if needed
conda activate pRESTO

Immcantation_dir="PATH/"

cd ${Immcantation_dir}

files=$(cat PATH/uniq_id_n105samples.txt)

# Start timing
start_time=$(date +%s)
start_date=$(date +"%Y-%m-%d %H:%M:%S")
echo ">>> Pipeline started at: $start_date"

for file in $files; do

  echo $file
  
  
  Threshlog="${Immcantation_dir}03_ChangeO/02_PostIgBlastFilter/${file}/${file}_igh_dist_nearest_and_threshold_log.txt"

  threshold=`grep "findThreshold - density method - round 2 digits : " $Threshlog | sed "s/findThreshold.*: //g"`
  
    # If threshold is NA or threshold is greater than 0.15, then replace by 0.15
  if [[ "$threshold" == "NA" || $(echo "$threshold > 0.15" | bc -l) -eq 1 ]]; then
    threshold=0.15
  fi
  
echo "Runing DefineClones.py"
DefineClones.py -d "${Immcantation_dir}03_ChangeO/02_PostIgBlastFilter/${file}/${file}_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_Collapse-unique_at-least2-count_IGH-dist.tsv" --model ham --norm len --dist $threshold --mode gene --act set --nproc 20 --outdir "${Immcantation_dir}03_ChangeO/02_PostIgBlastFilter/${file}/" --format airr --log "${Immcantation_dir}03_ChangeO/02_PostIgBlastFilter/${file}/${file}_DefineClone.log" >> "${Immcantation_dir}pipeline_log/${file}_ChangeO_and_PostProcess_perf.log" 2>&1

echo "Runing CreateGermlines.py"
CreateGermlines.py -d "${Immcantation_dir}03_ChangeO/02_PostIgBlastFilter/${file}/${file}_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_Collapse-unique_at-least2-count_IGH-dist_clone-pass.tsv" -r PATH/germlines/imgt/human/vdj -g dmask --format airr --cloned --outdir "${Immcantation_dir}03_ChangeO/02_PostIgBlastFilter/${file}/" >> "${Immcantation_dir}pipeline_log/${file}_ChangeO_and_PostProcess_perf.log" 2>&1

done


# End time and date
end_time=$(date +%s)
end_date=$(date +"%Y-%m-%d %H:%M:%S")
duration=$((end_time - start_time))
hours=$((duration / 3600))
minutes=$(( (duration % 3600) / 60 ))
seconds=$((duration % 60))

# Final report
echo ">>> Pipeline ended at: $end_date"
echo ">>> Total runtime: ${hours}h ${minutes}m ${seconds}s (${duration} seconds)"

# nano > Copy paste the script from the shebang to the done of the loop > DefineClones_loop_all_samples.sh

# As executable file
chmod +x DefineClones_loop_all_samples.sh

# run the loop as nohup so that it do not keyboad interupt when the session disconnect
nohup ./DefineClones_loop_all_samples.sh > pipeline_log/DefineClones_loop_nohup__all_samples_log.txt 2>&1 &


```

### Clonal lineage tree analysis

```{r,eval=FALSE}
library(dowser)
library(alakazam)
library(airr)
library(dplyr)

#Reload metadata
## reload from the step 3 (Collapse Duplicates) for all 105 samples
Metadata_QC <- readRDS( paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","Metadata_Step3_CollapseDuplicates_All_105samples.rds"))

# Add in the Metadata the threshold of nearest distance inferred for clonal grouping by density method / the threshold applied if modified

for(file in FileID){
  Threshlog <- read.delim(paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igh_dist_nearest_and_threshold_log.txt"),header = F,sep = ":")
  thresh<-as.numeric(gsub(" ","",Threshlog[grepl("density",Threshlog$V1),"V2"])) # take the threshold, remove space, convert numeric
  Metadata_QC[file,"Threshold_IGHclonal_grouping_inferred"]<-thresh
  
  if(is.na(thresh)){
    Metadata_QC[file,"Threshold_IGHclonal_grouping_applied"]<-0.15
  }else{
    if(thresh>0.15){
      Metadata_QC[file,"Threshold_IGHclonal_grouping_applied"]<-0.15
    }else{
      Metadata_QC[file,"Threshold_IGHclonal_grouping_applied"]<-thresh
    }
  }
}
```

```{r,eval=F}

# Open a file connection for writing a log file ##un comment if needed
sink(paste0(ImmcantationFolder,"pipeline_log/","R_Immunoverse_IGH_clonal_lineage_tree_dowser.log"), append = FALSE, type = c("output"))

  t1<-Sys.time()
  cat(paste(">>> Pipeline started :",t1))
  cat("\n")

for(file in FileID){
  
  cat(file)
  cat("\n")
  
  cat("Start")
  cat("\n")
  t11 <- Sys.time()
  
  # Do not need to adapt from here - path of Igblast quality filtered database
  file_path <- paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_Collapse-unique_at-least2-count_IGH-dist_clone-pass_germ-pass.tsv")
  
  # Load airr 
  options(readr.show_progress = FALSE)
  airr <- read_rearrangement(file_path)
  
  # Compute for the Metadata
  Num<-length(unique(airr$clone_id))
  cat(paste("number of unique clone_id :",Num))
  cat("\n")

  ##### Fill metadata
  Metadata_QC[file,"IGH_clones_Total"]<-Num  # fill Metadata
  
  # 1. Format
  
  airr$read_count <- as.numeric(airr$read_count)
  airr$umi_count <- as.numeric(airr$umi_count)
  
  clones <- try(formatClones(airr, traits=c("isotype"),num_fields=c("read_count","umi_count"),text_fields = c("clonotype_id","unique_umi"), columns=c("d_call"), minseq=2,nproc=20),
                silent = T)
  if (inherits(clones, "try-error")){
    # 0 clones with 2 clonotypes
    cat(paste("IGH clone with more than 1 clonotype :","0"))
    cat("\n")
    ##### Fill metadata
    Metadata_QC[file,"IGH_clones_with_more_1clonotype"]<-0  # fill Metadata
    
    # Num totol = num clones with 1 clonotypes
    cat(paste("IGH clone with only 1 clonotype :",Num))
    cat("\n")
    ##### Fill metadata
    Metadata_QC[file,"IGH_clones_with_only_1clonotype"]<-Num  # fill Metadata
  }else{
    # Num_clones_more than 1 seq
    Num_more1seq <- nrow(clones)
    cat(paste("IGH clone with more than 1 clonotype :",Num_more1seq))
    cat("\n")
    ##### Fill metadata
    Metadata_QC[file,"IGH_clones_with_more_1clonotype"]<-Num_more1seq  # fill Metadata
    
    # Only 1 seq = Num total - Num more than 1  
    Num_only1seq <-Num - Num_more1seq
    cat(paste("IGH clone with only 1 clonotype :",Num_only1seq))
    cat("\n")
    ##### Fill metadata
    Metadata_QC[file,"IGH_clones_with_only_1clonotype"]<-Num_only1seq  # fill Metadata
    
    # 2. Build the Trees
    # build maximum parsimony trees 
    clones <- getTrees(clones,nproc = 20)
    # clones <- scaleBranches(clones, edge_type = "mutations") # make a scale in mutations number (easier to interprete)
  
    # 3. Visualize
    # Plot the trees. Save them in a list of plots. 
    # Use tip metadata: isotype and collapse_count
    
    color_palette <- c("IGHM"="#253494", "IGHD"="#1D91C0", "IGHG1"="#d62728", "IGHG2"="#ff7f0e",  "IGHG3"="#FF6EC7", "IGHG4"="#FCCF05", "IGHA1"="#2ca02c", "IGHA2"="#36e40a","IGHE"="#8c564b","IGH_NA"="grey")
    
    tree_plots <- plotTrees(clones, tips="isotype", tipsize="umi_count", tip_palette=c(color_palette, "Germline"="#000000"),layout = "roundrect")
    names(tree_plots)<-paste0("IGHclone",clones$clone_id)
    
    ## Convert as dendrogram and save
    tree_plots <- lapply(tree_plots,function(plot){
      plot<-plot+
        ggtree::layout_dendrogram() # switch as a dendrogram with germline in the top
      
      plot
    })

    ####folder to save plot
  if(!file.exists(paste0(res_folder,"IGHclones/",file))) dir.create(paste0(res_folder,"IGHclones/",file)) 
    ### Save in the R session only the top 10 max, do not save for now the plot in pdf but keep the full object in RDS for latter plot saving if necessary
    if(length(tree_plots)>10){
      for(clnID in names(tree_plots)[1:10]){
        plot<-tree_plots[[clnID]]
        ggsave(plot = plot,
             filename = paste0(res_folder,"IGHclones/",file,"/","Dowser_MaxParsimonyTree_",clnID,".pdf"),
             width = 8,height = 4)
      }
    }else{
      for(clnID in names(tree_plots)){
        plot<-tree_plots[[clnID]]
        ggsave(plot = plot,
             filename = paste0(res_folder,"IGHclones/",file,"/","Dowser_MaxParsimonyTree_",clnID,".pdf"),
             width = 8,height = 4)
      }
    }
    
    #Save the RDS in 03_ChangeO/02_PostIgBlastFilter
    qsave(clones,paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_Dowser_MaxParsimonyTree_igh_clones_datatable.qs"))
    qsave(tree_plots, paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_Dowser_MaxParsimonyTree_igh_clones_treePlot.qs"))
  }

  t22 <- Sys.time()
  Diff<-t22-t11
  cat(file,">>> Total runtime:",paste(round(as.numeric(Diff),2), attr(Diff, "units")))
  cat("\n")
  
  cat("End")
  cat("\n")
  
  rm(list=setdiff(ls(),c("FileID","ImmcantationFolder","res_folder","Metadata_QC","t1")))
  gc()
  
}
  t2<-Sys.time()
  cat(paste(">>> Pipeline complete for all samples:",t2))
  cat("\n")
  DiffGlobal<-t2-t1
  cat(">>> Total runtime:",paste(round(as.numeric(DiffGlobal),2), attr(DiffGlobal, "units")))
  cat("\n")
  cat("\n")
  
sink()


# Save the Metadata in RDS to generate plot later
saveRDS(object = Metadata_QC,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","Metadata_Step4_IGHclones_All_105samples.rds"))
```


## Generate merged clonotype database

### All light and heavy chains clonotypes

```{r,eval=FALSE}
airr_imm_db<-list()
# Load all the final repertoire for all samples
for(file in FileID){
  tmp<-read_rearrangement(paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_Collapse-unique_at-least2-count.tsv"))
  tmp$sample_id <-file
  
  #Frequency in all repertoire (IGK+IGL+IGH)
  tmp$frequency_all_chain <-tmp$umi_count/sum(tmp$umi_count)
    
  # Frequency per chain 
  tmp <- split(tmp,tmp$locus)
  tmp<- lapply(tmp,function(sub){
    sub$frequency_per_locus <- sub$umi_count/sum(sub$umi_count)
    sub
  })
  tmp <- Reduce(rbind,tmp)
  #store in list
  airr_imm_db[[file]] <- tmp
}
airr_imm_db <- Reduce(rbind,airr_imm_db)
rm(tmp)
gc()

qsave(x = airr_imm_db,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20250714_Immcantation_clonotypes_merged_AllSamples.qs"))

## Renamed after anonymization of sample_id: 
# saveRDS(object = airr_imm_db,file = paste0("PATH/","20260724_Immcantation_clonotypes_IGHKL_merged_AllSamples_anonymized.rds"))

```

### IGH with clone id identifier

```{r,eval=F}
Igh_imm_db<-list()
# Load all the final repertoire for all samples
for(file in FileID){
  tmp<-read_rearrangement(paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/",file,"/",file,"_igblast-db_quality_filter-allFilter_germ-pass_chimeric-filter_Collapse-unique_at-least2-count_IGH-dist_clone-pass_germ-pass.tsv"))
  tmp$sample_id <-file
  
  #store in list
  Igh_imm_db[[file]] <- tmp
}
Igh_imm_db <- Reduce(rbind,Igh_imm_db)
rm(tmp)
gc()

qsave(x = Igh_imm_db,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20250714_Immcantation_IGH_clonal_grouped_merged_AllSamples.qs"))

## Renamed after anonymization of sample_id: 
# saveRDS(object = Igh_imm_db,file = paste0("PATH/","20260724_Immcantation_IGH_clonal_grouped_merged_AllSamples_anonymized.rds"))

```


## Clonal diversity and frequency - Hill numbers

### All IGH clones

Compute Hill diversity according to clone frequency unweighted (num of sequence per clone) or weighted by umi_count (sum of umi_count of all sequences in a clone). As it was done for clone frequency, compute both from uniformly ressampled (all samples with same size : minimum size) or not uniformly ressampled (all sample with their original size). 

NB : min_n increased to 1000 because it is recommended by the author to avoid artifacts in diversity calculations https://bitbucket.org/kleinstein/alakazam/issues/100/clonal-diversity 
Also to avoid a too hard downsampling of all samples

NB : At the end, I choosed the following parameters : unweighted (less biased by PC high counts) and uniformly resampled (uniform=T >> to correct the different sample sizes).

```{r,eval=F}
# Only 1 time to compute - then reload
# Create a sub annotated with all sample related metadata and filter only ICI
Igh_imm_db_ICIunique<-Igh_imm_db # NB here it correspond to the same table in input (metadata column are not necessary for clonal statistics analyis )

###########################################################################
###########################################################################
#### Edit alpha diversity according to copy/min_n/uniform?
###########################################################################
###########################################################################
var_to_iterate <- list(Unweighted_UniformlyResampled=list(min_n=1000,
                         uniform=T,
                         copy=c(NULL)),
                       Unweighted_OriginalSize=list(min_n=1000,
                         uniform=F, # not uniform
                         copy=c(NULL)),
                       umi_weighted_UniformlyResampled=list(min_n=6000,
                         uniform=T,
                         copy="umi_count"),
                       umi_weighted_OriginalSize=list(min_n=6000,
                         uniform=F, # not uniform
                         copy="umi_count"))


abund <- lapply(var_to_iterate,function(var_list){
  alakazam::estimateAbundance(Igh_imm_db_ICIunique, group="sample_id", nboot=100,
                              min_n = var_list[["min_n"]],
                              copy = var_list[["copy"]],
                              uniform = var_list[["uniform"]])
})

# NB : Warning message:
# In alakazam::estimateAbundance(Igh_imm_db_ICIunique, group = "sample_id",  :
#   Not all groups passed threshold min_n=1000. Excluded: PT_ZCG7VN_Tumor, PT_KQ9X6Y_Tumor


### Note : many clone imputed by alakazam to estimate and correct sampling bias : identified as "U..."

var_to_iterate<-setNames(names(var_to_iterate),nm = names(var_to_iterate)) # modify the vector to iterate on
div <- lapply(var_to_iterate,function(var_list){
  alakazam::alphaDiversity(abund[[var_list]], step_q=1, max_q=10)
})

#### Save estimateAbundance and alphaDiversity objects from Alakazam as qs to reload later if necessary
qsave(x = abund,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20250812_Immcantation_IGH_clonal_alakazamEstimateAbundance_list.qs"))
qsave(x = div,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20250812_Immcantation_IGH_clonal_alakazamAlphaDiversity_list.qs"))

#renamed after anonymization + SELECT only Unweighted+Uniform=T in parameter & abundance/diversity table in anonymization process (only table used at the end in the published figures)
# div <- div$Unweighted_UniformlyResampled
# abund_list <- abund_list$Unweighted_UniformlyResampled
# abund_list <- abund_list@abundance
# div <- div@diversity 
### renamed as follow :
# saveRDS(object = abund_list,file = paste0("PATH/","20260724_Immcantation_IGH_clonal_alakazamEstimateAbundance_Unweighted_Uniform_anonymized.rds"))
# 
# saveRDS(object = div,file = paste0("PATH/","20260724_Immcantation_IGH_clonal_alakazamAlphaDiversity_Unweighted_Uniform_anonymized.rds"))

```

### Only class switched IGH clones (remove IgM/IgD)

Perform again the comparaison but only for class switched isotypes - ie remove IgM and IgD clonotypes - to perform a more "fair" comparison of Tumor and Blood samples (hence blood samples are mainly composed of unswitched isotypes which is not the case for tumor samples that are mainly switched).

First filter IGHM/IGHD (and IGH_NA) clonotypes, filters also the samples previously filtered in the global analysis due to too low count of Ig clonotypes. Then compute abundance and diversity with different parameters (unweighted / umi_weighted ; downsampled or original size) + different filters bases on a minimum number of clone/sequences (min_n filter) : no_filter; filter on 1000 or 6000 clones/sequences.
Finally : choose Unweighted, Uniform=T (downsampled) & filter on 1000 sequences minimum.

```{r,eval=FALSE}

Igh_imm_db_ICIunique<-Igh_imm_db # NB here it correspond to the same table in input (metadata column are not necessary for clonal statistics analyis )


###########################################################################
###########################################################################
#### Edit alpha diversity according to copy/min_n/uniform?
###########################################################################
###########################################################################
var_to_iterate <- list(Unweighted_UniformlyResampled=list(min_n=1000,
                         uniform=T,
                         copy=c(NULL)),
                       Unweighted_OriginalSize=list(min_n=1000,
                         uniform=F, # not uniform
                         copy=c(NULL)),
                       umi_weighted_UniformlyResampled=list(min_n=6000,
                         uniform=T,
                         copy="umi_count"),
                       umi_weighted_OriginalSize=list(min_n=6000,
                         uniform=F, # not uniform
                         copy="umi_count"))


# Grouping by isotype & sample_id
## first remove IGH_NA
Igh_imm_db_ICIunique <- Igh_imm_db_ICIunique[Igh_imm_db_ICIunique$isotype != "IGH_NA",]
## Remove also IgM and IgD
Igh_imm_db_ICIunique <- Igh_imm_db_ICIunique[!(Igh_imm_db_ICIunique$isotype %in% c("IGHM","IGHD")),]

## remove the samples filtered in previous analysis according to the min_n threshold >> try to Remove the min_n threshold here to compute diversity per isotypes irrespective of its original counts ...
Igh_imm_db_ICIunique <- Igh_imm_db_ICIunique[!(Igh_imm_db_ICIunique$sample_id %in% c("PT_ZCG7VN_Tumor", "PT_KQ9X6Y_Tumor")), ] ### NB : edited here by anonymiz id

## compute abundance 
### without min_n filter (downsample on the minimum number of clone/sequences ...)
abund <- lapply(var_to_iterate,function(var_list){
  alakazam::estimateAbundance(Igh_imm_db_ICIunique, 
                              group="sample_id",
                              nboot=100,
                              min_n = 1, # no min_n filter
                              # min_n = var_list[["min_n"]],
                              copy = var_list[["copy"]],
                              uniform = var_list[["uniform"]])
})

### With min_n filter on 1000 clones / sequences >>> THIS IS THE FINAL PARAMETER USED
##### In alakazam::estimateAbundance(Igh_imm_db_ICIunique, group = "sample_id",  :
  # Not all groups passed threshold min_n=1000. Excluded 5 in more than previous 2 : PT_BEL9H5_Tumor, PT_41SWDR_PBMC_T0, PT_41SWDR_PBMC_T6, PT_9NCS0V_PBMC_HD, PT_529R1P_PBMC_HD
abund_min1000 <- lapply(var_to_iterate,function(var_list){
  alakazam::estimateAbundance(Igh_imm_db_ICIunique, 
                              group="sample_id",
                              nboot=100,
                              min_n = 1000, # no min_n filter
                              # min_n = var_list[["min_n"]],
                              copy = var_list[["copy"]],
                              uniform = var_list[["uniform"]])
})

### finally same but with min_n filter on 6000 clones/sequences (a bit too hard for unweighted but maybe relevent for umi_weighted ...)
# unweighted : In alakazam::estimateAbundance(Igh_imm_db_ICIunique, group = "sample_id",  :
#   Not all groups passed threshold min_n=6000. Excluded: a lot sample excluded [... edited but parameter not used ...]
# umi_weighted: In alakazam::estimateAbundance(Igh_imm_db_ICIunique, group = "sample_id",  :
#   Not all groups passed threshold min_n=6000. Excluded: PT_41SWDR_PBMC_T0, PT_41SWDR_PBMC_T6, PT_9NCS0V_PBMC_HD, PT_529R1P_PBMC_HD
abund_min6000 <- lapply(var_to_iterate,function(var_list){
  alakazam::estimateAbundance(Igh_imm_db_ICIunique, 
                              group="sample_id",
                              nboot=100,
                              min_n = 6000, # no min_n filter
                              # min_n = var_list[["min_n"]],
                              copy = var_list[["copy"]],
                              uniform = var_list[["uniform"]])
})

var_to_iterate<-setNames(names(var_to_iterate),nm = names(var_to_iterate)) # modify the vector to iterate on

# Hill Diversity
### withtout min_n
div <- lapply(var_to_iterate,function(var_list){
  print(paste("start",var_list))
  alakazam::alphaDiversity(abund[[var_list]], step_q=1, max_q=10)
})

### min_n 1000 >>> THIS IS THE FINAL PARAMETER USED
div_min1000 <- lapply(var_to_iterate,function(var_list){
  print(paste("start",var_list))
  alakazam::alphaDiversity(abund_min1000[[var_list]], step_q=1, max_q=10)
})

### min_n 6000
div_min6000 <- lapply(var_to_iterate,function(var_list){
  print(paste("start",var_list))
  alakazam::alphaDiversity(abund_min6000[[var_list]], step_q=1, max_q=10)
})

#### Save estimateAbundance and alphaDiversity objects from Alakazam as qs to reload later if necessary
# no min_n
qsave(x = abund,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260408_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamEstimateAbundance_no_minSeqFilter_list.qs"))
qsave(x = div,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260408_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamAlphaDiversity_no_minSeqFilter_list.qs"))

# min_n 1000 seq >>> THIS IS THE FINAL PARAMETER USED
qsave(x = abund_min1000,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260408_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamEstimateAbundance_min1000seqFilter_list.qs"))
qsave(x = div_min1000,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260408_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamAlphaDiversity_min1000seqFilter_list.qs"))


# min_ 6000 seq
qsave(x = abund_min6000,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260408_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamEstimateAbundance_min6000seqFilter_list.qs"))
qsave(x = div_min6000,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260408_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamAlphaDiversity_min6000seqFilter_list.qs"))




#renamed after anonymization + SELECT only Unweighted+Uniform=T + min_n=1000 in parameter & abundance/diversity table in anonymization process (only table used at the end in the published figures)
# div <- div$Unweighted_UniformlyResampled
# abund_list <- abund_list$Unweighted_UniformlyResampled
# abund_list <- abund_list@abundance
# div <- div@diversity 
### renamed as follow :
# saveRDS(object = abund_list,file = paste0("PATH/","20260724_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamEstimateAbundance_Unweighted_Uniform_min1000seqFilter_anonymized.rds"))
# 
# saveRDS(object = div,file = paste0("PATH/","20260724_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamAlphaDiversity_Unweighted_Uniform_min1000seqFilter_anonymized.rds"))

```



### Only IgGs or IgAs pures IGH clones

Same but measuring diversity of all IgGs or all IgAs.

Filter at 500 sequences instead of 1000 (to filter less samples because there are fewer sequences after removing all non IgG or all non IgA clones ...).


```{r,eval=FALSE}
Igh_imm_db_ICIunique<-Igh_imm_db # NB here it correspond to the same table in input (metadata column are not necessary for clonal statistics analyis )


###########################################################################
###########################################################################
#### Edit alpha diversity according to copy/min_n/uniform?
###########################################################################
###########################################################################
var_to_iterate <- list(Unweighted_UniformlyResampled=list(min_n=1000,
                         uniform=T,
                         copy=c(NULL)),
                       Unweighted_OriginalSize=list(min_n=1000,
                         uniform=F, # not uniform
                         copy=c(NULL)),
                       umi_weighted_UniformlyResampled=list(min_n=6000,
                         uniform=T,
                         copy="umi_count"),
                       umi_weighted_OriginalSize=list(min_n=6000,
                         uniform=F, # not uniform
                         copy="umi_count"))


# Grouping by isotype & sample_id
## first remove IGH_NA
Igh_imm_db_ICIunique <- Igh_imm_db_ICIunique[Igh_imm_db_ICIunique$isotype != "IGH_NA",]

## remove the samples filtered in previous analysis according to the min_n threshold >> try to Remove the min_n threshold here to compute diversity per isotypes irrespective of its original counts ...
Igh_imm_db_ICIunique <- Igh_imm_db_ICIunique[!(Igh_imm_db_ICIunique$sample_id %in% c("PT_ZCG7VN_Tumor", "PT_KQ9X6Y_Tumor")), ] ### NB : edited here by anonymiz id

# Vector of isotypes to iterate
iso_vect <- list(IgGs=c("IGHG1","IGHG2","IGHG3","IGHG4"),IgAs=c("IGHA1","IGHA2"))

## compute abundance 
### without min_n filter (downsample on the minimum number of clone/sequences ...)
abund <- lapply(iso_vect,function(iso){
  # subset correct isotypes
  tmp<-Igh_imm_db_ICIunique[Igh_imm_db_ICIunique$isotype %in% iso,] # keep only IgG or IgA
  #list of abundance
  lapply(var_to_iterate,function(var_list){
    alakazam::estimateAbundance(tmp, 
                                group="sample_id",
                                nboot=100,
                                min_n = 1, # no min_n filter
                                # min_n = var_list[["min_n"]],
                                copy = var_list[["copy"]],
                                uniform = var_list[["uniform"]])
  })
})


### With min_n filter=1000 / too many samples filtered
abund_min1000 <- lapply(iso_vect,function(iso){
  # subset correct isotypes
  tmp<-Igh_imm_db_ICIunique[Igh_imm_db_ICIunique$isotype %in% iso,] # keep only IgG or IgA
  #list of abundance
  lapply(var_to_iterate,function(var_list){
    alakazam::estimateAbundance(tmp, 
                                group="sample_id",
                                nboot=100,
                                min_n = 1000, # no min_n filter
                                # min_n = var_list[["min_n"]],
                                copy = var_list[["copy"]],
                                uniform = var_list[["uniform"]])
  })
})

### finally same but with min_n filter on 500 clones/sequences (a bit less hard)
# Warning messages:
# IgGs (unweighted + Uniform=T+min_n=500): In alakazam::estimateAbundance(tmp, group = "sample_id", nboot = 100,  :
#   Not all groups passed threshold min_n=500. Excluded 5 in more than previous 2: PT_J401JQ_PBMC_T6, PT_41SWDR_PBMC_T0, PT_41SWDR_PBMC_T6, PT_9NCS0V_PBMC_HD, PT_529R1P_PBMC_HD

# IgAs (unweighted + Uniform=T+min_n=500): In alakazam::estimateAbundance(tmp, group = "sample_id", nboot = 100,  :
#   Not all groups passed threshold min_n=500. Excluded 11 in more than previous 2: PT_Y6XN2H_PBMC_T6, PT_BEL9H5_Tumor, PT_41SWDR_PBMC_T0, PT_41SWDR_PBMC_T6, PT_41SWDR_Tumor, PT_DTP8BV_PBMC_T6, PT_22QBL1_PBMC_T6, PT_22QBL1_Tumor, PT_LGG57X_Tumor, PT_9NCS0V_PBMC_HD, PT_529R1P_PBMC_HD


abund_min500 <- lapply(iso_vect,function(iso){
  # subset correct isotypes
  tmp<-Igh_imm_db_ICIunique[Igh_imm_db_ICIunique$isotype %in% iso,] # keep only IgG or IgA
  #list of abundance
  lapply(var_to_iterate,function(var_list){
    alakazam::estimateAbundance(tmp, 
                                group="sample_id",
                                nboot=100,
                                min_n = 500, # no min_n filter
                                # min_n = var_list[["min_n"]],
                                copy = var_list[["copy"]],
                                uniform = var_list[["uniform"]])
  })
})


var_to_iterate<-setNames(names(var_to_iterate),nm = names(var_to_iterate)) # modify the vector to iterate on
iso_vect <- setNames(names(iso_vect),nm = names(iso_vect)) # modify the vector to iterate on

# Hill Diversity
### withtout min_n
div <- lapply(iso_vect,function(iso){
  # keep only correct iso
  abund_iso <- abund[[iso]]
  # list var
  lapply(var_to_iterate,function(var_list){
    print(paste("start",var_list))
    alakazam::alphaDiversity(abund_iso[[var_list]], step_q=1, max_q=10)
  })
})

### min_n 1000
div_min1000 <- lapply(iso_vect,function(iso){
  # keep only correct iso
  abund_iso <- abund_min1000[[iso]]
  # list var
  lapply(var_to_iterate,function(var_list){
    print(paste("start",var_list))
    alakazam::alphaDiversity(abund_iso[[var_list]], step_q=1, max_q=10)
  })
})

### min_n 500
div_min500 <- lapply(iso_vect,function(iso){
  # keep only correct iso
  abund_iso <- abund_min500[[iso]]
  # list var
  lapply(var_to_iterate,function(var_list){
    print(paste("start",var_list))
    alakazam::alphaDiversity(abund_iso[[var_list]], step_q=1, max_q=10)
  })
})

#### Save estimateAbundance and alphaDiversity objects from Alakazam as qs to reload later if necessary
# no min_n
qsave(x = abund,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260420_Immcantation_IGH_IgGs_or_IgAs_clonal_alakazamEstimateAbundance_no_minSeqFilter_list.qs"))
qsave(x = div,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260420_Immcantation_IGH_IgGs_or_IgAs_clonal_alakazamAlphaDiversity_no_minSeqFilter_list.qs"))

# min_n 1000 seq
qsave(x = abund_min1000,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260420_Immcantation_IGH_IgGs_or_IgAs_clonal_alakazamEstimateAbundance_min1000seqFilter_list.qs"))
qsave(x = div_min1000,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260420_Immcantation_IGH_IgGs_or_IgAs_clonal_alakazamAlphaDiversity_min1000seqFilter_list.qs"))


# min_ 500 seq >>>>> THIS ONE WAS SELECTED FOR ANALYSIS
qsave(x = abund_min500,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260420_Immcantation_IGH_IgGs_or_IgAs_clonal_alakazamEstimateAbundance_min500seqFilter_list.qs"))
qsave(x = div_min500,file = paste0(ImmcantationFolder,"03_ChangeO/02_PostIgBlastFilter/","20260420_Immcantation_IGH_IgGs_or_IgAs_clonal_alakazamAlphaDiversity_min500seqFilter_list.qs"))




#renamed after anonymization + SELECT only Unweighted+Uniform=T + min_n=500 in parameter & abundance/diversity table in anonymization process (only table used at the end in the published figures)
# div <- div$Unweighted_UniformlyResampled
# abund_list <- abund_list$Unweighted_UniformlyResampled
# abund_list <- abund_list@abundance
# div <- div@diversity 

### renamed as follow :
#### For IgGs
# saveRDS(object = abund_list,file = paste0("PATH/","20260724_Immcantation_IGH_IgGs_clonal_alakazamEstimateAbundance_Unweighted_Uniform_min500seqFilter_anonymized.rds"))
# 
# saveRDS(object = div,file = paste0("PATH/","20260724_Immcantation_IGH_IgGs_clonal_alakazamAlphaDiversity_Unweighted_Uniform_min500seqFilter_anonymized.rds"))

#### For IgAs
# saveRDS(object = abund_list,file = paste0("PATH/","20260724_Immcantation_IGH_IgAs_clonal_alakazamEstimateAbundance_Unweighted_Uniform_min500seqFilter_anonymized.rds"))
# 
# saveRDS(object = div,file = paste0("PATH/","20260724_Immcantation_IGH_IgAs_clonal_alakazamAlphaDiversity_Unweighted_Uniform_min500seqFilter_anonymized.rds"))

```


# Reload merged dataset

Reload data

```{r}
# Metadata ANONYMIZED - NB : Response to ICI is not shared due to patient privacy requirement but can be asked to the study sponsor unpon reasonable request (see article for contact)
# Immcantation_meta_ICI_unique <-readRDS(paste0("PATH/","20260723_Immunoverse_Metadata_AnonymId_ICIresponse.rds")) # edit : study sponsor accepted to share the information
Immcantation_meta_ICI_unique <- read.csv(paste0(res_folder,"20260723_Immunoverse_Metadata_AnonymId_ICIresponse.csv"),row.names = 1) # use csv since EGA refused rds
## Generate back factors ...
Immcantation_meta_ICI_unique$sample_origin <- factor(Immcantation_meta_ICI_unique$sample_origin,levels = c("Tumor" ,"PBMC" ))
Immcantation_meta_ICI_unique$sample_timepoint <- factor(Immcantation_meta_ICI_unique$sample_timepoint,levels = c("T0" ,"T6" ))
Immcantation_meta_ICI_unique$sample_type <- factor(Immcantation_meta_ICI_unique$sample_type,levels = c("Tumor.T0" ,"PBMC.T0" ,"PBMC.T6"))
Immcantation_meta_ICI_unique$sample_type2 <- factor(Immcantation_meta_ICI_unique$sample_type2,levels = c("Tumor.T0" ,"PBMC.T0" ,"PBMC.T6","PBMC.HD"))
Immcantation_meta_ICI_unique$R_NR <- factor(Immcantation_meta_ICI_unique$R_NR,levels = c("NR" ,"R" ,"HD"))


### airr dataset of all clonotypes (light chain and heavy chains) with anonymized id - lot of  memory necessary 
#### Not directly necessary to reproduce plot because count of clonotypes are in the metadata
# airr_imm_db<-readRDS(paste0(res_folder,"20260724_Immcantation_clonotypes_IGHKL_merged_AllSamples_anonymized.rds"))
# airr_imm_db <- airr::read_rearrangement(paste0(res_folder,"20260724_Immcantation_clonotypes_IGHKL_merged_AllSamples_anonymized.tsv")) # use tsv since EGA refused rds


#### airr dataset of all IGH clonotypes with clone_id identifier
# Igh_imm_db<-readRDS(paste0(res_folder,"20260724_Immcantation_IGH_clonal_grouped_merged_AllSamples_anonymized.rds"))
Igh_imm_db <- airr::read_rearrangement(paste0(res_folder,"20260724_Immcantation_IGH_clonal_grouped_merged_AllSamples_anonymized.tsv")) # use tsv since EGA refused rds


# Create a sub annotated with all sample related metadata and filter only ICI
Igh_imm_db_ICIunique<-merge(Igh_imm_db,Immcantation_meta_ICI_unique[,c("sample_id","patient_id","sample_origin","sample_timepoint","sample_type" ,"sample_type2","R_NR")],by="sample_id")

#reload clone abundance and diversity for all IGH clones
# abund_allIGH <- readRDS(paste0(res_folder,"20260724_Immcantation_IGH_clonal_alakazamEstimateAbundance_Unweighted_Uniform_anonymized.rds"))
abund_allIGH <- read.csv(paste0(res_folder,"20260724_Immcantation_IGH_clonal_alakazamEstimateAbundance_Unweighted_Uniform_anonymized.csv")) # use csv since EGA refused rds
# div_allIGH <- readRDS(paste0(res_folder,"20260724_Immcantation_IGH_clonal_alakazamAlphaDiversity_Unweighted_Uniform_anonymized.rds"))
div_allIGH <- read.csv(paste0(res_folder,"20260724_Immcantation_IGH_clonal_alakazamAlphaDiversity_Unweighted_Uniform_anonymized.csv")) # use csv since EGA refused rds



#reload clone abundance and diversity only for class-switched IGH clones
# abund_ClassSwitch <- readRDS(paste0(res_folder,"20260724_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamEstimateAbundance_Unweighted_Uniform_min1000seqFilter_anonymized.rds"))
abund_ClassSwitch <- read.csv(paste0(res_folder,"20260724_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamEstimateAbundance_Unweighted_Uniform_min1000seqFilter_anonymized.csv")) # use csv since EGA refused rds
# div_ClassSwitch <- readRDS(paste0(res_folder,"20260724_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamAlphaDiversity_Unweighted_Uniform_min1000seqFilter_anonymized.rds"))
div_ClassSwitch <- read.csv(paste0(res_folder,"20260724_Immcantation_IGH_ONLY_SWITCHED_clonal_alakazamAlphaDiversity_Unweighted_Uniform_min1000seqFilter_anonymized.csv")) # use csv since EGA refused rds


#reload clone abundance and diversity only for IgGs only IGH clones
# abund_IgGs <- readRDS(paste0(res_folder,"20260724_Immcantation_IGH_IgGs_clonal_alakazamEstimateAbundance_Unweighted_Uniform_min500seqFilter_anonymized.rds"))
abund_IgGs <- read.csv(paste0(res_folder,"20260724_Immcantation_IGH_IgGs_clonal_alakazamEstimateAbundance_Unweighted_Uniform_min500seqFilter_anonymized.csv")) # use csv since EGA refused rds
# div_IgGs <- readRDS(paste0(res_folder,"20260724_Immcantation_IGH_IgGs_clonal_alakazamAlphaDiversity_Unweighted_Uniform_min500seqFilter_anonymized.rds"))
div_IgGs <- read.csv(paste0(res_folder,"20260724_Immcantation_IGH_IgGs_clonal_alakazamAlphaDiversity_Unweighted_Uniform_min500seqFilter_anonymized.csv")) # use csv since EGA refused rds


#reload clone abundance and diversity only for IgAs only IGH clones
# abund_IgAs <- readRDS(paste0(res_folder,"20260724_Immcantation_IGH_IgAs_clonal_alakazamEstimateAbundance_Unweighted_Uniform_min500seqFilter_anonymized.rds"))
abund_IgAs <- read.csv(paste0(res_folder,"20260724_Immcantation_IGH_IgAs_clonal_alakazamEstimateAbundance_Unweighted_Uniform_min500seqFilter_anonymized.csv")) # use csv since EGA refused rds
# div_IgAs <- readRDS(paste0(res_folder,"20260724_Immcantation_IGH_IgAs_clonal_alakazamAlphaDiversity_Unweighted_Uniform_min500seqFilter_anonymized.rds"))
div_IgAs <- read.csv(paste0(res_folder,"20260724_Immcantation_IGH_IgAs_clonal_alakazamAlphaDiversity_Unweighted_Uniform_min500seqFilter_anonymized.csv")) # use csv since EGA refused rds

```



# Figures

## Figure 4A - Experimental scheme (biorender)

**Figure 4A compressed > see paper ...** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"Figure4A_to_render_compressed.pdf"),density = 300)
plot(img)
```


## Figure 4B - Ig clonotypes in each sample type


```{r}
# Number of clonotype Sample_type in full cohort 

## get clonotypes column
Clonotypes_col <- grep("clonotypes",colnames(Immcantation_meta_ICI_unique),value = T)

## melt to plot
melt_df<-Immcantation_meta_ICI_unique[,c("sample_id","sample_type","sample_type2","R_NR",Clonotypes_col)]
colnames(melt_df)<- gsub("IG_","",gsub("Final_clonotypes","IG",colnames(melt_df))) ### simplify variables names
melt_df <- reshape::melt.data.frame(melt_df,id.vars =c("sample_id","sample_type","sample_type2","R_NR") )


### With HD PBMC
df<-melt_df[(melt_df$variable != "IGH_NA"),]
df<-df[(df$variable != "IG"),] # remove also IG total
df_n<-unique(df[,c("sample_id","sample_type","R_NR")])
df_n<-table(df_n$R_NR,df_n$sample_type)

##### compare sample_type2
  p<-ggplot(df,aes(x=sample_type2,y=value))+
    geom_violin(aes(fill=sample_type2),draw_quantiles = 0.5)+
    scale_fill_manual(values = ColorPal$sample_type2)+
    geom_jitter(size=0.7)+
    facet_wrap(~variable,nrow = 2)+
    scale_y_continuous(trans = "log1p", breaks = base_breaksLog1p(),labels = prettyNum)+
    stat_compare_means(size=3,label = "p.format")+
    theme_bw()+
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
    labs(x=NULL,y=paste("Number of clonotypes"),
         title = "IG clonotypes in Tumor and PBMC samples from ICI treated patients or healthy donors",
         subtitle = "test : Kruskal-Wallis (>2 groups) or Wilcoxon-Mann-Whitney (2 groups)",
          caption= paste(paste(rep(colnames(df_n),each=3),rownames(df_n),sep = "-"),as.numeric(df_n),sep = ":",collapse = "; ") )
  
   ## This is Main Figure 4B
  
        ggsave(plot = p,
           filename = paste0(res_folder,"Figure4B_Immcantation_Number_clonotypes_all_chains_PBMC_Tumor_Compare_sampleTypes_noSplitResponse_V2ligns.pdf"),
           width = 11,height = 6)
  
```


**Figure 4B** 
```{r,fig.height=7,fig.width=12}
img <- magick::image_read_pdf(paste0(res_folder,"Figure4B_Immcantation_Number_clonotypes_all_chains_PBMC_Tumor_Compare_sampleTypes_noSplitResponse_V2ligns.pdf"),density = 300)
plot(img)
```

## Figure S9A - Ig clonotypes in each sample type vs reponse to ICI


```{r}
# Number of clonotype Sample_type in full cohort 

## get clonotypes column
Clonotypes_col <- grep("clonotypes",colnames(Immcantation_meta_ICI_unique),value = T)

## melt to plot
melt_df<-Immcantation_meta_ICI_unique[,c("sample_id","sample_type","sample_type2","R_NR",Clonotypes_col)]
colnames(melt_df)<- gsub("IG_","",gsub("Final_clonotypes","IG",colnames(melt_df))) ### simplify variables names
melt_df <- reshape::melt.data.frame(melt_df,id.vars =c("sample_id","sample_type","sample_type2","R_NR") )


### With HD PBMC
df<-melt_df[(melt_df$variable != "IGH_NA"),]
df<-df[(df$variable != "IG"),] # remove also IG total
df_n<-unique(df[,c("sample_id","sample_type","R_NR")])
df_n<-table(df_n$R_NR,df_n$sample_type)
p<-ggplot(df,aes(x=R_NR,y=value))+
    geom_violin(aes(fill=R_NR),draw_quantiles = 0.5)+
    scale_fill_manual(values = ColorPal$R_NR)+
    geom_jitter(size=0.7)+
    facet_grid(sample_type~variable,scales="free")+
    scale_y_continuous(trans = "log1p", breaks = base_breaksLog1p(),labels = prettyNum)+
    stat_compare_means(size=3,label = "p.format")+
    theme_bw()+
    labs(x=NULL,y=paste("Number of clonotypes"),
         title = "IG clonotypes in Tumor and PBMC samples from ICI treated patients or healthy donors - Immcantation pipeline",
         subtitle = "test : Kruskal-Wallis (>2 groups) or Wilcoxon-Mann-Whitney (2 groups)",
         caption = paste(paste(rep(colnames(df_n),each=3),rownames(df_n),sep = "-"),as.numeric(df_n),sep = ":",collapse = "; ") )


    ## This is Supplementary Figure S9A
    ggsave(plot = p,
           filename = paste0(res_folder,"S9A_Immcantation_Number_clonotypes_all_chains_PBMC_Tumor_R_vs_NR.pdf"),
           width = 10,height = 8)


```


**Figure S9A** 
```{r,fig.height=7,fig.width=12}
img <- magick::image_read_pdf(paste0(res_folder,"S9A_Immcantation_Number_clonotypes_all_chains_PBMC_Tumor_R_vs_NR.pdf"),density = 300)
plot(img)
```



## Figure 4C - Ig clones diversity in each sample type

```{r}
# Plot diversity R/NR or sample_type for all kind of parameters
## NB : initially a loop on all parameters. Here only the unweighted and uniform=T parameter are selected
# var_to_iterate<-setNames(names(div),nm = names(div)) # modify the vector to iterate on

var<-"Unweighted_UniformlyResampled"
# for(var in var_to_iterate){
  hill_div <- div_allIGH # Diversity of all IGH clones with Unweighted and Uniform=T and min_n=1000 parameters

  ## Merge back with metadata info
  hill_div<-merge(hill_div,Immcantation_meta_ICI_unique[,c("sample_id","patient_id","sample_origin","sample_timepoint","sample_type","sample_type2","R_NR")],by="sample_id")
 
  ## convert q in a factor rather than a numeric for plot
  hill_div$q<-factor(hill_div$q)
  

  ##################
  #####Tumor vs PBMC 
  ################

  ## Compute wilcox/kruskal test between Diversity~R_NR for each value of q
  pval_df <- hill_div %>%
    group_by(q) %>%
    summarise(
      n_groups = n_distinct(sample_type2),
      p_value = if (n_groups < 2) {
        NA_real_
      } else if (n_groups == 2) {
        tryCatch(wilcox.test(d ~ sample_type2)$p.value, error = function(e) NA_real_)
      } else {
        tryCatch(kruskal.test(d ~ sample_type2)$p.value, error = function(e) NA_real_)
      },
      .groups = "drop"
    ) %>%
    mutate(
      p_label = case_when(
        is.na(p_value) ~ "NA",
        p_value < 0.001 ~ "***",
        p_value < 0.01 ~ "**",
        p_value < 0.05 ~ "*",
        TRUE ~ "ns"
      )
    )
  
  ## add a y value to position the pvalue on plot on top +Inf
  pval_df$d<-+Inf
  
  p<-ggplot(hill_div, aes(x = q, y = d ,color=sample_type2)) +
    geom_line(aes(group = sample_id),alpha = 0.3) +
    scale_color_manual(values = ColorPal$sample_type2)+
    stat_summary(aes(group = sample_type2), fun = median, geom = "line", linewidth = 1.5) +
    labs(title = "Alpha Diversity (Hill numbers) by Sample type", x = "q", y = "Diversity",
         subtitle = paste("IGH clone frequency",var))+  
    # facet_grid(.~R_NR,scales = "free_x",space = "free_x")+
    geom_text(data = pval_df,aes(label = p_label),color = "black",size = 4,vjust=+1)+
    # stat_compare_means(label = "p.format")+
    theme_minimal()
  
  ##### THIS IS MAIN FIGURE 4C
  
  ggsave(plot = p,
         filename = paste0(res_folder,"Figure4C_Immcantation_AlphaDiversity_",var,"_DiversityCurves_split_Response_compare_SampleType.pdf"),
         height = 4,width = 6)

#}

```

**Figure 4C** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"Figure4C_Immcantation_AlphaDiversity_Unweighted_UniformlyResampled_DiversityCurves_split_Response_compare_SampleType.pdf"),density = 300)
plot(img)
```


## Figure S9B - Ig clones diversity in each sample type vs reponse to ICI

```{r}
# Plot diversity R/NR or sample_type for all kind of parameters
## NB : initially a loop on all parameters. Here only the unweighted and uniform=T parameter are selected
# var_to_iterate<-setNames(names(div),nm = names(div)) # modify the vector to iterate on

var<-"Unweighted_UniformlyResampled"
# for(var in var_to_iterate){
  hill_div <- div_allIGH # Diversity of all IGH clones with Unweighted and Uniform=T and min_n=1000 parameters

  ## Merge back with metadata info
  hill_div<-merge(hill_div,Immcantation_meta_ICI_unique[,c("sample_id","patient_id","sample_origin","sample_timepoint","sample_type","sample_type2","R_NR")],by="sample_id")
 
  ## convert q in a factor rather than a numeric for plot
  hill_div$q<-factor(hill_div$q)
  
  ## Compute wilcox/kruskal test between Diversity~R_NR for each value of q
  pval_df <- hill_div %>%
    group_by(sample_type, q) %>%
    summarise(
      n_groups = n_distinct(R_NR),
      p_value = if (n_groups < 2) {
        NA_real_
      } else if (n_groups == 2) {
        tryCatch(wilcox.test(d ~ R_NR)$p.value, error = function(e) NA_real_)
      } else {
        tryCatch(kruskal.test(d ~ R_NR)$p.value, error = function(e) NA_real_)
      },
      .groups = "drop"
    ) %>%
    mutate(
      p_label = case_when(
        is.na(p_value) ~ "NA",
        p_value < 0.001 ~ "***",
        p_value < 0.01 ~ "**",
        p_value < 0.05 ~ "*",
        TRUE ~ "ns"
      )
    )
  
  ## add a y value to position the pvalue on plot on top +Inf
  pval_df$d<-+Inf
  
  # diversity curve
  ## R vs NR
  p<-ggplot(hill_div, aes(x = q, y = d,color=R_NR)) +
    geom_line(aes(group = sample_id),alpha = 0.3) +
    scale_color_manual(values = ColorPal$R_NR)+
    stat_summary(aes(group = R_NR), fun = median, geom = "line", linewidth = 1.5) +
    labs(title = "Alpha Diversity (Hill numbers) by Response Group", x = "q", y = "Diversity",
         subtitle = paste("IGH clone frequency",var))+  
    facet_grid(.~sample_type,scales = "free_x",space = "free_x")+
    geom_text(data = pval_df,aes(label = p_label),color = "black",size = 4,vjust=+1)+
    theme_minimal() 
    
  
  #### THIS IS SUPPLEMENTARY FIGURE S9B
  
  ggsave(plot = p,
         filename = paste0(res_folder,"S9B_Immcantation_AlphaDiversity_",var,"_DiversityCurves_split_SampleType_compare_Response.pdf"),
         height = 4,width = 8)
  

#}
```

**Figure S9B** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"S9B_Immcantation_AlphaDiversity_Unweighted_UniformlyResampled_DiversityCurves_split_SampleType_compare_Response.pdf"),density = 300)
plot(img)
```

## Figure 4D, 4E, 4F, & S9C - Ig clones frequency in each sample type


```{r}
# Plot diversity R/NR or sample_type for all kind of parameters
## NB : Here only the unweighted and uniform=T parameter are selected

var<-"Unweighted_UniformlyResampled"

# Get correct abundance table
clone_frequency <- abund_allIGH # abundance of all IGH clones with Unweighted and Uniform=T and min_n=1000 parameters

## Merge back with metadata info
clone_frequency<-merge(clone_frequency,Immcantation_meta_ICI_unique[,c("sample_id","patient_id","sample_origin","sample_timepoint","sample_type","sample_type2","R_NR")],by="sample_id")


###########################################################################
# Clonal space (bining)
###########################################################################
## NB : sums the fre- quencies in each bin
### binning the clone frequencies into the bins [0.0, 0.001], [0.001, 0.01], [0.01, 0.1], and [0.1, 1]
###  Binning by rank is an alternative where the bins [1, 10], [11, 100], [101, 1000], and [1001, inf]

### binning the clone frequencies
bin_frequency <- clone_frequency
bin_frequency<-tapply(bin_frequency$p,bin_frequency$sample_id,function(freq_sample){
    bin_0_0001<-sum(freq_sample[freq_sample<=0.001])
    bin_0001_001<-sum(freq_sample[freq_sample>0.001 & freq_sample<=0.01])
    bin_001_01<-sum(freq_sample[freq_sample>0.01 & freq_sample<=0.1])
    bin_01_1<-sum(freq_sample[freq_sample>0.1])
    mat<-as.matrix(c("0.0, 0.001"=bin_0_0001,
             "0.001, 0.01"=bin_0001_001,
             "0.01, 0.1"=bin_001_01,
             "0.1, 1"=bin_01_1))
    colnames(mat)<-"frequency"
    # mat<-data.frame(mat)
    return(mat)
})

bin_frequency<-lapply(names(bin_frequency),function(nam){
  mat<-bin_frequency[[nam]]
  colnames(mat)<-nam
  mat
})

bin_frequency<-Reduce(cbind,bin_frequency)
bin_frequency<-as.data.frame(bin_frequency)

bin_frequency$bin_frequency <-rownames(bin_frequency)
bin_frequency$bin_frequency <- factor(bin_frequency$bin_frequency)

bin_frequency<-reshape::melt.data.frame(bin_frequency,id.vars = "bin_frequency")

## re merge with metadata
bin_frequency$sample_id<-bin_frequency$variable
bin_frequency<-merge(bin_frequency,Immcantation_meta_ICI_unique[,c("sample_id","patient_id","sample_origin","sample_timepoint","sample_type","sample_type2","R_NR")],by="sample_id")

## Order by R_NR - Stacked barplot
sort_id <-unique(as.character(bin_frequency$sample_id)[order(bin_frequency$sample_type2,bin_frequency$R_NR,bin_frequency$sample_id)])
bin_frequency$sample_id <- factor(as.character(bin_frequency$sample_id),
                                  levels =sort_id )

## create an anonym sample_id with Response type and then n° of patient (random ... just to simplify plot)
bin_frequency$patient_id2 <- paste0("P",as.numeric(factor(bin_frequency$patient_id))) # numero of patient
bin_frequency$sample_id2 <- paste(bin_frequency$R_NR,bin_frequency$patient_id2,sep = "_")

p<-ggplot(bin_frequency,aes(x =sample_id2,y= value))+
  geom_col(mapping = aes(fill=bin_frequency),color="black",position = "stack") +  # <- stack barplot
  # geom_text(data = bin_frequency_label,mapping = aes(label=R_NR),size=3,angle=90,hjust=-0.2)+
  facet_wrap(~sample_type2,scales = "free_x",drop = T,nrow = 2)+
  theme_minimal()+
  # scale_fill_manual(values = RColorBrewer::brewer.pal(length(levels(bin_frequency$bin_frequency)),"Spectral"))+
    # scale_fill_manual(values = rev(RColorBrewer::brewer.pal(length(levels(bin_frequency$bin_frequency)),"Spectral")))+
  scale_fill_viridis(option = "C",discrete = T)+
  # theme(axis.text.x = element_blank())+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1,size=6))+
  # scale_x_discrete(labels = bin_frequency$sample_id2)
  labs(y="Clone frequency",title = "Bin clone frequency",x=NULL,fill="clone frequency",
       subtitle = "Unweighted & uniformly resampled to min_n=1000")

###### THIS IS MAIN FIGURE 4D (NB: order of barplot not exactly the same after anonymization of sample ... not really important actually ... this is the same information)
ggsave(plot = p,
       filename = paste0(res_folder,"Figure4D_Immcantation_ClonalFrequency_Unweighted_UniformlyResampled_BinFrequency_BarplotStacked.pdf"),
       height = 5,width = 7)

### Statistical comparisons : boxplot
#### Tumor vs PBMC T0 vs PBMC T6
p<-ggplot(bin_frequency,aes(x =bin_frequency,y= value,fill=sample_type2))+
  geom_boxplot(outlier.size = 0.5,outlier.colour = "grey")+
  geom_point(size=0.5,position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75))+
  scale_fill_manual(values = ColorPal$sample_type2)+
  # scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum)+
  # facet_grid(.~R_NR,scales = "free_x",space = "free_x")+
  stat_compare_means(label = "p.format")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  # theme(legend.position = "none")+
  labs(y="Clone frequency",title = "Bin clone frequency",x="Clone frequency categories",
       subtitle = "Unweighted & uniformly resampled to min_n=1000")

###### THIS IS MAIN FIGURE 4E

ggsave(plot = p,
       filename = paste0(res_folder,"Figure4E_Immcantation_ClonalFrequency_Unweighted_UniformlyResampled_BinFrequency_compare_SampleType.pdf"),
       height = 4,width = 6)

#### R vs NR
p<-ggplot(bin_frequency,aes(x =bin_frequency,y= value,fill=R_NR))+
  geom_boxplot(outlier.size = 0.5,outlier.colour = "grey")+
  geom_point(size=0.5,position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75))+
  scale_fill_manual(values = ColorPal$R_NR)+
  # scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum)+
  facet_grid(.~sample_type,scales = "free_x",space = "free_x")+
  stat_compare_means(label = "p.signif")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  # theme(legend.position = "none")+
  labs(y="Clone frequency",title = "Bin clone frequency",x="Clone frequency categories",
       subtitle = "Unweighted & uniformly resampled to min_n=1000")

###### THIS IS SUPPLEMENTARY FIGURE 9C

ggsave(plot = p,
       filename = paste0(res_folder,"S9C_Immcantation_ClonalFrequency_Unweighted_UniformlyResampled_BinFrequency_split_SampleType_compare_Response.pdf"),
       height = 4,width = 5)

###########################################################################
# Top clones
###########################################################################
## top 10
clone_frequency_top10<-clone_frequency
clone_frequency_top10$rank_character<-as.character(clone_frequency_top10$rank)
clone_frequency_top10 <- clone_frequency_top10[clone_frequency_top10$rank_character %in% as.character(1:10),]#filter top 10
clone_frequency_top10$rank_character <- factor(clone_frequency_top10$rank_character,levels = as.character(1:10)) # as factor

### Tumor vs PBMC
p<-ggplot(clone_frequency_top10,aes(x =rank_character,y= p,fill=sample_type2))+
  geom_boxplot(outlier.size = 0.5,outlier.colour = "grey")+
  geom_point(size=0.4,position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75),color="darkgrey")+
  scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum)+
  scale_fill_manual(values = ColorPal$sample_type2)+
  # facet_grid(.~R_NR,scales = "free_x",space = "free_x")+
  stat_compare_means(label = "p.signif")+
  theme_minimal()+
  # theme(legend.position = "none")+
  labs(y="Clone frequency",title = "Rank clone abundance - Top 10 clones",
       x="rank top10 clones",
       subtitle = "Unweighted & uniformly resampled to min_n=1000")

###### THIS IS MAIN FIGURE 4F

ggsave(plot = p,
       filename = paste0(res_folder,"Figure4F_Immcantation_ClonalFrequency_Unweighted_UniformlyResampled_Top10clone_split_Response_compare_SampleType.pdf"),
       height = 4,width = 6)

```


**Figure 4D** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"Figure4D_Immcantation_ClonalFrequency_Unweighted_UniformlyResampled_BinFrequency_BarplotStacked.pdf"),density = 300)
plot(img)
```


**Figure 4E** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"Figure4E_Immcantation_ClonalFrequency_Unweighted_UniformlyResampled_BinFrequency_compare_SampleType.pdf"),density = 300)
plot(img)
```


**Figure 4F** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"Figure4F_Immcantation_ClonalFrequency_Unweighted_UniformlyResampled_Top10clone_split_Response_compare_SampleType.pdf"),density = 300)
plot(img)
```


**Figure S9C** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"S9C_Immcantation_ClonalFrequency_Unweighted_UniformlyResampled_BinFrequency_split_SampleType_compare_Response.pdf"),density = 300)
plot(img)
```


## Figure S9D - Ig clones diversity fold change pre-ICI to post-ICI in each sample type vs reponse to ICI



```{r}
# Plot diversity R/NR or sample_type for all kind of parameters
## NB : initially a loop on all parameters. Here only the unweighted and uniform=T parameter are selected

# Reload diversity list
div<-div_allIGH# Diversity of all IGH clones with Unweighted and Uniform=T and min_n=1000 parameters

# metadata at patient level
Immcantation_meta_ICI_unique_patient<-unique(Immcantation_meta_ICI_unique[,c("patient_id","R_NR")])
rownames(Immcantation_meta_ICI_unique_patient)<-Immcantation_meta_ICI_unique_patient$patient_id

# Reformat as patients_id data.frame (wider)
# div_patient<-lapply(div,function(sub_div){ # initially a list with 4 parameters. Here only 1

  # get diversity value
  # alpha_df <- sub_div@diversity # if raw results from alakazam function. Here this table is already obtained (see anonymisation)
  alpha_df <- div # only 1 table
  
  # get patient id and sample type for each sample
  alpha_df$patient_id <-gsub("_Tumor.*","",gsub("_PBMC.*","",alpha_df$sample_id))
  alpha_df$sample_type <-gsub(".*Tumor.*","Tumor",gsub(".*PBMC","PBMC",alpha_df$sample_id))
  
  # reshape wider by patient id
  alpha_df_wide <- pivot_wider(alpha_df,names_from = "sample_type",values_from = "d",id_cols =c("patient_id","q") )

  # compute T6/T0 and T6/Tumor fold change
  alpha_df_wide$FC_T6_on_T0 <- alpha_df_wide$PBMC_T6/alpha_df_wide$PBMC_T0
  alpha_df_wide$FC_T6_on_Tumor <- alpha_df_wide$PBMC_T6/alpha_df_wide$Tumor
  
  
  ## Merge back with metadata info
  alpha_df_wide<-merge(alpha_df_wide,Immcantation_meta_ICI_unique_patient,by="patient_id")

  ## convert q in a factor rather than a numeric for plot
  alpha_df_wide$q<-factor(alpha_df_wide$q)
  
  ## return patient level df
  # alpha_df_wide
# })

# now plot for each parameters used for estimation of alpha diversity
# var_to_iterate <- setNames(names(div_patient),names(div_patient))
var<-"Unweighted_UniformlyResampled"

# for(var in var_to_iterate){
  # retrive the patients lvl sample
  # hill_div <- div_patient[[var]] # if loop on 4 paramters
  hill_div <- alpha_df_wide#here only 1 parameter
  
  # create a melted na.omit df for raw value of diversity (paired test) / for FC
  hill_div_melt_raw <- reshape::melt.data.frame(hill_div[,setdiff(colnames(hill_div),c("FC_T6_on_T0", "FC_T6_on_Tumor"))],id.vars =c("patient_id" ,"q", "R_NR") )
  hill_div_melt_raw <- na.omit(hill_div_melt_raw)
  hill_div_melt_raw$variable<-factor(as.character(hill_div_melt_raw$variable),
                                 levels = c("Tumor","PBMC_T0", "PBMC_T6" , "PBMC_HD"))
  
  # FC
  hill_div_melt_FC <- reshape::melt.data.frame(hill_div[,setdiff(colnames(hill_div),c("Tumor","PBMC_T0", "PBMC_T6" , "PBMC_HD"))],id.vars =c("patient_id" ,"q", "R_NR") )
  hill_div_melt_FC <- na.omit(hill_div_melt_FC)
  hill_div_melt_FC$variable<-factor(as.character(hill_div_melt_FC$variable),
                                 levels = c("FC_T6_on_Tumor","FC_T6_on_T0"))
  
  ############################################################################
  ########## First : Plot FC and re-use previous script
  ############################################################################
  
  # diversity curve
  ## R vs NR
  ###log scale
  p<-ggplot(hill_div_melt_FC, aes(x = q, y = value, group = patient_id,color=R_NR)) +
    geom_line(alpha = 0.3) +
    geom_hline(yintercept = 1,color="black",linetype="dashed")+
    scale_color_manual(values = ColorPal$R_NR)+
    scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum)+
    # stat_summary(aes(group = R_NR), fun = median,geom = "line", linewidth = 1.5) +
    stat_summary(aes(group = R_NR),
                 fun = function(x) log10(median(10 ^ x)), # ADJUST IF LOG SCALE : order of mean/log
                 geom = "line", linewidth = 1.5) +
    labs(title = "Alpha Diversity (Hill numbers) - Fold change after treatment per patient", x = "q", y = "Diversity fold change",
         subtitle = paste("IGH clone frequency",var))+  
    facet_wrap(~variable,scales = "free")+
    theme_minimal()
  
  ##### THIS IS Sup S9D (FC T6 on T0 correspond to S9D. left part)
  
  ggsave(plot = p,
         filename = paste0(res_folder,"S9D_Immcantation_AlphaDiversity_",var,"_FoldChange_DiversityCurves_compare_Response_logScale.pdf"),
         height = 5,width = 10)


    ######## T6 on T0 >>> THIS one is actually S9D (left)
  p<-ggplot(hill_div_melt_FC[hill_div_melt_FC$variable=="FC_T6_on_T0",], aes(x = q, y = value, group = patient_id,color=R_NR)) +
    geom_line(alpha = 0.3) +
    geom_hline(yintercept = 1,color="black",linetype="dashed")+
    scale_color_manual(values = ColorPal$R_NR)+
    scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum)+
    # stat_summary(aes(group = R_NR), fun = median,geom = "line", linewidth = 1.5) +
    stat_summary(aes(group = R_NR),
                 fun = function(x) log10(median(10 ^ x)), # ADJUST IF LOG SCALE : order of mean/log
                 geom = "line", linewidth = 1.5) +
    labs(title = "Alpha Diversity (Hill numbers) - FC", x = "q", y = "Diversity fold change",
         subtitle = paste("IGH clone frequency",var))+  
    facet_wrap(~variable,scales = "free")+
    theme_minimal()
  
  ggsave(plot = p,
         filename = paste0(res_folder,"S9D_Immcantation_AlphaDiversity_",var,"_FoldChange_DiversityCurves_compare_Response_logScale_T6onT0.pdf"),
         height = 4,width = 5)

  
  
  # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
  # boxplot at specific q
  # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
  
  # alpha_q0 <- filter(hill_div_melt_FC, q == 0)
  
  ## R vs NR
  p<-ggplot(hill_div_melt_FC, aes(x = R_NR, y = value,fill=R_NR)) +
    geom_hline(yintercept = 1,color="black",linetype="dashed")+
    geom_boxplot(outlier.size = 0.5,outlier.colour = "grey")+
    scale_fill_manual(values = ColorPal$R_NR)+
    geom_point(size=0.5,position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75))+
    scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum)+
    stat_compare_means(label = "p.format")+
    labs(title = "Alpha Diversity (Hill numbers) - Fold change after treatment per patient", x = NULL, y = "Fold change",
         subtitle = paste("IGH clone frequency",var))+  
    # facet_grid(q~variable,scales = "free")+
    facet_grid(variable~q,scales = "free")+
    theme_minimal()
  
  ## This is S9D right 
  
  ggsave(plot = p,
         filename = paste0(res_folder,"S9D_Immcantation_AlphaDiversity_",var,"_FoldChange_BoxplotWithStat_AllQvalue_compare_Response.pdf"),
         height = 8,width = 10)
  
  
  ##### Save also individual plot rather than facet (in log scale)
  ######## T6 on T0 >>> This is actually S9D right
  p<-ggplot(hill_div_melt_FC[hill_div_melt_FC$variable=="FC_T6_on_T0",], aes(x = R_NR, y = value,fill=R_NR)) +
    geom_hline(yintercept = 1,color="black",linetype="dashed")+
    geom_boxplot(outlier.size = 0.5,outlier.colour = "grey")+
    scale_fill_manual(values = ColorPal$R_NR)+
    geom_point(size=0.5,position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75))+
    scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum)+
    stat_compare_means(label = "p.format")+
    labs(title = "Alpha Diversity (Hill numbers) - Fold change after treatment per patient", x = NULL, y = "Fold change",
         subtitle = paste("IGH clone frequency",var))+  
    # facet_grid(q~variable,scales = "free")+
    facet_grid(variable~q,scales = "free")+
    theme_minimal()
  
  ggsave(plot = p,
         filename = paste0(res_folder,"S9D_Immcantation_AlphaDiversity_",var,"_FoldChange_BoxplotWithStat_AllQvalue_compare_Response_T6onT0.pdf"),
         height = 4,width = 9)
  
  
  ############################################################################
  ############################################################################
  ############################################################################
  ########## Plot FC but add stat comparison of AUC per patients 
  ############################################################################
  ############################################################################
  ############################################################################

  # get back a q value in continuous data to compute AUC (rather than factor for plot)
  hill_div_melt_FC$q<-as.numeric(as.character(hill_div_melt_FC$q))
  
  # compute AUC with MESS package on all q value ...
  auc_data <- hill_div_melt_FC %>%
      group_by(patient_id,variable, R_NR) %>%
      summarise(AUC = auc(q, value))
  
  ###Evaluate normality of distribution  Shapiro–Wilk test
  # Perform Shapiro-Wilk test for each treatment group
  shapiro_results <- auc_data %>%
    group_by(variable, R_NR) %>%
    summarise(p_value = shapiro.test(AUC)$p.value)
  normality_pass <- all(shapiro_results$p_value > 0.05)

  # wilcoxon test
  #T6_on_T0
  auc_test_T6_on_T0 <- as.data.frame(auc_data)
  auc_test_T6_on_T0<-auc_test_T6_on_T0[auc_test_T6_on_T0$variable %in% "FC_T6_on_T0",]
  auc_test_T6_on_T0<-wilcox.test(AUC ~ R_NR, data = auc_test_T6_on_T0) #Test
  auc_test_T6_on_T0<-paste("wilcox.test on AUC\np-value =", format(auc_test_T6_on_T0$p.value, digits = 2, scientific = F))
  ##T6_on_Tumor
  auc_test_T6_on_Tumor <- as.data.frame(auc_data)
  auc_test_T6_on_Tumor<-auc_test_T6_on_Tumor[auc_test_T6_on_Tumor$variable %in% "FC_T6_on_Tumor",]
  auc_test_T6_on_Tumor<-wilcox.test(AUC ~ R_NR, data = auc_test_T6_on_Tumor) #Test
  auc_test_T6_on_Tumor<-paste("wilcox.test on AUC\np-value =", format(auc_test_T6_on_Tumor$p.value, digits = 2, scientific = F))
  
  ### create a dummy df to plot p_value
  pval_df <-hill_div_melt_FC %>%
    group_by(variable) %>%
    summarise(value = max(value)+1) # get max per variable to plot in top
  pval_df$q <- 8 # plot around q value of 8
  pval_df$label<-ifelse(pval_df$variable=="FC_T6_on_Tumor",
                        yes = auc_test_T6_on_Tumor,
                        no=auc_test_T6_on_T0)
  
  ###log scale
  p<-ggplot(hill_div_melt_FC, aes(x = q, y = value)) +
    geom_line(aes(group = patient_id,color=R_NR),alpha = 0.3) +
    # geom_hline(yintercept = 0,color="black",linetype="dashed")+
    geom_hline(yintercept = 1,color="black",linetype="dashed")+
    scale_color_manual(values = ColorPal$R_NR)+
    scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum,expand = expansion(mult = 0.1))+
    # scale_y_continuous(expand = expansion(mult = 0.1))+
    # stat_summary(aes(group = R_NR,color=R_NR), fun = median,geom = "line", linewidth = 1.5) +
    stat_summary(aes(group = R_NR,color=R_NR),
                 fun = function(x) log10(median(10 ^ x)), # ADJUST IF LOG SCALE : order of mean/log
                 geom = "line", linewidth = 1.5) +
    geom_text(aes(label=label),data = pval_df)+
    labs(title = "Alpha Diversity (Hill numbers) - Fold change after treatment per patient", x = "q", y = "Diversity Fold Change",
         subtitle = paste("IGH clone frequency",var))+  
    facet_wrap(~variable,scales = "free")+
    theme_minimal()

  
  ### THIS IS THE FINAL S9D left With test on AUC
  
  ggsave(plot = p,
         filename = paste0(res_folder,"S9D_Immcantation_AlphaDiversity_",var,"_FoldChange_DiversityCurves_compare_Response_logScale_WithAUC_wilcoxTest.pdf"),
         height = 5,width = 10)
  
  ##### Save also individual plot rather than facet (in log scale)
  ######## T6 on T0 >>> This one is the individual final plot ...
  p<-ggplot(hill_div_melt_FC[hill_div_melt_FC$variable %in% "FC_T6_on_T0",], aes(x = q, y = value)) +
    geom_line(aes(group = patient_id,color=R_NR),alpha = 0.3) +
    # geom_hline(yintercept = 0,color="black",linetype="dashed")+
    geom_hline(yintercept = 1,color="black",linetype="dashed")+
    scale_color_manual(values = ColorPal$R_NR)+
    scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum,expand = expansion(mult = 0.1))+
    # scale_y_continuous(expand = expansion(mult = 0.1))+
    # stat_summary(aes(group = R_NR,color=R_NR), fun = median,geom = "line", linewidth = 1.5) +
    stat_summary(aes(group = R_NR,color=R_NR),
                 fun = function(x) log10(median(10 ^ x)), # ADJUST IF LOG SCALE : order of mean/log
                 geom = "line", linewidth = 1.5) +
    geom_text(aes(label=label),data = pval_df[pval_df$variable %in% "FC_T6_on_T0",])+
    labs(title = "Alpha Diversity (Hill numbers) - FC", x = "q", y = "Diversity Fold Change",
         subtitle = paste("IGH clone frequency",var))+  
    facet_wrap(~variable,scales = "free")+
    theme_minimal()
  
  ggsave(plot = p,
         filename = paste0(res_folder,"S9D_Immcantation_AlphaDiversity_",var,"_FoldChange_BoxplotWithStat_AllQvalue_compare_Response_T6onT0_WithAUC_wilcoxTest.pdf"),
         height = 4,width = 5)
  
#}

  
```

**Figure S9D (left : with AUC test)** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"S9D_Immcantation_AlphaDiversity_Unweighted_UniformlyResampled_FoldChange_BoxplotWithStat_AllQvalue_compare_Response_T6onT0_WithAUC_wilcoxTest.pdf"),density = 300)
plot(img)
```


**Figure S9D (right)** 
```{r,fig.height=7,fig.width=10}
img <- magick::image_read_pdf(paste0(res_folder,"S9D_Immcantation_AlphaDiversity_Unweighted_UniformlyResampled_FoldChange_BoxplotWithStat_AllQvalue_compare_Response_T6onT0.pdf"),density = 300)
plot(img)
```


## Figure S10A - Class-switched Ig clones diversity and frequency in each sample type 


```{r}
# Create a list of diversity object according to different min_n filter (list containing lists ...)
## diversity
  ## NB : initially a loop on all parameters. Here only the unweighted and uniform=T parameter are selected
# div_list <- list(noFilter=div_noFilter,
#                  min1000=div_min1000,
#                  min6000=div_min6000)
# 
# 
# ## abundance
# abund_list <- list(noFilter=abund_noFilter,
#                    min1000=abund_min1000,
#                    min6000=abund_min6000)

# Create to vector to iterate on
## different min_n filter
# minN_to_iterate<-setNames(names(div_list),nm = names(div_list)) 

minN <- "min1000" # set this here

## variables in each list : downsample or not / unweighted or umi_weighted
# var_to_iterate<-setNames(names(div_list$noFilter),nm = names(div_list$noFilter)) 
var <- "Unweighted_UniformlyResampled"

# Iterate plot diversity R/NR or sample_type for all kind of parameters
## NB : min6000 raised an error for the correlation analysis > to few samples > do not consider this filter which is too stringent .......

# for(minN in minN_to_iterate){
#   print(paste("start:",minN))
#   
#   for(var in var_to_iterate){
#     print(paste("start:",var))
    
    # get correct diversity
    # hill_div <- div_list[[minN]][[var]]@diversity
    hill_div <- div_ClassSwitch # Diversity of class switched IGH clones with Unweighted and Uniform=T and min_n=1000 parameters
  
    ## Merge back with metadata info
    hill_div<-merge(hill_div,Immcantation_meta_ICI_unique[,c("sample_id","patient_id","sample_origin","sample_timepoint","sample_type","sample_type2","R_NR")],by="sample_id")
    
    ## convert q in a factor rather than a numeric for plot
    hill_div$q<-factor(hill_div$q)
    
    ########    ########    ########    ########    ########
    # I - GLOBAL diversity curve
    ########    ########    ########    ########    ########
    ## Ia- sample_type global 
    
    ### Compute number of sample in the corresponding filter/parameter
    num_samples <- table(hill_div[hill_div$q =="0",c("sample_type2")])
    num_samples<-data.frame(num_samples)
    num_samples <- num_samples[num_samples$Freq !=0,] # remove lines without any samples
    final_label <- paste(paste("Tumor:",paste(num_samples$Freq[num_samples$Var1 %in% "Tumor.T0"],collapse = ", ")),
                         paste("PBMC.T0:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.T0"],collapse = ", ")),
                         paste("PBMC.T6:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.T6"],collapse = ", ")),
                         paste("PBMC.HD:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.HD"],collapse = ", ")),
                         sep = " ; ")
  
    #### after removing PBMC HD
    
    ### diversity curve
    p<-ggplot(hill_div[!grepl("HD",hill_div$sample_id),], aes(x = q, y = d,color=sample_type2)) + # reomve HD
      geom_line(aes(group = sample_id),alpha = 0.3) +
      stat_summary(aes(group = sample_type2), fun = median, geom = "line", linewidth = 1.5) +
      scale_color_manual(values = ColorPal$sample_type2)+
      stat_compare_means(label = "p.signif",size=3.5)+
      # stat_compare_means(size=3.5)+
      labs(title = "Diversity of class-switched IGH", x = "q", y = "Diversity (class-switched IGH)",
           subtitle = gsub("...PBMC.HD:..","",final_label), # reomve HD
           color="sample_type",
           caption = paste(minN,"sequences, ",var))+  
        theme_classic()
    
    # THIS IS SUP S10A left
    ggsave(plot = p,
           filename = paste0(res_folder,"S10A_Immcantation_AlphaDiversity_ClassSwitchedIGH_",minN,"_",var,"_DiversityCurves_Compare_SampleType_RemoveHD.pdf"),
           height = 5,width = 7)
    
   
    
    ########    ########    ########    ########    ######## ################################
    # Abundance of clones
    ########    ########    ########    ########    ######## ################################

    # Get correct clone_frequency according to variables in the loop
    # clone_frequency <- abund_list[[minN]][[var]]@abundance
    clone_frequency <- abund_ClassSwitch # Abundance of class switched IGH clones with Unweighted and Uniform=T and min_n=1000 parameters
    
    ## Merge back with metadata info
    clone_frequency<-merge(clone_frequency,Immcantation_meta_ICI_unique[,c("sample_id","patient_id","sample_origin","sample_timepoint","sample_type","sample_type2","R_NR")],by="sample_id")

    ### Compute number of sample in the corresponding filter/parameter
    clone_frequency$rank_character<-as.character(clone_frequency$rank)
    
    ###### number of samples split by R_NR group
    num_samples <- table(clone_frequency[clone_frequency$rank_character =="1",c("R_NR","sample_type")])
    num_samples<-data.frame(num_samples)
    num_samples <- num_samples[num_samples$Freq !=0,] # remove lines without any samples
    num_samples$label <- paste(num_samples$Freq,num_samples$R_NR,sep = "-") # paste n and R_NR group
    final_label_R_NR <- paste(paste("Tumor:",paste(num_samples$label[num_samples$sample_type %in% "Tumor.T0"],collapse = ", ")),
                         paste("PBMC.T0:",paste(num_samples$label[num_samples$sample_type %in% "PBMC.T0"],collapse = ", ")),
                         paste("PBMC.T6:",paste(num_samples$label[num_samples$sample_type %in% "PBMC.T6"],collapse = ", ")),
                         sep = " ; ")
    
    ###### number of samples overall (only by sample type)
    num_samples <- table(clone_frequency[clone_frequency$rank_character =="1",c("sample_type2")])
    num_samples<-data.frame(num_samples)
    num_samples <- num_samples[num_samples$Freq !=0,] # remove lines without any samples
    final_label_SampleType <- paste(paste("Tumor:",paste(num_samples$Freq[num_samples$Var1 %in% "Tumor.T0"],collapse = ", ")),
                         paste("PBMC.T0:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.T0"],collapse = ", ")),
                         paste("PBMC.T6:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.T6"],collapse = ", ")),
                         paste("PBMC.HD:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.HD"],collapse = ", ")),
                         sep = " ; ")
      
      
    ###########################################################################
    # Top clones
    ###########################################################################
    ## top 10
    clone_frequency_top10<-clone_frequency
    clone_frequency_top10$rank_character<-as.character(clone_frequency_top10$rank)
    clone_frequency_top10 <- clone_frequency_top10[clone_frequency_top10$rank_character %in% as.character(1:10),]#filter top 10
    clone_frequency_top10$rank_character <- factor(clone_frequency_top10$rank_character,levels = as.character(1:10)) # as factor
    
    ##  without HD PBMC
    p<-ggplot(clone_frequency_top10[!grepl("HD",clone_frequency_top10$sample_id),],aes(x =rank_character,y= p,fill=sample_type2))+ # remove HD
      geom_boxplot(outlier.size = 0.5,outlier.colour = "grey")+
      geom_point(size=0.4,position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75),color="darkgrey")+
      scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum)+
      scale_fill_manual(values = ColorPal$sample_type2)+
      # facet_grid(.~R_NR,scales = "free_x",space = "free_x")+
      stat_compare_means(label = "p.signif")+
      theme_classic()+
      # theme(legend.position = "none")+
      labs(y="Clone frequency of class-switched IGH",
           title = "Rank clone abundance of class-switched IGH - Top 10 clones",
           x="Rank of top 10 clones",
           subtitle = gsub("...PBMC.HD...","",final_label_SampleType), # remove HD
           fill="sample_type",
           caption = paste(minN,"sequences, ",var))
    
    ### THIS IS SUP S10A right
    ggsave(plot = p,
           filename = paste0(res_folder,"S10A_Immcantation_AlphaDiversity_ClassSwitchedIGH_",minN,"_",var,"_Top10clone_compare_SampleType_RemoveHD.pdf"),
           height = 4,width = 6)
    
#   }
# }

```

**Figure S10A (left)** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"S10A_Immcantation_AlphaDiversity_ClassSwitchedIGH_min1000_Unweighted_UniformlyResampled_DiversityCurves_Compare_SampleType_RemoveHD.pdf"),density = 300)
plot(img)
```


**Figure S10A (right)** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"S10A_Immcantation_AlphaDiversity_ClassSwitchedIGH_min1000_Unweighted_UniformlyResampled_Top10clone_compare_SampleType_RemoveHD.pdf"),density = 300)
plot(img)
```


## Figure S10B & S10C - IgGs or IgAs pure clones diversity and frequency in each sample type 


```{r}
# Create a list of diversity object according to different min_n filter (list containing lists ...)
## diversity
  ## NB : initially a loop on all parameters. Here only the unweighted and uniform=T parameter are selected
# 
# div_list <- list(noFilter=div_noFilter,
#                  min500=div_min500,
#                  min1000=div_min1000
#                  )
# 
# 
# ## abundance
# abund_list <- list(noFilter=abund_noFilter,
#                    min500=abund_min500,
#                    min1000=abund_min1000
#                    )

# Create to vector to iterate on
## different min_n filter
# minN_to_iterate<-setNames(names(div_list),nm = names(div_list)) 

minN <- "min500" # parameter set. Initially loop to test multiple

## Isotype factor

div_list <- list(IgGs=div_IgGs,
                 IgAs=div_IgAs) # set a list on isotype

abund_list <- list(IgGs=abund_IgGs,
                 IgAs=abund_IgAs) # same for abundance

# iso_vect <- setNames(names(div_list$noFilter),nm = names(div_list$noFilter)) 
iso_vect <- setNames(names(div_list),nm = names(div_list)) 

## variables in each list : downsample or not / unweighted or umi_weighted
# var_to_iterate<-setNames(names(div_list$noFilter$IgGs),nm = names(div_list$noFilter$IgGs)) 
var <- "Unweighted_UniformlyResampled" # parameter set. Initially loop to test multiple

# Iterate plot diversity R/NR or sample_type for all kind of parameters

# for(minN in minN_to_iterate){
#   print(paste("start:",minN))
#   
  for(iso in iso_vect){
    # print(paste("start:",iso))
    
      # for(var in var_to_iterate){
      # print(paste("start:",var))
      # 
      # get correct diversity
      # hill_div <- div_list[[minN]][[iso]][[var]]@diversity ### Add isotype level of the list
    
       hill_div <- div_list[[iso]] # Diversity of IgGs or IgAs clones with Unweighted and Uniform=T and min_n=500 parameters
  

      ## Merge back with metadata info
      hill_div<-merge(hill_div,Immcantation_meta_ICI_unique[,c("sample_id","patient_id","sample_origin","sample_timepoint","sample_type","sample_type2","R_NR")],by="sample_id")
      
      ## convert q in a factor rather than a numeric for plot
      hill_div$q<-factor(hill_div$q)
      
        
      ########    ########    ########    ########    ########
      # I - GLOBAL diversity curve
      ########    ########    ########    ########    ########
      ## Ia- sample_type global 
      
      ### Compute number of sample in the corresponding filter/parameter
      num_samples <- table(hill_div[hill_div$q =="0",c("sample_type2")])
      num_samples<-data.frame(num_samples)
      num_samples <- num_samples[num_samples$Freq !=0,] # remove lines without any samples
      final_label <- paste(paste("Tumor:",paste(num_samples$Freq[num_samples$Var1 %in% "Tumor.T0"],collapse = ", ")),
                           paste("PBMC.T0:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.T0"],collapse = ", ")),
                           paste("PBMC.T6:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.T6"],collapse = ", ")),
                           paste("PBMC.HD:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.HD"],collapse = ", ")),
                           sep = " ; ")
     
      ####  after removing PBMC HD
      
      ### diversity curve
      p<-ggplot(hill_div[!grepl("HD",hill_div$sample_id),], aes(x = q, y = d,color=sample_type2)) + # reomve HD
        geom_line(aes(group = sample_id),alpha = 0.3) +
        stat_summary(aes(group = sample_type2), fun = median, geom = "line", linewidth = 1.5) +
        scale_color_manual(values = ColorPal$sample_type2)+
        stat_compare_means(label = "p.signif",size=3.5)+
        # stat_compare_means(size=3.5)+
        labs(title = paste("Diversity of",iso,"clones"), x = "q", y = paste("Diversity of",iso),
             subtitle = gsub("...PBMC.HD:..","",final_label), # reomve HD
             color="sample_type",
             caption = paste(minN,"sequences, ",var))+  
          theme_classic()
      
      ### This is SUP S10B or S10C left (depending on isotype)
      prefixe <- ifelse(iso=="IgGs",yes = "S10B_",no = "S10C_")
      ggsave(plot = p,
             filename = paste0(res_folder,prefixe,"Immcantation_AlphaDiversity_All_",iso,"_",minN,"_",var,"_DiversityCurves_Compare_SampleType_RemoveHD.pdf"),
             height = 5,width = 7)
      
      
      ########    ########    ########    ########    ######## ################################
      # Abundance of clones
      ########    ########    ########    ########    ######## ################################
  
      # Get correct clone_frequency according to variables in the loop
      # clone_frequency <- abund_list[[minN]][[iso]][[var]]@abundance ### Add the isotye level in list
      clone_frequency <- abund_list[[iso]] # Abundance of IgGs or IgAs clones with Unweighted and Uniform=T and min_n=500 parameters
      
      ## Merge back with metadata info
      clone_frequency<-merge(clone_frequency,Immcantation_meta_ICI_unique[,c("sample_id","patient_id","sample_origin","sample_timepoint","sample_type","sample_type2","R_NR")],by="sample_id")
  
      ### Compute number of sample in the corresponding filter/parameter
      clone_frequency$rank_character<-as.character(clone_frequency$rank)
      
      ###### number of samples split by R_NR group
      num_samples <- table(clone_frequency[clone_frequency$rank_character =="1",c("R_NR","sample_type")])
      num_samples<-data.frame(num_samples)
      num_samples <- num_samples[num_samples$Freq !=0,] # remove lines without any samples
      num_samples$label <- paste(num_samples$Freq,num_samples$R_NR,sep = "-") # paste n and R_NR group
      final_label_R_NR <- paste(paste("Tumor:",paste(num_samples$label[num_samples$sample_type %in% "Tumor.T0"],collapse = ", ")),
                           paste("PBMC.T0:",paste(num_samples$label[num_samples$sample_type %in% "PBMC.T0"],collapse = ", ")),
                           paste("PBMC.T6:",paste(num_samples$label[num_samples$sample_type %in% "PBMC.T6"],collapse = ", ")),
                           sep = " ; ")
      
      ###### number of samples overall (only by sample type)
      num_samples <- table(clone_frequency[clone_frequency$rank_character =="1",c("sample_type2")])
      num_samples<-data.frame(num_samples)
      num_samples <- num_samples[num_samples$Freq !=0,] # remove lines without any samples
      final_label_SampleType <- paste(paste("Tumor:",paste(num_samples$Freq[num_samples$Var1 %in% "Tumor.T0"],collapse = ", ")),
                           paste("PBMC.T0:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.T0"],collapse = ", ")),
                           paste("PBMC.T6:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.T6"],collapse = ", ")),
                           paste("PBMC.HD:",paste(num_samples$Freq[num_samples$Var1 %in% "PBMC.HD"],collapse = ", ")),
                           sep = " ; ")
        
        
      ###########################################################################
      # Top clones
      ###########################################################################
      ## top 10
      clone_frequency_top10<-clone_frequency
      clone_frequency_top10$rank_character<-as.character(clone_frequency_top10$rank)
      clone_frequency_top10 <- clone_frequency_top10[clone_frequency_top10$rank_character %in% as.character(1:10),]#filter top 10
      clone_frequency_top10$rank_character <- factor(clone_frequency_top10$rank_character,levels = as.character(1:10)) # as factor

      ## without HD PBMC
      p<-ggplot(clone_frequency_top10[!grepl("HD",clone_frequency_top10$sample_id),],aes(x =rank_character,y= p,fill=sample_type2))+ # remove HD
        geom_boxplot(outlier.size = 0.5,outlier.colour = "grey")+
        geom_point(size=0.4,position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75),color="darkgrey")+
        scale_y_continuous(trans = "log10", breaks = base_breaksLog10(),labels = prettyNum)+
        scale_fill_manual(values = ColorPal$sample_type2)+
        # facet_grid(.~R_NR,scales = "free_x",space = "free_x")+
        stat_compare_means(label = "p.signif")+
        theme_classic()+
        # theme(legend.position = "none")+
        labs(y=paste("Clone frequency of",iso),
             title = paste("Rank clone abundance of",iso," - Top 10 clones"),
             x="Rank of top 10 clones",
             subtitle = gsub("...PBMC.HD...","",final_label_SampleType), # remove HD
             fill="sample_type",
             caption = paste(minN,"sequences, ",var))
      
      ### This is SUP S10B or S10C right (depending on isotype)
      prefixe <- ifelse(iso=="IgGs",yes = "S10B_",no = "S10C_")
      ggsave(plot = p,
             filename = paste0(res_folder,prefixe,"Immcantation_AlphaDiversity_All_",iso,"_",minN,"_",var,"_Top10clone_compare_SampleType_RemoveHD.pdf"),
             height = 4,width = 6)
      
    # } # end variable loop
    
    
  } # end Isotype loop

# } # end min n loop

```


**Figure S10B (left)** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"S10B_Immcantation_AlphaDiversity_All_IgGs_min500_Unweighted_UniformlyResampled_DiversityCurves_Compare_SampleType_RemoveHD.pdf"),density = 300)
plot(img)
```


**Figure S10B (right)** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"S10B_Immcantation_AlphaDiversity_All_IgGs_min500_Unweighted_UniformlyResampled_Top10clone_compare_SampleType_RemoveHD.pdf"),density = 300)
plot(img)
```



**Figure S10C (left)** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"S10C_Immcantation_AlphaDiversity_All_IgAs_min500_Unweighted_UniformlyResampled_DiversityCurves_Compare_SampleType_RemoveHD.pdf"),density = 300)
plot(img)
```


**Figure S10C (right)** 
```{r,fig.height=7,fig.width=7}
img <- magick::image_read_pdf(paste0(res_folder,"S10C_Immcantation_AlphaDiversity_All_IgAs_min500_Unweighted_UniformlyResampled_Top10clone_compare_SampleType_RemoveHD.pdf"),density = 300)
plot(img)
```


# SessionInfo

```{r}
sessionInfo()
```
